### 2018-06-13 README.Rmd
Abstract and first analysis

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# RefractoryAnaOrg

This repository contains the data and code for our paper:

> Authors, (YYYY). _Title of paper_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _Title of paper_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for 'Title of paper'_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: <https://github.com/wolass/RefractoryAnaOrg/archive/master.zip>

Or you can install this compendium as an R package, RefractoryAnaOrg, from GitHub with:
```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("wolass/RefractoryAnaOrg")
```
### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-BY-4.0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2018-06-13 analysis/paper/paper.Rmd
Abstract and first analysis

---
title: "Refractory anaphylaxis. Data from the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik
  - Margitta Worm
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r, setup, echo = FALSE,warning=FALSE,message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

library(RefractoryAnaOrg) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
require(magrittr)
```

Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,

**Corresponding author: **
Prof. Dr. med. M. Worm margitta.worm@charite.de
Phone: +49 30 450 529 005;	  Fax: +49 30 450 529 902

**Keywords:** 
anaphylaxis, adrenaline (epinaphrine), Betablockers, Insect venom allergy, Drug allergic reactions

# Abstract: 
<!-- 350 Words unstructured -->
Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 mg adrenaline) is a rare and often fatal hypersensitivity reaction. Its definition, prevalence, and risk factors are only vaguely described.  
Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we were able to identify cases of refractory anaphylaxis (n = `r v$casesR`) and compare them to a control group of severe anaphylaxis cases (n = `r v$casesC`).
Drugs significantly more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).  Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r elicitExact[4,3]`% vs `r elicitExact[4,2]`%, p = `r elicitExact[4,4]`). The refractory cases occurred more frequently in a perioperative setting (`r v$perioperative[3]`% vs. `r v$perioperative[2]`, p < 0.0001). Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases was given intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001). Second line treatment options: vasopression with dopamine, methylene blue, glucagon i.v. were not used at all in the treatment of refractory cases. Mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).
Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these standards leaves room for improvement. Low use of intramuscular adrenaline as a first line therapy might contribute to refractoriness of an episode. Also, lack of suggested 2nd line medication use in refractory cases indicate that availability of these drug is restricted or anaphylaxis management guidelines may need revision in order to provide algorithms for severe refractory anaphylaxis.

# Introduction: 
The mainstay of Anaphylaxis management is the intramuscular dose of adrenaline, but in the most severe cases of anaphylaxis it might insufficient to restore a stable patient status.

Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which is unresponsive to the treatment with at least 2 doses of minimum 300 mg adrenaline (lack of expected normalization of clinical symptoms).

# Methods
We aimed to distinguish the prevalence of refractory anaphylaxis among anaphylaxis cases and factors which may increase the risk of a refractory anaphylaxis episode. 
Using the Data from the European Anaphylaxis Registry we selected cases where patients received at least two doses of adrenalin, and failed to recover adequately. The flowchart represents the detailed cases selection process.

The final database consisted of `r sum(countries)` cases from `r length(countries)` countries: `r paste0(names(countries),": ",countries,collapse=", ")`.

# Results

The European Anaphylaxis Registry contained `r data3[,1] %>% length` cases of anaphylaxis and in `r sum(sumAdre[3:6])` cases at lest 2 doses of adrenalin were administered to the patient.

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
```

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

# Figures

```{r flowchart, fig.cap="Flowchart illustrating the cases selection process for the final database."}
knitr::include_graphics("../figures/flow.png")
```

# Tables

```{r demoTab}
knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. ")
```

```{r elicitorTab}
knitr::kable(elicitorTab, caption = "Summary of elicitors in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ",
             col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]"))
```

```{r elicitExact}
knitr::kable(cbind(elicitExact[,5],elicitExact[,2:4]), caption = "Summary of elicitors in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ",
             col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value"))
```

* p-value derived from the Fischer exact test.

```{r therapyTab}
knitr::kable(therapyTab[,1:4], caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ",
             col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"))
```

* p-value derived from the Fischer exact test.

### 2018-06-13 R/analysis.R

@@ -1,0 +1,643 @@
+##### PACKAGES #####
+# Put all the necessary packages here
+# So that other could easily install them before compiling multiple times
+# I prefer to use require() insted of library() because
+# it will only load the library if it hasn't already.
+require(ggplot2)
+require(magrittr)
+
+##### DATA LOAD ######
+# Load your data in this section - it is crucial to know on what data you are actually working
+#load("analysis/data/raw_data/data3.R")
+# New database
+load("analysis/data/raw_data/data4.R")
+data3 <- data4
+
+##### FUNCTIONS #####
+# Put all the functions above the code below,
+# so that you can easily find them later and execute the code seemlesly
+
+sum_adrenalines <- function(vars,data){
+  lapply(data[,vars],function(var){
+    ifelse(!is.na(var),
+           ifelse(var=="yes", 1, 0),
+           0)}) %>%
+    Reduce(f=`+`) %>%
+    return()
+}
+
+rcalc <- function(var,level="yes"){
+  tab <- table(data3[,var],data3$rANA)
+  if(level %in% (data3[,var] %>% levels)){
+    pn <- tab[level,"no"]
+    pp <- tab[level,"yes"]
+    l <- level
+  } else {
+    pn <- tab[2,"no"]
+    pp <- tab[2,"yes"]
+    l <- levels(data3[,var])[2]
+  }
+  nn <- sum(tab[,"no"])-pn
+  np <- sum(tab[,"yes"])-pp
+  eval <- c(pn,pp,nn,np) %>% matrix(byrow = T,ncol = 2)
+  prt <- (prop.table(eval,2)*100) %>% {signif(.,3)}
+  p <- eval %>% fisher.test() %>% .$p.value %>% signif(3)
+  return(c(var,prt[1,1],prt[1,2],p,l))
+}
+
+##### VARIABLES #####
+# If you are making new variables than put them here for reference later.
+# It is always good to have all derived data show up early in the script
+# so that you would not have issues with code execution later.
+
+adr2 <- which(data3$d_560_adren2_v5=="yes") # Which patients recieved 2 times the adrenilin?
+data.r <- data3[adr2,]
+#table(data.r$d_severity_rm)
+
+sev <- data.r %>% {which(.$d_severity_rm%in%c("grade III","grade IV"))}
+data.r <- data.r[sev,]
+
+
+# So First get the idea who got 2 shots of adrenaline or more
+adre.no <- sum_adrenalines(vars= c("q_521_autoinj_v5",    # Patient self medicated
[…]

### 2018-07-02 analysis/paper/paper.Rmd
abstract submission files

@@
 ---
-title: "Refractory anaphylaxis. Data from the European Anaphylaxis Registry"
+title: "Refractory anaphylaxis: Data from the European Anaphylaxis Registry"
 author:
-  - Wojciech Francuzik
-  - Margitta Worm
+  - Wojciech Francuzik*
+  - Sabine Dölle-Bierke*
+  - Franziska Ruëff**
+  - Margitta Worm*
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
-Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
+*Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
+
+**Klinik und Poliklinik für Dermatologie und Allergologie, Frauenlobstraße 9–11, 80337 München
 
@@
 **Keywords:** 
-anaphylaxis, adrenaline (epinaphrine), Betablockers, Insect venom allergy, Drug allergic reactions
+anaphylaxis, adrenaline (epinaphrine), betablockers, Insect venom allergy, Drug allergic reactions
 
@@
 Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 mg adrenaline) is a rare and often fatal hypersensitivity reaction.
-Its definition, prevalence, and risk factors are only vaguely described.
-Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we were able to identify cases of refractory anaphylaxis (n = `r v$casesR`) and compare them to a control group of severe anaphylaxis cases (n = `r v$casesC`).
-Drugs significantly more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).
+Based on this, definition, its prevalence, and risk factors are only vaguely described.
+Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we were able to identify cases of refractory anaphylaxis (n = `r v$casesR`) and to analyse these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
+The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).
 Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r elicitExact[4,3]`% vs `r elicitExact[4,2]`%, p = `r elicitExact[4,4]`).
@@
 Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases was given intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001).
-Second line treatment options: vasopression with dopamine, methylene blue, glucagon i.v.
-were not used at all in the treatment of refractory cases.
-Mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).
-Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these standards leaves room for improvement.
-Low use of intramuscular adrenaline as a first line therapy might contribute to refractoriness of an episode.
-Also, lack of suggested 2nd line medication use in refractory cases indicate that availability of these drug is restricted or anaphylaxis management guidelines may need revision in order to provide algorithms for severe refractory anaphylaxis.
+Second line treatment options (e.g.
+vasopression with dopamine, methylene blue, glucagon) were not used at all in the treatment of refractory cases.
+The mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).
+Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these standards can be improved.
+The low use of intramuscular adrenaline as a first line therapy might contribute to refractoriness of an episode.
+Also, the lack of suggested second line medication use in refractory cases indicate that the availability of these drug is restricted or anaphylaxis management guidelines may need a revision in order to provide clear algorithms for severe refractory anaphylaxis.
 
+<!--
 # Introduction: 
-The mainstay of Anaphylaxis management is the intramuscular dose of adrenaline, but in the most severe cases of anaphylaxis it might insufficient to restore a stable patient status.
+The mainstay of anaphylaxis management is the intramuscular dose of adrenaline, but in the most severe cases of anaphylaxis it might insufficient to restore a stable patient status.
 
-Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which is unresponsive to the treatment with at least 2 doses of minimum 300 mg adrenaline (lack of expected normalization of clinical symptoms).
+Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which is unresponsive to the treatment with at least 2 doses of minimum 300 mg adrenaline without the expected normalization of clinical symptoms.
 
@@
 Using the Data from the European Anaphylaxis Registry we selected cases where patients received at least two doses of adrenalin, and failed to recover adequately.
-The flowchart represents the detailed cases selection process.
+The flowchart in figure \@ref(fig:flowchart) represents the detailed cases selection process.
 
@@
 
-The European Anaphylaxis Registry contained `r data3[,1] %>% length` cases of anaphylaxis and in `r sum(sumAdre[3:6])` cases at lest 2 doses of adrenalin were administered to the patient.
+## Demographics
+The European Anaphylaxis Registry contained `r {data3[,"rANA"]=="yes"} %>% which %>% length` cases of refractory anaphylaxis and in `r {data3[,"rANA"]=="no"} %>% which %>% length` cases severe, non-refractory anaphlaxis was diagnosed.
+The mean age at the reaction was `r {data3$d_age[data3[,"rANA"]=="yes"]} %>% mean(na.rm=T) %>% signif(3)` ± `r data3$d_age[data3[,"rANA"]=="yes"] %>% sd(na.rm=T)` years, and did not differ from severe, non-refractory cases, p = `r t.test(data3$d_age[data3[,"rANA"]=="yes"],data3$d_age[data3[,"rANA"]=="no"])$p.value %>% signif(3)`).
+There were however significantly more cases of concomitant malignant diseases in the refractory anaphylaxis group.
+The demographic summary of refractory cases can be found in table \@ref(tab:demoTab).
 
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
-```
+## Elicitors
+Refractory anaphylaxis was most commonly elicited by drugs (significantly more often than in severe, non-refractory cases), followed by food and insects (table \@ref(tab:elicitorTab)). 
+The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics and X-ray contrast medium.
+Insect-venom refractory cases were significantly more often elicited by bee-stings than severe, non-refractory cases, although the yellow-jacket-stings were most common in other anaphylaxis cases (\@ref(tab:elicitExact)).
+
+## Therapy
+Adrenaline  iv. as a first-line treatment of anaphylaxis was significantly more often given in refractory cases (`r therapyTab[2,3]`% vs `r therapyTab[2,3]`%).
 

### 2018-07-02 R/analysis.R

@@ -45,6 +45,19 @@ rcalc <- function(var,level="yes"){
   return(c(var,prt[1,1],prt[1,2],p,l))
 }
 
+f1 <- function(x){
+  names(x) <- NULL
+  return(do.call(t.test,x)$p.value %>% signif(3))
+}
+f2 <- function(x){
+  x %>% table(data3$rANA) %>% {.[1:2,]} %>% summary() %>% {.$p.value} %>% signif(3)
+}
+f3 <- function(x){
+  x %>%
+    lapply(function(x){(length(which(x=="yes"))/length(x)*100) %>% signif(3)})
+}
+
+
 ##### VARIABLES #####
 # If you are making new variables than put them here for reference later.
 # It is always good to have all derived data show up early in the script
@@ -493,19 +506,47 @@ v$casesInsect <- table(data3$q_340_insects[data3$rANA=="yes"])
 v$mortality <- rcalc("q_140_fatal")# ** Tables ----
 # This section deals with tables. Prepare them here and reference to them in the paper.Rmd
 
+sdb <- data3[which(data3$rANA=="no"),]
+
 #### 1. Demography ################
 demoTab <- cbind(n = rdb$b_sex %>% summary(),
-                      Age = rdb$d_age %>% split(rdb$b_sex) %>%
+                Age = rdb$d_age %>% split(rdb$b_sex) %>%
                               lapply(.,function(x){mean(x) %>% signif(3)}),
-                      Cardiologic = rdb$q_410_cardio_cur%>% split(rdb$b_sex) %>%
-                              lapply(.,function(x){(length(which(x=="yes"))/length(x)*100) %>% signif(3)}),
-                      DM = rdb$q_410_diab_cur_v6%>% split(rdb$b_sex) %>%
-                              lapply(.,function(x){(length(which(x=="yes"))/length(x)*100) %>% signif(3)}),
-                      `Food allergy` = rdb$q_410_foodallergy_cur_v6%>% split(rdb$b_sex) %>%
-                              lapply(.,function(x){(length(which(x=="yes"))/length(x)*100) %>% signif(3)})
+                Cardiologic = rdb$q_410_cardio_cur%>% split(rdb$b_sex) %>%
+                              f3,
+                DM = rdb$q_410_diab_cur_v6%>% split(rdb$b_sex) %>%
[…]

### 2018-08-15 analysis/paper/paper.Rmd
berfore changing after a while

@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
-abstract: |
-  Text of abstract
+csl: expert-review-of-clinical-immunology.csl # Insert path for the bib-style
+#abstract: 
 keywords: |
@@
 
+```{r}
+periopAna <- data3$rANA[which(data3$q_152_location=="medical practice, hospital")] %>% 
+  as.factor %>%  summary()
+
+```
+
 *Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
@@
 **Keywords:** 
-anaphylaxis, adrenaline (epinaphrine), betablockers, Insect venom allergy, Drug allergic reactions
+anaphylaxis, adrenaline (epinephrine), beta-blockers, Insect venom allergy, Drug allergic reactions
 
-# Abstract: 
+# Abstract
 <!-- 350 Words unstructured -->
 Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 mg adrenaline) is a rare and often fatal hypersensitivity reaction.
-Based on this, definition, its prevalence, and risk factors are only vaguely described.
+Based on this, its definition, prevalence, and risk factors are only vaguely described.
 Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we were able to identify cases of refractory anaphylaxis (n = `r v$casesR`) and to analyse these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
@@
 Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r elicitExact[4,3]`% vs `r elicitExact[4,2]`%, p = `r elicitExact[4,4]`).
-The refractory cases occurred more frequently in a perioperative setting (`r v$perioperative[3]`% vs.
+The refractory cases occurred more frequently in a perioperative setting E(`r v$perioperative[3]`% vs.
 `r v$perioperative[2]`, p < 0.0001).
@@
 
-<!--
-# Introduction: 
-The mainstay of anaphylaxis management is the intramuscular dose of adrenaline, but in the most severe cases of anaphylaxis it might insufficient to restore a stable patient status.
+# Introduction
 
-Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which is unresponsive to the treatment with at least 2 doses of minimum 300 mg adrenaline without the expected normalization of clinical symptoms.
+Anaphylaxis is a non-homogeneous clinical diagnosis, depending on various triggering, and augmenting factors.
+This variability introduces a wide range of possible reaction-symptom severity.
+Therefore multiple sub-types of anaphylaxis have been previously identified (i.e.
+food dependent exercise induced anaphylaxis, venom anaphylaxis, biphasic anaphylaxis).
+
+The mainstay of anaphylaxis management is the intramuscular dose of adrenaline, but in the most severe cases of anaphylaxis it might be insufficient to restore a stable patient status.
+Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least 2 doses of minimum 300 mg adrenaline, does not result in an expected normalization of clinical symptoms.
+Typical elicitors and symptoms of refractory anaphylaxis, as well as the therapeutic strategy for the most severe cases differ from the usual reactions and call for a specific research and targeted guideline development for refractory anaphylaxis cases.
+
+We aimed to distinguish the prevalence of refractory anaphylaxis among anaphylaxis cases and to describe its symptoms and factors which may increase the risk of a refractory anaphylaxis episode.
 
 # Methods
-We aimed to distinguish the prevalence of refractory anaphylaxis among anaphylaxis cases and factors which may increase the risk of a refractory anaphylaxis episode. 
-Using the Data from the European Anaphylaxis Registry we selected cases where patients received at least two doses of adrenalin, and failed to recover adequately.
+The European Anaphylaxis Registry (described in detail elsewhere [@Grabenhenrich2016]) provided data for this analysis status from May 2018.
+Based on  an electronic questionnaire (available at www.anaphylaxie.net), we selected cases where patients received at least two doses of adrenaline, and failed to recover adequately and assigned them to the "refractory anaphylaxis group".
 The flowchart in figure \@ref(fig:flowchart) represents the detailed cases selection process.
 
-The final database consisted of `r sum(countries)` cases from `r length(countries)` countries: `r paste0(names(countries),": ",countries,collapse=", ")`.
+The final database consisted of `r sum(countries)` cases of refractory anaphylaxis from `r length(countries)` countries: `r paste0(names(countries),": ",countries,collapse=", ")`.
+We compared these to a group of severe, non-refractory cases of anaphylaxis.
+Severe reactions were identified based on the definition of by NIAID/FAAN [@Sampson2006] and presented with significant hypoxia, cyanosis, hypotension (SBP < 90 mm Hg in adults), confusion, collapse, loss of conciusness, or incontinence.
+We compared the frequency of various elicitors, symptoms, and factors known to incease the risk of severe anphylaxis [@Worm2018] in both groups. We also compared how the cases were managed.
+
+<!-- ## Data handling and statisitcs -->
+The statistical analysis was performed in the R Statistical Package [@R].
+Simple comparison of nominal variables was performed using Fischer's exact test, continous variables were analysed using Mann-Whithney U test.
+We defined statistical significance as α = 0.05.
+Data along with the analysis script can be accessed here...
 
@@
 
-## Demographics
-The European Anaphylaxis Registry contained `r {data3[,"rANA"]=="yes"} %>% which %>% length` cases of refractory anaphylaxis and in `r {data3[,"rANA"]=="no"} %>% which %>% length` cases severe, non-refractory anaphlaxis was diagnosed.
-The mean age at the reaction was `r {data3$d_age[data3[,"rANA"]=="yes"]} %>% mean(na.rm=T) %>% signif(3)` ± `r data3$d_age[data3[,"rANA"]=="yes"] %>% sd(na.rm=T)` years, and did not differ from severe, non-refractory cases, p = `r t.test(data3$d_age[data3[,"rANA"]=="yes"],data3$d_age[data3[,"rANA"]=="no"])$p.value %>% signif(3)`).
+<!-- ## Demographics -->
+
+The European Anaphylaxis Registry contained `r {data3[,"rANA"]=="yes"} %>% which %>% length` cases of refractory anaphylaxis (reported in years 2007-2017) and in `r {data3[,"rANA"]=="no"} %>% which %>% length` cases severe, non-refractory anaphylaxis was diagnosed.
+The frequency of refractory anaphylaxis was therefore `r ((data3$rANA=="yes") %>% which() %>% length / (data3$d_year_reaction %in% 2007:2017) %>% which() %>% length ) %>% {.*100} %>% signif(2)`% of all anaphylaxis cases reported in the registry from European countries in years 2007 - 2017.
+It may be stated that each year `r table(data3$d_year_reaction, data3$rANA) %>% {cbind(.[,1]/(.[,1]+.[,2]),.[,2]/(.[,1]+.[,2]))[which(rownames(.) %in% 2007:2017),2]} %>% {paste0(mean(.) %>% signif(3)*100, "% $\\pm$ ", sd(.) %>% signif(3)*100,"%")}` of severe anaphylactic episodes are refractory to treatment with adrenaline.
+Moreover, if we consider patients undergoing a medical procedure (who undergo anaphylaxis in a perioperative setting), nearly `r (periopAna[2]/(periopAna[2]+periopAna[1])) %>% signif(3)*100`% (`r (periopAna[2]/(periopAna[2]+periopAna[1]) %>% signif(3)*100 / 0.853) %>% signif(3)` times more) patients present with reactions that do not respond to adrenaline.
+
+The percentage of males within the  refractory anaphylaxis group was `r (as.numeric(demoTab[2,2])/sum(as.numeric(demoTab[2,2])+as.numeric(demoTab[1,2]))*100) %>% {round(.,1)}`% and the mean age at the reaction was `r {data3$d_age[data3[,"rANA"]=="yes"]} %>% mean(na.rm=T) %>% signif(3)` ± `r data3$d_age[data3[,"rANA"]=="yes"] %>% sd(na.rm=T) %>% signif(3)` years, which did not differ from severe, non-refractory cases, p = `r t.test(data3$d_age[data3[,"rANA"]=="yes"],data3$d_age[data3[,"rANA"]=="no"])$p.value %>% signif(3)`).
 There were however significantly more cases of concomitant malignant diseases in the refractory anaphylaxis group.
-The demographic summary of refractory cases can be found in table \@ref(tab:demoTab).
+Patients with refractory reactions more frequently had a previous anaphylactic reaction in their medical history (p = `r data3$q_410_prev_v5 %>% table(data3$rANA) %>%{.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% signif(3)`).
+The demographic summary of refractory cases can be found in table \@ref(tab:demoTab)
 
-## Elicitors
+<!-- # ## Elicitors -->
+
+```{r}
+locTab <- data3$q_152_location %>% table(data3$rANA)
+ # data3$rANA[which(data3$q_152_location=="medical practice, hospital")] %>% as.factor %>%  summary()
+```
+
 Refractory anaphylaxis was most commonly elicited by drugs (significantly more often than in severe, non-refractory cases), followed by food and insects (table \@ref(tab:elicitorTab)).
+In `r elicitorTab[1,8]`% of food elicited cases the diagnosis of food allergy was previously confirmed.
+Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
+Patients with refractory anaphylaxis experienced the reaction while undergoing a medical procedure in `r (locTab[3,2]/sum(locTab[,2])*100) %>% {signif(.,3)}`% of the times (vs.
+`r (locTab[3,1]/sum(locTab[,1])*100) %>% {signif(.,3)}` in severe, non-refractory cases, p < 0.0001)
+
 The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics and X-ray contrast medium.
-Insect-venom refractory cases were significantly more often elicited by bee-stings than severe, non-refractory cases, although the yellow-jacket-stings were most common in other anaphylaxis cases (\@ref(tab:elicitExact)).
+Insect-venom-elicited, refractory cases were significantly more often elicited by bee-stings than severe, non-refractory cases, whereas the yellow-jacket-stings were most common in non-refractory anaphylaxis cases (table \@ref(tab:elicitExact)).
 
-## Therapy
-Adrenaline  iv. as a first-line treatment of anaphylaxis was significantly more often given in refractory cases (`r therapyTab[2,3]`% vs `r therapyTab[2,3]`%).
+<!-- ## Symptoms -->
+Milder anaphylaxis symptoms (i.e.
+pruritus, gastrointestinal symptoms, vertigo, chest and throat tightness) showed significantly more often in severe anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were associated more often with the refractory anaphylaxis cases.
+Table \@ref(tab:symptTab) summarizes the most prominent differences in anaphylaxis symptoms among both groups.
+
+<!-- ## Therapy -->
+When evaluating the therapeutic procedures, adrenaline iv.
+as a first-line treatment of anaphylaxis was significantly more often given in refractory cases (`r therapyTab[2,3]`% vs.
+`r therapyTab[1,3]`%, p < 0.0001) it has been also given more often as a second line therapy in refractory cases.
+
+```{r}
+med.time2 <-data3$q_530_adren2_time_in_min_v5 %>% as.character() %>% as.numeric %>% split(data3$rANA) %>% lapply(median, na.rm=T)
+```
+
+Median time to second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes).
+<!--  wilcox.test(data3$q_530_adren2_time_in_min_v5[data3$rANA=="yes"],data3$q_530_adren2_time_in_min_v5[data3$rANA=="no"])$p.value %>% signif(3)-->
+
+```{r}
+adr2prof <- rcalc("q_530_adren2_who_v5","professional")
+```
+
+In `r adr2prof[3]`% of refractory cases, the second dose of adrenaline was given by a professional (significantly more than in non-refractory cases `r adr2prof[2]`%, p < 0.0001).
+
+Corticosteroids iv.
+were the second most frequently administered group of drugs in refractory cases (as a first and second line treatment), outpacing antihistaminic drugs and volume replacement therapy, and were significantly more often given in refractory cases.
+Volume replacement therapy was given initially in `r therapyTab[4,3]`% of refractory cases but was sustained only in `r therapyTab[5,3]`% as the therapy progressed in the hospital environment.
+<!--(possible counfounding variable by patients who died!!! they did not receive second line therapy and might need to be discarded from the analysis)-->
+
+Second line medication like dopamine, glucagon and methylene blue were neglected in all refractory anaphylaxis cases as well as in severe non-refractory ones, however patients with refractory anaphylaxis were more frequently admitted to the hospital and treated at the ICUs.
+Table \@ref(tab:therapyTab) illustrated the therapy of refractory anaphylaxis cases.
+
+<!-- ## Death  -->
+Fatal reactions occurred after 30 minutes to 120 minutes - when the CPR was unsuccessful.
+In cases where patients responded to life support procedures, but failed to be reanimated due to post-resuscitative complication (e.g.
+hypoxic brain injury), death occured in the next 3-8 days.
+
+<!-- ## Cofactors -->
+Patients with refractory anaphylaxis more often had concomitant asthma and malignant diseases in the past.
+Also other unspecified concomitant conditions were significantly more often reported in refractory cases (although they were not explicitly stated in the questionnaire).
+It is worth noticing that concomitant cardiologic conditions, diabetes and mastocytosis were similarly frequent in both groups.
+
+Patients with refractory anaphylaxis more often used proton pump inhibitors and acetylsalicilic acid than patients with severe non-refractory anaphylaxis.
+Other (non-disclosed) medications were also more frequent in refractory cases.
+
+The intensity of of physical exercise exceeding the reaction was indifferent between groups, however the psychological burden was reported 3 times more frequently in refractory cases (see table \@ref(tab:cofactorsTab)).
+
+```{r}
+#data3$q_410_other_codisease_free[which(data3$rANA=="yes")]
+#rcalc("q_4212_exercise_intensity","moderate eg. vigorous housework or gardening")
+#data3$q_423_other_drug_free[which(data3$rANA=="yes")]
+
+```
 
@@
 
-# Conclusion
+## Key results
+
+### Frequency and patients at risk:
+The frequency of refractory anaphylaxis was less than 1% of severe anaphylaxis cases per year.
+This suggests that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapeutic strategy, commonly outlined in anaphylaxis management guidelines.
+Refractory anaphylaxis was most often iatrogenic: ie.
+elicited by drugs and occurring in a medical practice or a hospital.
+Perioperative cases of refractoriness have been previously published to be the most common in case series analysis [source].
+Therefore patients who undergo medical procedures and have a history of anaphylaxis should remain under special caution.
+
+### Elicitors
+Although drugs were the most commonly occurring elicitor, there were also relatively more refractory cases elicited by bee stings.
+Plausible explanations for this observation, might include either longer exposition to the allergen (longer stings predispose to deeper deposition of the venom and slower metabolizing) or more efficient vasodilation due to biochemical differences between wasp and bee venom.
+[insekt sting anaphylaxis, Golden, 2007].
+In such case - patients might benefit from using vasopressive medication.
+
+Antibiotics, although commonly triggering type IV hypersensitivity reactions, are rarely reported in the literature as the cause of a refractory anaphylaxis.
+Radio-contrast media (RCM) are however common in eliciting hypersensitivity episodes refractory to adrenaline.
+We saw multiple refractory reactions to antibiotics but in clinical practice, these might be overestimated, as patients in a perioperative setting receive multiple drugs along with antibiotics.
+
+### Cofactors 
+Although we have previously identified older age and male sex to be associated with more severe anaphylaxis, these factors seem to be less important when refractory anaphylaxis is considered.
+Asthma and malignant diseases in the medical history were more frequent in the refractory anaphylaxis group and the reason for this is unclear.
+It is however interesting that past malignant diseases were mostly recognized in male refractory anaphylaxis patients.
+
+After performing a thorough analysis of co-factors associated with an increased risk of severe anaphylaxis [@Worm2018] we saw that concomitant asthma rather had a protective effect.
+This association is probably non-linear as the present analysis indicated concomitant asthma as a risk factor for a refractory episode.
+The reason for this is unclear.
+Either patients receive already antihistaminic and mast cell blocking agents and their reaction to adrenaline is therefore restricted, or the episode elicits a bronchospasm that is refractory to adrenaline and therefore results in a prolonged anaphylactic episode.
+This would also clarify why we saw more respiratory distress symptoms in the refractory anaphylaxis cases.
+
+We have previously proven the role of acetysalicylic acid in increasing the risk of severe anaphylaxis.
+The role of proton pump inhibitors (PPI) however has not been identified as a co-factor, and there are only reports on PPI as elicitors of anaphylaxis.
+
+### adrenaline use inconsistent and too low
+After revising the therapeutic procedures we saw significantly more adrenaline use in the refractory cases, but it could be used even more extensively in these life-threatening reactions.
+This observation might indicate that in fact the emergency teams are using these life-saving measures too cautiously due to possible side effects.
+Methylene blue and vasopressors have been described to successfully restore refractory hypotension, but their actual use in anaphylaxis patients is scarce, although they are mentioned in recognized guidelines on the management of anaphylaxis.
+
+## Limitations and strength
+Low number of refractory cases prevented us from analyzing the data with more advanced statistical models, however, our analysis it is the first published report on over 30 refractory anaphylaxis cases.
+
+Definition  of refractory anaphylaxis is not universal, and confirming the refractory anaphylaxis based on the answers to our online questionnaire poses additional difficulties.
+If fatal reaction occurs before the second dose of adrenaline could be administered to the patient - it cannot be diagnosed as refractory (as we now require at least two doses of minimum 300 µg adrenaline).
+although it as well could be refractory.
+
+Comparison of refractory anaphylaxis to patients with severe anaphylaxis gave us the possibility to describe which patients have a higher risk of developing a refractory episode, however we can not answer the question which patients have the higher risk of developing anaphylaxis per se.
+
+## Conclusion
+
+Refractory anaphylaxis is a rare form of life-threatening hypersensitivity reaction with high mortality.
+We propose to increase the use of second line medication such as methylene blue, vasopressin and glucagon
+in cases where 2 doses of adrenaline did not result in rapid normalization of anaphylaxis symptoms.
 
 # Acknowledgements
+We thank the members of The European Anaphylaxis Registry.
+<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
 
-<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
 ##### pagebreak
@@
 ```{r demoTab}
-knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. ")
+knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. Age is represented by a mean value, other as fractions [%]")
 ```
@@
 ```{r elicitorTab}
-knitr::kable(elicitorTab, caption = "Summary of elicitors in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ",
+knitr::kable(elicitorTab, caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
+ANA - anaphylaxis, p-value derived from the Fisher exact test.",
              col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]"))
@@
 ```{r elicitExact}
-knitr::kable(cbind(elicitExact[,5],elicitExact[,2:4]), caption = "Summary of elicitors in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ",
+knitr::kable(cbind(elicitExact[,5],elicitExact[,2:4]), caption = "Summary of the elicitors broken down to specific elicitors fro meach group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control.
+ANA - anaphylaxis, p-value derived from the Fisher exact test.
+",
              col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value"))
@@
 
-* p-value derived from the Fischer exact test.
+p-value derived from the Fisher exact test.
+
+```{r symptTab}
+knitr::kable(symptTab[,1:4], caption = "Summary of the symptoms in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
+ANA - anaphylaxis, p-value derived from the Fisher exact test.
+",
+             col.names = c("Symptom","severe ANA [%]","refractory ANA [%]","p value"))
+```
 
 ```{r therapyTab}
-knitr::kable(therapyTab[,1:4], caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe anaphylaxis cases as a control. ",
+knitr::kable(therapyTab[,1:4], caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
+2nd line therapy was performed after the initial rescue individual called for a professional medical emergeny team.
+ANA - anaphylaxis, p-value derived from the Fisher exact test.
+",
              col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"))
@@
 
-* p-value derived from the Fischer exact test.
+```{r cofactorsTab}
+knitr::kable(cofactorsTabF[,1:4], caption = "Factors potentially increasing the risk of a severe anaphylaxis investigated in refractory cases.
+ANA - anaphylaxis, p-value derived from the Fisher exact test.
+",
+             col.names = c("Factor","severe ANA [%]","refractory ANA [%]","p value"))
+```

### 2018-08-15 R/analysis.R

@@ -50,7 +50,7 @@ f1 <- function(x){
   return(do.call(t.test,x)$p.value %>% signif(3))
 }
 f2 <- function(x){
-  x %>% table(data3$rANA) %>% {.[1:2,]} %>% summary() %>% {.$p.value} %>% signif(3)
+  x %>% table(data3$rANA) %>% {.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% signif(3)
 }
 f3 <- function(x){
   x %>%
@@ -213,8 +213,8 @@ data3 <- data
 # Patient has to get at least 2 doses of adrenalin and the reaction has to be either fatal or
 # refractory anaphylaxis is when you give adrenaline and the patient dies
 # Helpful variables in our case
-data3$q_521_autoinj_v5
-data3$q_521_other_v5 %>% summary
+#"q_521_autoinj_v5" %>% rcalc
+#"q_521_other_v5" %>% rcalc
 data3$q_521_other_free_v5 %>% summary # Mund-zu-Mund-Beatmung, ABC
 data3$q_522_adren_im %>% summary
 data3$q_522_adren_iv %>% summary
@@ -621,6 +621,27 @@ therapyTab <- rbind(
   rcalc("q_562_intensive_care_v6")
 )
 
+therapyTab[,1] <-
+  c("adrenaline im.",
+    "adrenaline iv.",
+    "adrenaline iv., 2nd line",
+    "volume",
+    "volume, 2nd line",
+    "antihistaminics iv.",
+    "antihistaminics iv. 2nd line",
+    "corticosteoids, all routes",
+    "corticosteroids iv.",
+    "corticosteroids iv. 2nd line",
+    "beta-2-mimetics iv.",
+    "beta-2-mimetics inh. 2nd line",
+    "theophyline iv.",
+    "100% oxygen",
+    "dopamine iv.",
+    "glucagon iv.",
+    "methylene blue",
+    "hospital admission",
+    "intensive care")
+
 #### 5. Fatal cases ####
 data3$q_140_fatal %>% table(data3$rANA)
 
@@ -633,9 +654,48 @@ elicitExact <- rbind(
   rcalc("q_340_insects", "yellow jacket")
 )
 
+#### 7. Symptoms tab ####
+sympt.all <- names(data3)[18:70] %>% lapply(rcalc) %>% do.call(what = rbind)
+symptTab <- sympt.all[c(4,6,14,16,18,21,22,24,26,29,30,31),] %>% rbind(rcalc("q_140_fatal"))
[…]

### 2018-08-22 analysis/paper/paper.Rmd
workday over changes to the discussion in progress

@@
 <!-- ## Demographics -->
+## Refractory anaphylaxis accounts for less than 0.5% of severe anaphylaxis cases in the register
 
-The European Anaphylaxis Registry contained `r {data3[,"rANA"]=="yes"} %>% which %>% length` cases of refractory anaphylaxis (reported in years 2007-2017) and in `r {data3[,"rANA"]=="no"} %>% which %>% length` cases severe, non-refractory anaphylaxis was diagnosed.
-The frequency of refractory anaphylaxis was therefore `r ((data3$rANA=="yes") %>% which() %>% length / (data3$d_year_reaction %in% 2007:2017) %>% which() %>% length ) %>% {.*100} %>% signif(2)`% of all anaphylaxis cases reported in the registry from European countries in years 2007 - 2017.
-It may be stated that each year `r table(data3$d_year_reaction, data3$rANA) %>% {cbind(.[,1]/(.[,1]+.[,2]),.[,2]/(.[,1]+.[,2]))[which(rownames(.) %in% 2007:2017),2]} %>% {paste0(mean(.) %>% signif(3)*100, "% $\\pm$ ", sd(.) %>% signif(3)*100,"%")}` of severe anaphylactic episodes are refractory to treatment with adrenaline.
-Moreover, if we consider patients undergoing a medical procedure (who undergo anaphylaxis in a perioperative setting), nearly `r (periopAna[2]/(periopAna[2]+periopAna[1])) %>% signif(3)*100`% (`r (periopAna[2]/(periopAna[2]+periopAna[1]) %>% signif(3)*100 / 0.853) %>% signif(3)` times more) patients present with reactions that do not respond to adrenaline.
+The European Anaphylaxis Registry captured `r {data3[,"rANA"]=="yes"} %>% which %>% length` cases of refractory anaphylaxis and `r {data3[,"rANA"]=="no"} %>% which %>% length` severe, non-refractory anaphylaxis.
+The frequency of refractory anaphylaxis was `r ((data3$rANA=="yes") %>% which() %>% length / (data3$d_year_reaction %in% 2007:2017) %>% which() %>% length ) %>% {.*100} %>% signif(2)`% of all anaphylaxis cases reported in the registry from European countries in years 2007 - 2017.
+Each year approximately 1% (`r table(data3$d_year_reaction, data3$rANA) %>% {cbind(.[,1]/(.[,1]+.[,2]),.[,2]/(.[,1]+.[,2]))[which(rownames(.) %in% 2007:2017),2]} %>% {paste0(mean(.) %>% signif(3)*100, "% $\\pm$ ", sd(.) %>% signif(3)*100,"%")}`) of severe anaphylactic episodes are refractory to treatment with adrenaline.
+When considering patients undergoing a medical procedure (who experience anaphylaxis in a perioperative setting), nearly `r (periopAna[2]/(periopAna[2]+periopAna[1])) %>% signif(3)*100`% (`r (periopAna[2]/(periopAna[2]+periopAna[1]) %>% signif(3)*100 / 0.853) %>% signif(3)` times more) patients present with reactions that do not respond to adrenaline.
 
-The percentage of males within the  refractory anaphylaxis group was `r (as.numeric(demoTab[2,2])/sum(as.numeric(demoTab[2,2])+as.numeric(demoTab[1,2]))*100) %>% {round(.,1)}`% and the mean age at the reaction was `r {data3$d_age[data3[,"rANA"]=="yes"]} %>% mean(na.rm=T) %>% signif(3)` ± `r data3$d_age[data3[,"rANA"]=="yes"] %>% sd(na.rm=T) %>% signif(3)` years, which did not differ from severe, non-refractory cases, p = `r t.test(data3$d_age[data3[,"rANA"]=="yes"],data3$d_age[data3[,"rANA"]=="no"])$p.value %>% signif(3)`).
-There were however significantly more cases of concomitant malignant diseases in the refractory anaphylaxis group.
-Patients with refractory reactions more frequently had a previous anaphylactic reaction in their medical history (p = `r data3$q_410_prev_v5 %>% table(data3$rANA) %>%{.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% signif(3)`).
-The demographic summary of refractory cases can be found in table \@ref(tab:demoTab)
+## Increased Frequency of previous reactions in patients with refractory anaphylaxis
+
+The mean age at the reaction was `r {data3$d_age[data3[,"rANA"]=="yes"]} %>% mean(na.rm=T) %>% signif(3)` ± `r data3$d_age[data3[,"rANA"]=="yes"] %>% sd(na.rm=T) %>% signif(3)` years, which did not differ from severe, non-refractory cases, p = `r t.test(data3$d_age[data3[,"rANA"]=="yes"],data3$d_age[data3[,"rANA"]=="no"])$p.value %>% signif(3)`).
+The percentage of males within the refractory anaphylaxis group was `r (as.numeric(demoTab[2,2])/sum(as.numeric(demoTab[2,2])+as.numeric(demoTab[1,2]))*100) %>% {round(.,1)}`%.
+Significantly more patients suffered from a concomitant malignant disease in the refractory anaphylaxis group.
+Most strikingly, patients with refractory reactions more often had a previous anaphylactic reaction in their medical history (p = `r data3$q_410_prev_v5 %>% table(data3$rANA) %>%{.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% signif(3)`).
+The demographic summary of refractory cases is shown in table \@ref(tab:demoTab).
 
 <!-- # ## Elicitors -->
+## Drugs are the most frequent elicitors of refractory anaphylaxis
 
@@
 Refractory anaphylaxis was most commonly elicited by drugs (significantly more often than in severe, non-refractory cases), followed by food and insects (table \@ref(tab:elicitorTab)).
-In `r elicitorTab[1,8]`% of food elicited cases the diagnosis of food allergy was previously confirmed.
-Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
-Patients with refractory anaphylaxis experienced the reaction while undergoing a medical procedure in `r (locTab[3,2]/sum(locTab[,2])*100) %>% {signif(.,3)}`% of the times (vs.
+The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics and X-ray contrast medium.
+Patients with refractory anaphylaxis more frequently experienced the reaction while undergoing a medical procedure (`r (locTab[3,2]/sum(locTab[,2])*100) %>% {signif(.,3)}`% vs.
 `r (locTab[3,1]/sum(locTab[,1])*100) %>% {signif(.,3)}` in severe, non-refractory cases, p < 0.0001)
 
-The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics and X-ray contrast medium.
+`r elicitorTab[1,8]`% of food elicited cases had a previously confirmed diagnosis of food allergy. Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
 Insect-venom-elicited, refractory cases were significantly more often elicited by bee-stings than severe, non-refractory cases, whereas the yellow-jacket-stings were most common in non-refractory anaphylaxis cases (table \@ref(tab:elicitExact)).
 
+## Refractory anaphylaxis is life-threathening
 <!-- ## Symptoms -->
@@
 Table \@ref(tab:symptTab) summarizes the most prominent differences in anaphylaxis symptoms among both groups.
+<!-- ## Death  -->
+Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the CPR was unsuccessful.
+In cases where patients responded to life support procedures, but failed to be reanimated due to post-resuscitative complication (e.g.
+hypoxic brain injury), death occured in the next 3-8 days.
 
 <!-- ## Therapy -->
+## Adrenaline i.v. as first-line treatment was given frequently in refractory anaphylaxis
+
+```{r}
+med.time2 <-data3$q_530_adren2_time_in_min_v5 %>% as.character() %>% as.numeric %>% split(data3$rANA) %>% lapply(median, na.rm=T)
+adr2prof <- rcalc("q_530_adren2_who_v5","professional")
+
+```
+
 When evaluating the therapeutic procedures, adrenaline iv.
@@
 `r therapyTab[1,3]`%, p < 0.0001) it has been also given more often as a second line therapy in refractory cases.
-
-```{r}
-med.time2 <-data3$q_530_adren2_time_in_min_v5 %>% as.character() %>% as.numeric %>% split(data3$rANA) %>% lapply(median, na.rm=T)
-```
-
-Median time to second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes).
+Median time to the second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes).
 <!--  wilcox.test(data3$q_530_adren2_time_in_min_v5[data3$rANA=="yes"],data3$q_530_adren2_time_in_min_v5[data3$rANA=="no"])$p.value %>% signif(3)-->
-
-```{r}
-adr2prof <- rcalc("q_530_adren2_who_v5","professional")
-```
-
 In `r adr2prof[3]`% of refractory cases, the second dose of adrenaline was given by a professional (significantly more than in non-refractory cases `r adr2prof[2]`%, p < 0.0001).
@@
 
-<!-- ## Death  -->
-Fatal reactions occurred after 30 minutes to 120 minutes - when the CPR was unsuccessful.
-In cases where patients responded to life support procedures, but failed to be reanimated due to post-resuscitative complication (e.g.
-hypoxic brain injury), death occured in the next 3-8 days.
-
+## Cofactors of refractory anaphylaxis
 <!-- ## Cofactors -->
@@
 
-### Frequency and patients at risk:
-The frequency of refractory anaphylaxis was less than 1% of severe anaphylaxis cases per year.
-This suggests that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapeutic strategy, commonly outlined in anaphylaxis management guidelines.
+### Frequency and patients at risk
+Our findings suggest that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapy with adrenaline, commonly outlined in anaphylaxis management guidelines.
 Refractory anaphylaxis was most often iatrogenic: ie.
 elicited by drugs and occurring in a medical practice or a hospital.
-Perioperative cases of refractoriness have been previously published to be the most common in case series analysis [source].
+
+### Elicitors
+Drugs were the most commonly occurring elicitor, what is in concordance with our previous findings [@Francuzik2018].
+Multiple medications given in a perioperative setting increase the chance of a hypersensitivity reaction and drug interactions.
+Perioperative cases of refractoriness have been previously published to be the most common in case series analysis [@Francuzik2018].
 Therefore patients who undergo medical procedures and have a history of anaphylaxis should remain under special caution.
 
-### Elicitors
-Although drugs were the most commonly occurring elicitor, there were also relatively more refractory cases elicited by bee stings.
+Antibiotics, although commonly triggering type IV hypersensitivity reactions, are rarely reported in the literature as the cause of a refractory anaphylaxis.
+We saw multiple refractory reactions to antibiotics but in clinical practice, these might be overestimated, as patients in a perioperative setting receive multiple drugs along with antibiotics.
+
+Radio-contrast media (RCM) are however common in eliciting hypersensitivity episodes refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [Greenberger1991].
+One of the plausible explanation to this is that RCM promote CARPA and pseudoallergic reactions which might depend on a different mechanism than anaphylaxis [@Wang2011].
+
+Literature describes many cases of hypersenitivity due to cancer drugs.
+We saw more cases of refractory anaphylaxis in patient who reported having a malignant disease in the past.
+Is it possible that the cancer drugs made them more susceptible to refractory anaphylaxis?
+```{r}
+temp <- data3[data3$rANA=="yes",]
+mean_age_malig <- data3$d_age %>% split(data3$q_410_malig_prev_v5 %>% factor) %>% lapply(mean)
+```
+It also might be that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]]` vs `r mean_age_malig[[2]]` p < ) and therefore less able to compensate the anaphylactic shock.
+
+#### Bees vs Wasps
+There were also relatively less refractory cases elicited by yellow jacket stings (but not bee-stings).
+The question we could ask here is if chemical properties of bee and yellow jacket venom have various affinities to mast cells?
 Plausible explanations for this observation, might include either longer exposition to the allergen (longer stings predispose to deeper deposition of the venom and slower metabolizing) or more efficient vasodilation due to biochemical differences between wasp and bee venom.
@@
 
-Antibiotics, although commonly triggering type IV hypersensitivity reactions, are rarely reported in the literature as the cause of a refractory anaphylaxis.
-Radio-contrast media (RCM) are however common in eliciting hypersensitivity episodes refractory to adrenaline.
-We saw multiple refractory reactions to antibiotics but in clinical practice, these might be overestimated, as patients in a perioperative setting receive multiple drugs along with antibiotics.
+Mastoparans are a potent family of mast cell activators through G-protein coupled receptors (GPCRs).
+It is likely that their affinity to GPCRs might vary and the composition of mastoparans might depend on the species of the
 
@@
 
+RCM was the most frequent elicitor of refractory anaphylaxis.
+
 ### adrenaline use inconsistent and too low
@@
 Methylene blue and vasopressors have been described to successfully restore refractory hypotension, but their actual use in anaphylaxis patients is scarce, although they are mentioned in recognized guidelines on the management of anaphylaxis.
+Adrenaline i.v.
+as first-line therapy use was higher in the refractory cases, probably due to most of them happening in a medical setting, however, other second-line therapeutic options were rarely used.
 
@@
 ## Conclusion
+
+Refractoriness may be described in three mechanisms:
+1. Lack of the natural ability to compensate anaphylactic shock.
+2. Specific trigger or risk factor which is avoidable or tangible- beta blockers, RCM, 
+3. Lack of proper managment of the episode leading to its refractoriness.
+
+All of these three mechanisms are interconnecd and influece the outcome of a reaction. 
+<Figure like the Virchow's TRIAD>
 
@@
 ```{r elicitExact}
-knitr::kable(cbind(elicitExact[,5],elicitExact[,2:4]), caption = "Summary of the elicitors broken down to specific elicitors fro meach group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control.
+knitr::kable(cbind(elicitExact[,5],elicitExact[,2:4]), caption = "Summary of the elicitors broken down to specific elicitors from each group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control.
 ANA - anaphylaxis, p-value derived from the Fisher exact test.

### 2018-08-22 R/analysis.R

@@ -37,8 +37,8 @@ rcalc <- function(var,level="yes"){
     pp <- tab[2,"yes"]
     l <- levels(data3[,var])[2]
   }
-  nn <- sum(tab[,"no"])-pn
-  np <- sum(tab[,"yes"])-pp
+  nn <- data3$rANA %>% factor %>% summary %>% {.[1]}-pn
+  np <- data3$rANA %>% factor %>% summary %>% {.[2]}-pp
   eval <- c(pn,pp,nn,np) %>% matrix(byrow = T,ncol = 2)
   prt <- (prop.table(eval,2)*100) %>% {signif(.,3)}
   p <- eval %>% fisher.test() %>% .$p.value %>% signif(3)
@@ -649,6 +649,7 @@ data3$q_140_fatal %>% table(data3$rANA)
 elicitExact <- rbind(
   rcalc("d_330_drug_group","antibiotics"),
   rcalc("d_330_drug_group","xray_cm"),
+  rcalc("d_330_drug_group","muscle relaxant"),
   rcalc("d_320_food_group", "legumes"),
   rcalc("q_340_insects", "bee"),
   rcalc("q_340_insects", "yellow jacket")

### 2018-08-27 analysis/paper/paper.Rmd
discussion

@@
 
-## Key results
+## Frequency and patients at risk
+Our findings suggest that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapy with adrenaline, commonly outlined in anaphylaxis management guidelines.
+Such cases have the highest risk of a fatal outcome, and therefore need to be treated adequately.
 
-### Frequency and patients at risk
-Our findings suggest that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapy with adrenaline, commonly outlined in anaphylaxis management guidelines.
-Refractory anaphylaxis was most often iatrogenic: ie.
-elicited by drugs and occurring in a medical practice or a hospital.
+Overall, the severity of anaphylaxis and its probable transition into a refractory episode might be described as in the figure \@ref(fig:theory).
+Patients who undergo anaphylaxis have four main aspects that influence the severity of a particular episode: 1) compensation - how well is the body able to compensate symptoms i.e.
+hypotension; 2) elicitors - what was the eliciting agent and duration of the exposition; 3) cofactors - is the presence of other factors which may increase the severity of a given reaction i.e.
+concomitant use of betablocker etc.; 4) therapy - how fast and accurate the therapeutic procedures were introduced.
+<!-- This theory might explain why younger patients are able to better compensate the same trigger which would elicit a refractory reaction in eldery patients.  -->
 
-### Elicitors
+## Compensation
+Patients who underwent refractory anahpylaxis more often had perioperative reactions  and therefore a decreased ablity to compensate the reaction symptoms.
+Age is also a factor that may correlate with the inabillity to retain homeostasis as many comorbidities are more likely to occure along with older age.
+
+## Elicitors
 Drugs were the most commonly occurring elicitor, what is in concordance with our previous findings [@Francuzik2018].
@@
 Radio-contrast media (RCM) are however common in eliciting hypersensitivity episodes refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [Greenberger1991].
-One of the plausible explanation to this is that RCM promote CARPA and pseudoallergic reactions which might depend on a different mechanism than anaphylaxis [@Wang2011].
+RCM was the most frequent elicitor of refractory anaphylaxis.
+One of the plausible explanation to this is that RCM promote CARPA and pseudoallergic reactions which might depend on a different mechanism than anaphylaxis [@Wang2011], and therfore the treatment with adrenaline may be less effective in these reactions.
 
@@
 Is it possible that the cancer drugs made them more susceptible to refractory anaphylaxis?
+
 ```{r}
@@
 ```
+
 It also might be that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]]` vs `r mean_age_malig[[2]]` p < ) and therefore less able to compensate the anaphylactic shock.
 
-#### Bees vs Wasps
 There were also relatively less refractory cases elicited by yellow jacket stings (but not bee-stings).
@@
 [insekt sting anaphylaxis, Golden, 2007].
-In such case - patients might benefit from using vasopressive medication.
+In such case - patients might benefit from vasopressive medication [source].
 
 Mastoparans are a potent family of mast cell activators through G-protein coupled receptors (GPCRs).
-It is likely that their affinity to GPCRs might vary and the composition of mastoparans might depend on the species of the
+It is likely that their affinity to GPCRs might vary and the composition of mastoparans might depend on the species [source].
 
-### Cofactors 
+## Cofactors 
 Although we have previously identified older age and male sex to be associated with more severe anaphylaxis, these factors seem to be less important when refractory anaphylaxis is considered.
@@
 This association is probably non-linear as the present analysis indicated concomitant asthma as a risk factor for a refractory episode.
-The reason for this is unclear.
 Either patients receive already antihistaminic and mast cell blocking agents and their reaction to adrenaline is therefore restricted, or the episode elicits a bronchospasm that is refractory to adrenaline and therefore results in a prolonged anaphylactic episode.
@@
 
-RCM was the most frequent elicitor of refractory anaphylaxis.
+## Therapy 
+After revising the therapeutic procedures we saw significantly more adrenaline use in the refractory cases, but it could have been used even more extensively in these life-threatening reactions.
+This observation might indicate that in fact the emergency teams are using these life-saving measures too cautiously.
 
-### adrenaline use inconsistent and too low
-After revising the therapeutic procedures we saw significantly more adrenaline use in the refractory cases, but it could be used even more extensively in these life-threatening reactions.
-This observation might indicate that in fact the emergency teams are using these life-saving measures too cautiously due to possible side effects.
 Methylene blue and vasopressors have been described to successfully restore refractory hypotension, but their actual use in anaphylaxis patients is scarce, although they are mentioned in recognized guidelines on the management of anaphylaxis.
@@
 ## Conclusion
-
-Refractoriness may be described in three mechanisms:
-1. Lack of the natural ability to compensate anaphylactic shock.
-2. Specific trigger or risk factor which is avoidable or tangible- beta blockers, RCM, 
-3. Lack of proper managment of the episode leading to its refractoriness.
-
-All of these three mechanisms are interconnecd and influece the outcome of a reaction. 
-<Figure like the Virchow's TRIAD>
 
@@
 knitr::include_graphics("../figures/flow.png")
+```
+
+```{r theory, fig.cap="Visual representation of the three forces determining the severity of anaphylaxis.
+The natural ability of the body to compensate the anaphylaxis symptoms and therapeutic measures act to restore homeostasis to a patient by whom elicitors and cofactors influence synergistically the severity of a given episode."}
+
+knitr::include_graphics("../figures/diag.png")
 ```

### 2018-08-29 analysis/paper/paper.Rmd
corrections from Sabine introduced

@@
 **Keywords:** 
-anaphylaxis, adrenaline (epinephrine), beta-blockers, Insect venom allergy, Drug allergic reactions
+anaphylaxis, adrenaline (epinephrine), beta-blockers, insect venom allergy, drug allergic reactions, vasopression, refractory
 
@@
 Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we were able to identify cases of refractory anaphylaxis (n = `r v$casesR`) and to analyse these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
+
 The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).
 Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r elicitExact[4,3]`% vs `r elicitExact[4,2]`%, p = `r elicitExact[4,4]`).
-The refractory cases occurred more frequently in a perioperative setting E(`r v$perioperative[3]`% vs.
+The refractory cases occurred more frequently in a perioperative setting (`r v$perioperative[3]`% vs.
 `r v$perioperative[2]`, p < 0.0001).
@@
 The mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).
-Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these standards can be improved.
+
+Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these guidelines can be improved.
 The low use of intramuscular adrenaline as a first line therapy might contribute to refractoriness of an episode.
-Also, the lack of suggested second line medication use in refractory cases indicate that the availability of these drug is restricted or anaphylaxis management guidelines may need a revision in order to provide clear algorithms for severe refractory anaphylaxis.
+Also, the lack of suggested second line medication use in refractory cases indicate that the availability of these drugs is either restricted or anaphylaxis management guidelines may need to be revised to provide clear algorithms for severe refractory anaphylaxis.
 
@@
 
-Anaphylaxis is a non-homogeneous clinical diagnosis, depending on various triggering, and augmenting factors.
-This variability introduces a wide range of possible reaction-symptom severity.
+Anaphylaxis is a non-homogeneous clinical diagnosis, depending on various triggering and augmenting factors.
+This variability introduces a wide range of possible reaction-symptom severities.
 Therefore multiple sub-types of anaphylaxis have been previously identified (i.e.
@@
 The mainstay of anaphylaxis management is the intramuscular dose of adrenaline, but in the most severe cases of anaphylaxis it might be insufficient to restore a stable patient status.
-Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least 2 doses of minimum 300 mg adrenaline, does not result in an expected normalization of clinical symptoms.
-Typical elicitors and symptoms of refractory anaphylaxis, as well as the therapeutic strategy for the most severe cases differ from the usual reactions and call for a specific research and targeted guideline development for refractory anaphylaxis cases.
+Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least 2 doses of minimum 300 mg adrenaline, does not respond with an expected normalization of clinical symptoms.
+Typical elicitors and symptoms of refractory anaphylaxis, as well as the therapeutic strategy for the most severe cases, differ from the usual reactions [@Francuzik2018] and call for a specific research and targeted guideline development for refractory anaphylaxis cases.
 
@@
 The European Anaphylaxis Registry (described in detail elsewhere [@Grabenhenrich2016]) provided data for this analysis status from May 2018.
-Based on  an electronic questionnaire (available at www.anaphylaxie.net), we selected cases where patients received at least two doses of adrenaline, and failed to recover adequately and assigned them to the "refractory anaphylaxis group".
+We selected cases where patients received at least two doses of adrenaline, and failed to recover adequately and assigned them to the "refractory anaphylaxis group".
 The flowchart in figure \@ref(fig:flowchart) represents the detailed cases selection process.
 
-The final database consisted of `r sum(countries)` cases of refractory anaphylaxis from `r length(countries)` countries: `r paste0(names(countries),": ",countries,collapse=", ")`.
+The final database consisted of `r sum(countries)` cases of refractory anaphylaxis from `r length(countries)` countries: `r paste0(names(countries[order(countries,decreasing = T)]),": ",countries[order(countries,decreasing = T)],collapse=", ")`.
 We compared these to a group of severe, non-refractory cases of anaphylaxis.
-Severe reactions were identified based on the definition of by NIAID/FAAN [@Sampson2006] and presented with significant hypoxia, cyanosis, hypotension (SBP < 90 mm Hg in adults), confusion, collapse, loss of conciusness, or incontinence.
+Severe reactions were identified based on the definition by NIAID/FAAN [@Sampson2006] and presented with significant hypoxia, hypotension, confusion, collapse, loss of consciousness, or incontinence.
 We compared the frequency of various elicitors, symptoms, and factors known to incease the risk of severe anphylaxis [@Worm2018] in both groups. We also compared how the cases were managed.
@@
 We defined statistical significance as α = 0.05.
-Data along with the analysis script can be accessed here...
+Data along with the analysis script can be accessed at github.com/wolass/RefractoryAnaOrg.
 
@@
 The European Anaphylaxis Registry captured `r {data3[,"rANA"]=="yes"} %>% which %>% length` cases of refractory anaphylaxis and `r {data3[,"rANA"]=="no"} %>% which %>% length` severe, non-refractory anaphylaxis.
-The frequency of refractory anaphylaxis was `r ((data3$rANA=="yes") %>% which() %>% length / (data3$d_year_reaction %in% 2007:2017) %>% which() %>% length ) %>% {.*100} %>% signif(2)`% of all anaphylaxis cases reported in the registry from European countries in years 2007 - 2017.
+The frequency of refractory anaphylaxis was `r ((data3$rANA=="yes") %>% which() %>% length / (data3$d_year_reaction %in% 2007:2017) %>% which() %>% length ) %>% {.*100} %>% signif(2)`% of all anaphylaxis cases reported in the registry.
 Each year approximately 1% (`r table(data3$d_year_reaction, data3$rANA) %>% {cbind(.[,1]/(.[,1]+.[,2]),.[,2]/(.[,1]+.[,2]))[which(rownames(.) %in% 2007:2017),2]} %>% {paste0(mean(.) %>% signif(3)*100, "% $\\pm$ ", sd(.) %>% signif(3)*100,"%")}`) of severe anaphylactic episodes are refractory to treatment with adrenaline.
@@
 Refractory anaphylaxis was most commonly elicited by drugs (significantly more often than in severe, non-refractory cases), followed by food and insects (table \@ref(tab:elicitorTab)).
-The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics and X-ray contrast medium.
+The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics (`r elicitExact[1,3]`%) and Radio-contrast media (RCM, `r elicitExact[2,3]`%).
 Patients with refractory anaphylaxis more frequently experienced the reaction while undergoing a medical procedure (`r (locTab[3,2]/sum(locTab[,2])*100) %>% {signif(.,3)}`% vs.
@@
 `r elicitorTab[1,8]`% of food elicited cases had a previously confirmed diagnosis of food allergy. Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
-Insect-venom-elicited, refractory cases were significantly more often elicited by bee-stings than severe, non-refractory cases, whereas the yellow-jacket-stings were most common in non-refractory anaphylaxis cases (table \@ref(tab:elicitExact)).
+Insect-venom-elicited, refractory cases were significantly less often elicited by yellow-jacket-stings (but not by bee-stings) than severe, non-refractory cases (table \@ref(tab:elicitExact)).
 
@@
 Milder anaphylaxis symptoms (i.e.
-pruritus, gastrointestinal symptoms, vertigo, chest and throat tightness) showed significantly more often in severe anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were associated more often with the refractory anaphylaxis cases.
+pruritus, gastrointestinal symptoms, vertigo, chest and throat tightness) were significantly less often present in refractory anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were more often associated with the refractory anaphylaxis cases.
 Table \@ref(tab:symptTab) summarizes the most prominent differences in anaphylaxis symptoms among both groups.
 <!-- ## Death  -->
-Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the CPR was unsuccessful.
+Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the ardiopulmonary resuscitation was unsuccessful.
 In cases where patients responded to life support procedures, but failed to be reanimated due to post-resuscitative complication (e.g.
@@
 med.time2 <-data3$q_530_adren2_time_in_min_v5 %>% as.character() %>% as.numeric %>% split(data3$rANA) %>% lapply(median, na.rm=T)
+med.timep<- data3$q_530_adren2_time_in_min_v5 %>% as.character() %>% as.numeric %>% 
+  {wilcox.test(.[data3$rANA=="no"],.[data3$rANA=="yes"])}
 adr2prof <- rcalc("q_530_adren2_who_v5","professional")
@@
 `r therapyTab[1,3]`%, p < 0.0001) it has been also given more often as a second line therapy in refractory cases.
-Median time to the second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes).
+Median time to the second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes vs.
+15 minutes in non-refractory cases, p = `r med.timep$p.value %>% signif(3)`).
 <!--  wilcox.test(data3$q_530_adren2_time_in_min_v5[data3$rANA=="yes"],data3$q_530_adren2_time_in_min_v5[data3$rANA=="no"])$p.value %>% signif(3)-->
-In `r adr2prof[3]`% of refractory cases, the second dose of adrenaline was given by a professional (significantly more than in non-refractory cases `r adr2prof[2]`%, p < 0.0001).
+<!--In `r adr2prof[3]`% of refractory cases, the second dose of adrenaline was given by a professional (significantly more than in non-refractory cases `r adr2prof[2]`%, p < 0.0001).  -->
 
-Corticosteroids iv.
+Corticosteroids i.v.
 were the second most frequently administered group of drugs in refractory cases (as a first and second line treatment), outpacing antihistaminic drugs and volume replacement therapy, and were significantly more often given in refractory cases.
@@
 Second line medication like dopamine, glucagon and methylene blue were neglected in all refractory anaphylaxis cases as well as in severe non-refractory ones, however patients with refractory anaphylaxis were more frequently admitted to the hospital and treated at the ICUs.
-Table \@ref(tab:therapyTab) illustrated the therapy of refractory anaphylaxis cases.
+Table \@ref(tab:therapyTab) illustrates the therapy of refractory anaphylaxis cases.
 
@@
 Patients with refractory anaphylaxis more often had concomitant asthma and malignant diseases in the past.
-Also other unspecified concomitant conditions were significantly more often reported in refractory cases (although they were not explicitly stated in the questionnaire).
+Also other unspecified concomitant conditions were significantly more often reported in refractory cases.
 It is worth noticing that concomitant cardiologic conditions, diabetes and mastocytosis were similarly frequent in both groups.
 
-Patients with refractory anaphylaxis more often used proton pump inhibitors and acetylsalicilic acid than patients with severe non-refractory anaphylaxis.
-Other (non-disclosed) medications were also more frequent in refractory cases.
+Patients with refractory anaphylaxis more often used proton pump inhibitors (PPI) and acetylsalicilic acid (ASA) than patients with severe non-refractory anaphylaxis.
+Other (not specified) medications were also more frequent in refractory cases.
 
-The intensity of of physical exercise exceeding the reaction was indifferent between groups, however the psychological burden was reported 3 times more frequently in refractory cases (see table \@ref(tab:cofactorsTab)).
+The intensity of physical exercise exceeding the reaction was indifferent between groups, however the psychological burden was reported 3 times more frequently in refractory cases (see table \@ref(tab:cofactorsTab)).
 
@@
 ## Frequency and patients at risk
-Our findings suggest that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapy with adrenaline, commonly outlined in anaphylaxis management guidelines.
+Our findings suggest that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapy with adrenaline, commonly outlined in anaphylaxis management guidelines [@Muraro2014].
 Such cases have the highest risk of a fatal outcome, and therefore need to be treated adequately.
@@
 Overall, the severity of anaphylaxis and its probable transition into a refractory episode might be described as in the figure \@ref(fig:theory).
-Patients who undergo anaphylaxis have four main aspects that influence the severity of a particular episode: 1) compensation - how well is the body able to compensate symptoms i.e.
+Patients who undergo anaphylaxis have four main aspects that influence the severity of a particular episode: 1) compensation - how well is the body able to compensate reaction symptoms e.g.
 hypotension; 2) elicitors - what was the eliciting agent and duration of the exposition; 3) cofactors - is the presence of other factors which may increase the severity of a given reaction i.e.
@@
 
-## Compensation
+## Symptom compensation
+Age is the most important factor influencing the risk of developing severe anaphylaxis [@Worm2018].
+It has been shown that older age may correlate with the decreased abillity to retain homeostasis on strain [@fleg1995].
 Patients who underwent refractory anahpylaxis more often had perioperative reactions  and therefore a decreased ablity to compensate the reaction symptoms.
-Age is also a factor that may correlate with the inabillity to retain homeostasis as many comorbidities are more likely to occure along with older age.
 
@@
 
-Radio-contrast media (RCM) are however common in eliciting hypersensitivity episodes refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [Greenberger1991].
+RCM however, commonly elicit hypersensitivity episodes refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [@Greenberger1991].
 RCM was the most frequent elicitor of refractory anaphylaxis.
@@
 
-## Cofactors 
-Although we have previously identified older age and male sex to be associated with more severe anaphylaxis, these factors seem to be less important when refractory anaphylaxis is considered.
+## Risk factors 
+Although we have previously identified older age and male sex to be associated with more severe anaphylaxis [@Worm2018], these factors seem to be less important when refractory anaphylaxis is considered.
 Asthma and malignant diseases in the medical history were more frequent in the refractory anaphylaxis group and the reason for this is unclear.
@@
 
-After performing a thorough analysis of co-factors associated with an increased risk of severe anaphylaxis [@Worm2018] we saw that concomitant asthma rather had a protective effect.
-This association is probably non-linear as the present analysis indicated concomitant asthma as a risk factor for a refractory episode.
+We previously saw that concomitant asthma was rather associated with less severe  anaphylaxis [@Worm2018] but the present analysis indicated concomitant asthma as a risk factor for a refractory episode.
 Either patients receive already antihistaminic and mast cell blocking agents and their reaction to adrenaline is therefore restricted, or the episode elicits a bronchospasm that is refractory to adrenaline and therefore results in a prolonged anaphylactic episode.
@@
 
-We have previously proven the role of acetysalicylic acid in increasing the risk of severe anaphylaxis.
-The role of proton pump inhibitors (PPI) however has not been identified as a co-factor, and there are only reports on PPI as elicitors of anaphylaxis.
+<!-- We have previously proven the role of ASA in increasing the risk of severe anaphylaxis [@].  Could not find the reference -->
+The role of PPI however has not been identified as a risk factor, and there are only reports on PPI as elicitors of anaphylaxis.
+
+<!--Low-Dose Ethanol Alters the Cardiovascular, Metabolic, and Respiratory Compensation for Severe Blood Loss-->
 
 ## Therapy 
-After revising the therapeutic procedures we saw significantly more adrenaline use in the refractory cases, but it could have been used even more extensively in these life-threatening reactions.
+<!-- After revising the therapeutic procedures we saw significantly more adrenaline use in the refractory cases, but it could have been used even more extensively in these life-threatening reactions.
 This observation might indicate that in fact the emergency teams are using these life-saving measures too cautiously.
+-->
 
-Methylene blue and vasopressors have been described to successfully restore refractory hypotension, but their actual use in anaphylaxis patients is scarce, although they are mentioned in recognized guidelines on the management of anaphylaxis.
 Adrenaline i.v.
@@
 
+Methylene blue and vasopressors have been described to successfully restore refractory hypotension, but their actual use in anaphylaxis patients is scarce, although they are mentioned in recognized guidelines on the management of anaphylaxis [@Muraro2014].
+
 ## Limitations and strength
-Low number of refractory cases prevented us from analyzing the data with more advanced statistical models, however, our analysis it is the first published report on over 30 refractory anaphylaxis cases.
+Low number of refractory cases prevented us from analyzing the data with more advanced statistical models, however, our analysis is the first published report on over 30 refractory anaphylaxis cases.
 
 Definition  of refractory anaphylaxis is not universal, and confirming the refractory anaphylaxis based on the answers to our online questionnaire poses additional difficulties.
-If fatal reaction occurs before the second dose of adrenaline could be administered to the patient - it cannot be diagnosed as refractory (as we now require at least two doses of minimum 300 µg adrenaline).
-although it as well could be refractory.
+If fatal reaction occurs before the second dose of adrenaline could be administered to the patient - it cannot be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline) although it as well could be refractory.
 
@@
 ```{r demoTab}
-knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. Age is represented by a mean value, other as fractions [%]")
+knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. Age is represented by a mean value, other as fractions [%]. DM - diabetes mellitus",)
 ```
@@
 ANA - anaphylaxis, p-value derived from the Fisher exact test.",
-             col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]"))
+             col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]"),
+             digits = 1)
 ```

### 2018-08-29 R/analysis.R

@@ -54,7 +54,7 @@ f2 <- function(x){
 }
 f3 <- function(x){
   x %>%
-    lapply(function(x){(length(which(x=="yes"))/length(x)*100) %>% signif(3)})
+    lapply(function(x){(length(which(x=="yes"))/length(x)*100) %>% round(0)})
 }
 
 
@@ -562,18 +562,16 @@ countries <- rdb$d_centres_country %>% summary() %>% .[{which(.!=0)}]
 
 #rdb[rdb$d_severity_rm=="grade II",]
 
-
-
 #### 2. Elicitor tab ####
 
 elicitorTab <- cbind(n = rdb$d_elicitor_gr5 %>% summary(),
-                     percent = rdb$d_elicitor_gr5 %>% {summary(.)/42*100} %>% signif(3),
-                     percANA = control$d_elicitor_gr5 %>% {summary(.)/length(.)*100} %>% signif(3),
+                     percent = rdb$d_elicitor_gr5 %>% {summary(.)/42*100} %>% round(1),
+                     percANA = control$d_elicitor_gr5 %>% {summary(.)/length(.)*100} %>% round(1),
                      p = c(rcalc("d_elicitor_gr5","food")[4],
                            rcalc("d_elicitor_gr5","drugs")[4],
                            rcalc("d_elicitor_gr5","insects")[4],
                            rcalc("d_elicitor_gr5","other")[4],
-                           rcalc("d_elicitor_gr5","unknown")[4]),
+                           rcalc("d_elicitor_gr5","unknown")[4]) %>% as.numeric(),
                  Age = rdb$d_age %>% split(rdb$d_elicitor_gr5) %>%
                    lapply(.,function(x){mean(x) %>% signif(3)}),
                  Male = rdb$b_sex%>% split(rdb$d_elicitor_gr5) %>%
@@ -622,22 +620,22 @@ therapyTab <- rbind(
 )
 
 therapyTab[,1] <-
-  c("adrenaline im.",
-    "adrenaline iv.",
-    "adrenaline iv., 2nd line",
+  c("adrenaline i.m.",
+    "adrenaline i.v.",
+    "adrenaline i.v. 2nd line",
     "volume",
     "volume, 2nd line",
-    "antihistaminics iv.",
-    "antihistaminics iv. 2nd line",
+    "antihistaminics i.v.",
+    "antihistaminics i.v. 2nd line",
     "corticosteoids, all routes",
[…]

### 2018-08-30 analysis/paper/paper.Rmd
discussion half done

@@
 Age is the most important factor influencing the risk of developing severe anaphylaxis [@Worm2018].
-It has been shown that older age may correlate with the decreased abillity to retain homeostasis on strain [@fleg1995].
-Patients who underwent refractory anahpylaxis more often had perioperative reactions  and therefore a decreased ablity to compensate the reaction symptoms.
+It has been shown that older age may correlate with the decreased abillity to retain homeostasis on increased strain [@fleg1995].
+Patients who underwent refractory anahpylaxis more often had perioperative reactions  and therefore a decreased ablity to compensate the reaction symptoms with reflexory renal or cardiopulmonary response [@Rogers2013].
 
@@
 Drugs were the most commonly occurring elicitor, what is in concordance with our previous findings [@Francuzik2018].
-Multiple medications given in a perioperative setting increase the chance of a hypersensitivity reaction and drug interactions.
-Perioperative cases of refractoriness have been previously published to be the most common in case series analysis [@Francuzik2018].
-Therefore patients who undergo medical procedures and have a history of anaphylaxis should remain under special caution.
+Multiple medications given in a perioperative setting increase the chance of a hypersensitivity reaction and drug interactions [@Ebo2007].
+Therefore patients who undergo medical procedures and have a history of anaphylaxis should remain under extraordinary caution.
 
-Antibiotics, although commonly triggering type IV hypersensitivity reactions, are rarely reported in the literature as the cause of a refractory anaphylaxis.
+Antibiotics, although commonly triggering pseudoallergic reactions[@saxon1983], are rarely reported in the literature as the cause of a refractory anaphylaxis.
 We saw multiple refractory reactions to antibiotics but in clinical practice, these might be overestimated, as patients in a perioperative setting receive multiple drugs along with antibiotics.
@@
 RCM however, commonly elicit hypersensitivity episodes refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [@Greenberger1991].
-RCM was the most frequent elicitor of refractory anaphylaxis.
-One of the plausible explanation to this is that RCM promote CARPA and pseudoallergic reactions which might depend on a different mechanism than anaphylaxis [@Wang2011], and therfore the treatment with adrenaline may be less effective in these reactions.
-
-Literature describes many cases of hypersenitivity due to cancer drugs.
-We saw more cases of refractory anaphylaxis in patient who reported having a malignant disease in the past.
-Is it possible that the cancer drugs made them more susceptible to refractory anaphylaxis?
-
-```{r}
-temp <- data3[data3$rANA=="yes",]
-mean_age_malig <- data3$d_age %>% split(data3$q_410_malig_prev_v5 %>% factor) %>% lapply(mean)
-```
-
-It also might be that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]]` vs `r mean_age_malig[[2]]` p < ) and therefore less able to compensate the anaphylactic shock.
+We previously saw RCM to be the most frequent elicitor of refractory anaphylaxis [@Francuzik2018].
+One of the plausible explanation to this is that RCM promote complement activation related pseudo allergy (CARPA) [@Wang2011], and therfore the treatment with adrenaline may be less effective in these reactions.
 
 There were also relatively less refractory cases elicited by yellow jacket stings (but not bee-stings).
-The question we could ask here is if chemical properties of bee and yellow jacket venom have various affinities to mast cells?
-Plausible explanations for this observation, might include either longer exposition to the allergen (longer stings predispose to deeper deposition of the venom and slower metabolizing) or more efficient vasodilation due to biochemical differences between wasp and bee venom.
-[insekt sting anaphylaxis, Golden, 2007].
-In such case - patients might benefit from vasopressive medication [source].
+It means that the ratio of yellow jacket to honey bee venom elicited anaphylaxis was inverted in the refractory group.
+Plausible explanations for this observation might include either longer exposition to the allergen (bee's sting remains in the skin) or the venom volume being 10 times higher than in a wasp sting.
+Patients should remove the bee sting as quickly as possible (optimal < 2 seconds) [@Visscher1996].
 
-Mastoparans are a potent family of mast cell activators through G-protein coupled receptors (GPCRs).
+<!-- Mastoparans are a potent family of mast cell activators through G-protein coupled receptors (GPCRs).
 It is likely that their affinity to GPCRs might vary and the composition of mastoparans might depend on the species [source].
+-->
 
@@
 It is however interesting that past malignant diseases were mostly recognized in male refractory anaphylaxis patients.
+
+Literature describes many cases of hypersenitivity due to cancer drugs, but malignancy in medical history has never been previously associated with severe anaphylaxis. 
+<!-- Is it possible that the cancer drugs made them more susceptible to refractory anaphylaxis? -->
+```{r}
+temp <- data3[data3$rANA=="yes",]
+mean_age_malig <- data3$d_age %>% split(data3$q_410_malig_prev_v5 %>% factor) %>% lapply(mean)
+```
+It might be possible that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]]` vs `r mean_age_malig[[2]]` p <
+<!-- `r fisher.test(data3$d_age[which(data3$rANA=="yes")],data3$q_410_malig_prev_v5[which(data3$rANA=="yes")])`  --> 
+0.001) and therefore less able to compensate the anaphylactic shock.
 

### 2018-08-31 analysis/paper/paper.Rmd
discussion ready

@@
 Our findings suggest that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapy with adrenaline, commonly outlined in anaphylaxis management guidelines [@Muraro2014].
-Such cases have the highest risk of a fatal outcome, and therefore need to be treated adequately.
+Such cases have the highest risk of a fatal outcome, and therefore need to be treated accordingly.
 
@@
 hypotension; 2) elicitors - what was the eliciting agent and duration of the exposition; 3) cofactors - is the presence of other factors which may increase the severity of a given reaction i.e.
-concomitant use of betablocker etc.; 4) therapy - how fast and accurate the therapeutic procedures were introduced.
+concomitant use of betablocker etc.; 4) therapy - how fast and accurately the therapeutic procedures were introduced.
 <!-- This theory might explain why younger patients are able to better compensate the same trigger which would elicit a refractory reaction in eldery patients.  -->
@@
 Antibiotics, although commonly triggering pseudoallergic reactions[@saxon1983], are rarely reported in the literature as the cause of a refractory anaphylaxis.
-We saw multiple refractory reactions to antibiotics but in clinical practice, these might be overestimated, as patients in a perioperative setting receive multiple drugs along with antibiotics.
+We saw multiple refractory reactions to antibiotics and since literature describes symptomatically identical pseudoallergic drug reactions to anaphylaxis [@Subramanian2016] it would be only possible to identify these patients after investigating their specific mast cell receptors [@Mcneil2015].
 
@@
 
-There were also relatively less refractory cases elicited by yellow jacket stings (but not bee-stings).
+There were also relatively less refractory cases elicited by yellow-jacket-stings (but not bee-stings).
 It means that the ratio of yellow jacket to honey bee venom elicited anaphylaxis was inverted in the refractory group.
@@
 -->
+
+<!-- Carpa ???-->
+Strikingly, the drugs which were most often reported in patients with refractory anaphylaxis usually elicit pseudoallergic reactions (eg.
+CARPA or MARPA), meaning that a big part of the refractory cases might not be classically associated with a type I hypersensitivity.
+This could explain why the therapy of anaphylaxis in these patients was refractory.
 
@@
 
-We previously saw that concomitant asthma was rather associated with less severe  anaphylaxis [@Worm2018] but the present analysis indicated concomitant asthma as a risk factor for a refractory episode.
-Either patients receive already antihistaminic and mast cell blocking agents and their reaction to adrenaline is therefore restricted, or the episode elicits a bronchospasm that is refractory to adrenaline and therefore results in a prolonged anaphylactic episode.
+We previously saw that concomitant asthma was rather associated with less severe  anaphylaxis [@Worm2018] but the present analyss indicated concomitant asthma as a risk factor for a refractory episode.
+It is possible that the refractory episode elicits a bronchospasm that is refractory to adrenaline and therefore results in a prolonged anaphylactic episode.
 This would also clarify why we saw more respiratory distress symptoms in the refractory anaphylaxis cases.
 
+Psychological stress in temporal proximity to the reaction might increase the severity of the reaction as we previously reported [@Worm2018].
+Until now the plausible explanation for this observation is lacking, and its further investigation is warranted.
+
 <!-- We have previously proven the role of ASA in increasing the risk of severe anaphylaxis [@].  Could not find the reference -->
-The role of PPI however has not been identified as a risk factor, and there are only reports on PPI as elicitors of anaphylaxis.
+The role of PPI as a risk factor for severe anaphylaxis has been indicated in prolonging the exposition to an oral allergen (preventing its degradation due to lower gastric acid production) [@Wlbing2013], and there are reports on PPI as elicitors of anaphylaxis [@Natsch2000].
+We saw a significantly higher rate of PPI intake in the refractory anaphylaxis group but its mechanistic contribution to the increase in  severity of an episode or being responsible for its refractoriness cannot be estimated based on this observation.
 
@@
 ## Therapy 
-<!-- After revising the therapeutic procedures we saw significantly more adrenaline use in the refractory cases, but it could have been used even more extensively in these life-threatening reactions.
+<!-- After revising the therapeutic procedures
+
+We saw significantly more adrenaline use in the refractory cases, but it could have been used even more extensively in these life-threatening reactions.
 This observation might indicate that in fact the emergency teams are using these life-saving measures too cautiously.
@@
 as first-line therapy use was higher in the refractory cases, probably due to most of them happening in a medical setting, however, other second-line therapeutic options were rarely used.
-
-Methylene blue and vasopressors have been described to successfully restore refractory hypotension, but their actual use in anaphylaxis patients is scarce, although they are mentioned in recognized guidelines on the management of anaphylaxis [@Muraro2014].
+Methylene blue and vasopressors have been described to successfully restore refractory hypotension and recommended by managment guidelines [@Muraro2014], but their actual use in anaphylaxis patients is scarce.
+Grabenhenrich et al.
+[@Grabenhenrich2018] recently evaluated the epinephrine use in anaphylaxis patients and concluded that, even in this state of the art drug, large discrepancies between recommenden use and actual treatment practice exist.
+Therefore, more effort needs to be dedicated to promoting and developing the consensus guidelenes as practically as possiblle in order to increasing adherence.
 
@@
 Definition  of refractory anaphylaxis is not universal, and confirming the refractory anaphylaxis based on the answers to our online questionnaire poses additional difficulties.
-If fatal reaction occurs before the second dose of adrenaline could be administered to the patient - it cannot be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline) although it as well could be refractory.
+If a fatal reaction occurs before the second dose of adrenaline could be administered to the patient - it cannot be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline) although it as well could be refractory.
 
@@
 Refractory anaphylaxis is a rare form of life-threatening hypersensitivity reaction with high mortality.
-We propose to increase the use of second line medication such as methylene blue, vasopressin and glucagon
-in cases where 2 doses of adrenaline did not result in rapid normalization of anaphylaxis symptoms.
+Its elicitors and cofactors differ from other anaphylaxis cases, and the managment of refractory anaphylaxis needs to improve.
+We propose to increase the use of second line medication such as methylene blue, vasopressin and (in suspicion of a beta adrenergic blockade) glucagon in cases where 2 doses of adrenaline did not result in a rapid normalization of anaphylaxis symptoms.
 

### 2018-09-14 analysis/paper/paper.Rmd
Anaphylaxis corrections from Prof worm introduced

@@
 Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 mg adrenaline) is a rare and often fatal hypersensitivity reaction.
-Based on this, its definition, prevalence, and risk factors are only vaguely described.
-Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we were able to identify cases of refractory anaphylaxis (n = `r v$casesR`) and to analyse these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
+Comprehensive data on its definition, prevalence, and risk factors are missing.
+Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified refractory anaphylaxis cases (n = `r v$casesR`) and analysed these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
 
@@
 `r v$perioperative[2]`, p < 0.0001).
-Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases was given intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001).
+Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases it was applied intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001).
 Second line treatment options (e.g.
@@
 Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these guidelines can be improved.
-The low use of intramuscular adrenaline as a first line therapy might contribute to refractoriness of an episode.
-Also, the lack of suggested second line medication use in refractory cases indicate that the availability of these drugs is either restricted or anaphylaxis management guidelines may need to be revised to provide clear algorithms for severe refractory anaphylaxis.
+The lack of second line medication (i.e.
+methylene blue or dopamine) use in refractory cases has to be addressed in future efforts to improve the managment (and guidelines adherence) of severe refractory anaphylaxis cases.
 
@@
 
-Anaphylaxis is a non-homogeneous clinical diagnosis, depending on various triggering and augmenting factors.
+Anaphylaxis is a non-homogeneous clinical diagnosis, depending on various triggering and augmenting factors [@Simons2015].
 This variability introduces a wide range of possible reaction-symptom severities.
@@
 
-The mainstay of anaphylaxis management is the intramuscular dose of adrenaline, but in the most severe cases of anaphylaxis it might be insufficient to restore a stable patient status.
-Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least 2 doses of minimum 300 mg adrenaline, does not respond with an expected normalization of clinical symptoms.
+The mainstay of anaphylaxis management is the intramuscular dose of adrenaline [@Grabenhenrich2018], but in the most severe cases of anaphylaxis it might be insufficient to restore a stable patient status.
+Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least 2 doses of minimum 300 µg adrenaline, does not achieve normalization of clinical symptoms.
 Typical elicitors and symptoms of refractory anaphylaxis, as well as the therapeutic strategy for the most severe cases, differ from the usual reactions [@Francuzik2018] and call for a specific research and targeted guideline development for refractory anaphylaxis cases.
@@
 Milder anaphylaxis symptoms (i.e.
-pruritus, gastrointestinal symptoms, vertigo, chest and throat tightness) were significantly less often present in refractory anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were more often associated with the refractory anaphylaxis cases.
+pruritus, gastrointestinal symptoms, vertigo, chest and throat tightness) were significantly less present in refractory anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were more often associated with the refractory anaphylaxis cases.
 Table \@ref(tab:symptTab) summarizes the most prominent differences in anaphylaxis symptoms among both groups.
 <!-- ## Death  -->
-Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the ardiopulmonary resuscitation was unsuccessful.
+Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the cardiopulmonary resuscitation was unsuccessful.
 In cases where patients responded to life support procedures, but failed to be reanimated due to post-resuscitative complication (e.g.
@@
 
-When evaluating the therapeutic procedures, adrenaline iv.
+When evaluating the therapeutic procedures, adrenaline i.v.
 as a first-line treatment of anaphylaxis was significantly more often given in refractory cases (`r therapyTab[2,3]`% vs.
-`r therapyTab[1,3]`%, p < 0.0001) it has been also given more often as a second line therapy in refractory cases.
-Median time to the second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes vs.
-15 minutes in non-refractory cases, p = `r med.timep$p.value %>% signif(3)`).
+`r therapyTab[1,3]`%, p < 0.0001).
+Median time to the second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes vs. 15 minutes in non-refractory cases, p <0.0001). 
+<!-- `r med.timep$p.value %>% signif(3)`). -->
 <!--  wilcox.test(data3$q_530_adren2_time_in_min_v5[data3$rANA=="yes"],data3$q_530_adren2_time_in_min_v5[data3$rANA=="no"])$p.value %>% signif(3)-->
@@
 
-Second line medication like dopamine, glucagon and methylene blue were neglected in all refractory anaphylaxis cases as well as in severe non-refractory ones, however patients with refractory anaphylaxis were more frequently admitted to the hospital and treated at the ICUs.
+Second line medication like dopamine, glucagon and methylene blue were not given in all refractory anaphylaxis cases as well as in severe non-refractory ones.
+However, patients with refractory anaphylaxis were more frequently admitted to the hospital (`r therapyTab[18,3]`%) and treated at the ICUs (`r therapyTab[19,3]`%).
 Table \@ref(tab:therapyTab) illustrates the therapy of refractory anaphylaxis cases.
@@
 <!-- ## Cofactors -->
-Patients with refractory anaphylaxis more often had concomitant asthma and malignant diseases in the past.
+Patients with refractory anaphylaxis more often had concomitant asthma and malignant diseases in medical history.
 Also other unspecified concomitant conditions were significantly more often reported in refractory cases.
@@
 
-Patients with refractory anaphylaxis more often used proton pump inhibitors (PPI) and acetylsalicilic acid (ASA) than patients with severe non-refractory anaphylaxis.
+Patients with refractory anaphylaxis more often reported concomitant proton pump inhibitors (PPI) and acetylsalicilic acid (ASA) use compared to patients with severe non-refractory anaphylaxis.
 Other (not specified) medications were also more frequent in refractory cases.
@@
 Such cases have the highest risk of a fatal outcome, and therefore need to be treated accordingly.
+An early use of drugs (e.g.
+methylene blue and dopamine) might increse the survival chance.
 
-Overall, the severity of anaphylaxis and its probable transition into a refractory episode might be described as in the figure \@ref(fig:theory).
-Patients who undergo anaphylaxis have four main aspects that influence the severity of a particular episode: 1) compensation - how well is the body able to compensate reaction symptoms e.g.
-hypotension; 2) elicitors - what was the eliciting agent and duration of the exposition; 3) cofactors - is the presence of other factors which may increase the severity of a given reaction i.e.
-concomitant use of betablocker etc.; 4) therapy - how fast and accurately the therapeutic procedures were introduced.
+Overall, the severity of an anaphylaxis and its probable transition into a refractory episode might be described by four co-influencing mechanisms: 
+1) elicitors - the type of an eliciting agent and duration of the exposition, which correlates to the amount of the allergen and its metabolisation; 
+2) cofactors - the presence of other factors which may increase the severity of a given reaction i.e. concomitant use of a betablocker etc.;
+3) compensation - how well is the body able to compensate reaction symptoms e.g. hypotension; 
+4) therapy - how fast and accurately the therapeutic procedures were introduced (Fig. \@ref(fig:theory)).
 <!-- This theory might explain why younger patients are able to better compensate the same trigger which would elicit a refractory reaction in eldery patients.  -->
 
-## Symptom compensation
-Age is the most important factor influencing the risk of developing severe anaphylaxis [@Worm2018].
-It has been shown that older age may correlate with the decreased abillity to retain homeostasis on increased strain [@fleg1995].
-Patients who underwent refractory anahpylaxis more often had perioperative reactions  and therefore a decreased ablity to compensate the reaction symptoms with reflexory renal or cardiopulmonary response [@Rogers2013].
+## Elicitors
+Drugs were the most commonly occurring elicitor, which is in concordance with the literature and our previous findings [@Francuzik2018].
+Multiple medications with mast cell activation potential [@Hepner2003] given in a perioperative setting increase the chance of a hypersensitivity reaction and drug interactions [@Ebo2007].
+Recent study on IgE-independent anaphylaxis showed that peroperative drugs (i.e.
+rocuronium, turbocurarine, fluorochinolones, atracurium) may trigger anaphylaxis by activating mast cells directly through the MrgprX2 receptor [@Mcneil2015].
+Therefore patients who undergo surgical procedures and have a history of anaphylaxis should remain under extraordinary caution.
 
-## Elicitors
-Drugs were the most commonly occurring elicitor, what is in concordance with our previous findings [@Francuzik2018].
-Multiple medications given in a perioperative setting increase the chance of a hypersensitivity reaction and drug interactions [@Ebo2007].
-Therefore patients who undergo medical procedures and have a history of anaphylaxis should remain under extraordinary caution.
+Antibiotics, although commonly triggering IgE-dependent reactions [@saxon1983], are rarely reported in the literature as the cause of a refractory anaphylaxis.
+RCM on the other hand, commonly elicit IgE-independent hypersensitivity episodes that are refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [@Greenberger1991].
+We previously saw RCM to be the most frequent elicitor of refractory anaphylaxis [@Francuzik2018].
+It might be that RCM promote  complement activation [@Wang2011] and thus - unspecific activation of multiple immune cell classes (mast cells, basophiles, platelets and neutophiles).
+Therefore, treatment with adrenaline may be less effective in these reactions.
 
-Antibiotics, although commonly triggering pseudoallergic reactions[@saxon1983], are rarely reported in the literature as the cause of a refractory anaphylaxis.
-We saw multiple refractory reactions to antibiotics and since literature describes symptomatically identical pseudoallergic drug reactions to anaphylaxis [@Subramanian2016] it would be only possible to identify these patients after investigating their specific mast cell receptors [@Mcneil2015].
-
-RCM however, commonly elicit hypersensitivity episodes refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [@Greenberger1991].
-We previously saw RCM to be the most frequent elicitor of refractory anaphylaxis [@Francuzik2018].
-One of the plausible explanation to this is that RCM promote complement activation related pseudo allergy (CARPA) [@Wang2011], and therfore the treatment with adrenaline may be less effective in these reactions.
-
-There were also relatively less refractory cases elicited by yellow-jacket-stings (but not bee-stings).
-It means that the ratio of yellow jacket to honey bee venom elicited anaphylaxis was inverted in the refractory group.
-Plausible explanations for this observation might include either longer exposition to the allergen (bee's sting remains in the skin) or the venom volume being 10 times higher than in a wasp sting.
-Patients should remove the bee sting as quickly as possible (optimal < 2 seconds) [@Visscher1996].
+Only few refractory cases were elicited by yellow-jacket-stings.
+Therefore, the ratio of yellow-jacket to honey-bee venom elicited anaphylaxis was inverted in the refractory group.
+Reason for this observation might include higher allergen exposition due to either longer exposition to the allergen (bee's sting remains in the skin) or the venom volume being 10 times higher than in a wasp sting [@Visscher1996].
+<!-- Patients should remove the bee sting as quickly as possible (optimal < 2 seconds) . -->
 
@@
 <!-- Carpa ???-->
-Strikingly, the drugs which were most often reported in patients with refractory anaphylaxis usually elicit pseudoallergic reactions (eg.
-CARPA or MARPA), meaning that a big part of the refractory cases might not be classically associated with a type I hypersensitivity.
-This could explain why the therapy of anaphylaxis in these patients was refractory.
 
 ## Risk factors 
-Although we have previously identified older age and male sex to be associated with more severe anaphylaxis [@Worm2018], these factors seem to be less important when refractory anaphylaxis is considered.
-Asthma and malignant diseases in the medical history were more frequent in the refractory anaphylaxis group and the reason for this is unclear.
+Although we have previously identified older age and male sex to be associated with severe anaphylaxis [@Worm2018], these factors seem to be less important when refractory anaphylaxis is considered.
+Asthma and malignant diseases in the medical history were more frequently present in the refractory anaphylaxis group.
+The underlying reason is unclear.
 It is however interesting that past malignant diseases were mostly recognized in male refractory anaphylaxis patients.
 
-Literature describes many cases of hypersenitivity due to cancer drugs, but malignancy in medical history has never been previously associated with severe anaphylaxis. 
 <!-- Is it possible that the cancer drugs made them more susceptible to refractory anaphylaxis? -->
@@
 ```
-It might be possible that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]]` vs `r mean_age_malig[[2]]` p <
+Many cases of hypersenitivity due to cancer drugs were described in the literature, but malignancy in medical history has never been previously associated with severe anaphylaxis. 
+It can be possible that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]] %>% round(1)` vs `r mean_age_malig[[2]] %>% round(1)` years, p <
 <!-- `r fisher.test(data3$d_age[which(data3$rANA=="yes")],data3$q_410_malig_prev_v5[which(data3$rANA=="yes")])`  --> 
@@
 
-We previously saw that concomitant asthma was rather associated with less severe  anaphylaxis [@Worm2018] but the present analyss indicated concomitant asthma as a risk factor for a refractory episode.
-It is possible that the refractory episode elicits a bronchospasm that is refractory to adrenaline and therefore results in a prolonged anaphylactic episode.
-This would also clarify why we saw more respiratory distress symptoms in the refractory anaphylaxis cases.
+We previously observed concomitant asthma to be associated with less severe  anaphylaxis [@Worm2018].
+However, the present analysis suggests concomitant asthma as a risk factor for a refractory episode.
+It is possible that during a refractory episode the bronchospasm is mediated via other mediators (deriving from basophiles, and eosinophiles) what makes it refractory to adrenaline and therefore results in a prolonged anaphylactic episode.
+This might also explain why we saw more respiratory distress symptoms in the refractory anaphylaxis cases.
 
 Psychological stress in temporal proximity to the reaction might increase the severity of the reaction as we previously reported [@Worm2018].
-Until now the plausible explanation for this observation is lacking, and its further investigation is warranted.
+It has been reported that stress induces the complement cascade activation [@Burns2008].
+Substance P which is released during stress [@Arck2003] is a known vasodilator [@Bos92] and can activate mast cells directly [@Singh1999].
+These mediators may contribute to an increased mast cell response and more severe anaphylaxis.
 
 <!-- We have previously proven the role of ASA in increasing the risk of severe anaphylaxis [@].  Could not find the reference -->
-The role of PPI as a risk factor for severe anaphylaxis has been indicated in prolonging the exposition to an oral allergen (preventing its degradation due to lower gastric acid production) [@Wlbing2013], and there are reports on PPI as elicitors of anaphylaxis [@Natsch2000].
-We saw a significantly higher rate of PPI intake in the refractory anaphylaxis group but its mechanistic contribution to the increase in  severity of an episode or being responsible for its refractoriness cannot be estimated based on this observation.
+The role of PPI as a risk factor for severe anaphylaxis has been indicated in prolonging the exposition to an oral allergen (preventing its degradation due to lower gastric acid production) [@Wlbing2013].
+Moreover, there are reports on PPI as elicitors of anaphylaxis [@Natsch2000].
+We observed a significantly higher rate of PPI intake in the refractory anaphylaxis group.
+Its mechanistic contribution to the increase in  severity of an episode or being responsible for its refractoriness cannot be estimated based on this observation.
 
 <!--Low-Dose Ethanol Alters the Cardiovascular, Metabolic, and Respiratory Compensation for Severe Blood Loss-->
+
+## Symptom compensation
+Age is the most important factor influencing the risk of developing severe anaphylaxis [@Worm2018].
+We and others have shown that older age may correlate with the decreased abillity to retain homeostasis on increased strain [@fleg1995].
+Patients who underwent refractory anahpylaxis more often had perioperative reactions  and therefore a decreased ablity to compensate the reaction symptoms with reflexory renal or cardiopulmonary response [@Rogers2013].
 
@@
 Adrenaline i.v.
-as first-line therapy use was higher in the refractory cases, probably due to most of them happening in a medical setting, however, other second-line therapeutic options were rarely used.
-Methylene blue and vasopressors have been described to successfully restore refractory hypotension and recommended by managment guidelines [@Muraro2014], but their actual use in anaphylaxis patients is scarce.
+as first-line therapy use was high in the refractory cases.
+This was probably due to the fact that most of them occur in a medical setting.
+However, other second-line therapeutic options were rarely used.
 Grabenhenrich et al.
 [@Grabenhenrich2018] recently evaluated the epinephrine use in anaphylaxis patients and concluded that, even in this state of the art drug, large discrepancies between recommenden use and actual treatment practice exist.
-Therefore, more effort needs to be dedicated to promoting and developing the consensus guidelenes as practically as possiblle in order to increasing adherence.
+Therefore, more effort needs to be dedicated to promote and develop the consensus guidelenes as practically as possible in order to increase adherence.
 
-## Limitations and strength
-Low number of refractory cases prevented us from analyzing the data with more advanced statistical models, however, our analysis is the first published report on over 30 refractory anaphylaxis cases.
+Methylene blue and vasopressors have been described to successfully restore refractory hypotension and recommended by managment guidelines [@Muraro2014], but their actual use in anaphylaxis patients is scarce.
+There are multiple reports of successful methylene blue use in refractory anaphylaxis [@Francuzik2018].
+Evora [@Evora2000] reported 6 cases of refractory, peroperative anaphylaxis (to RCM and penicillin) which responded to methylene blue i.v.
+within minutes.
+Its mode of action is by blocking the guanylate cyclase and therefore preventing further nitric-oxide-dependent vasodilation in a distributive shock [@dfaz].
 
-Definition  of refractory anaphylaxis is not universal, and confirming the refractory anaphylaxis based on the answers to our online questionnaire poses additional difficulties.
-If a fatal reaction occurs before the second dose of adrenaline could be administered to the patient - it cannot be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline) although it as well could be refractory.
+## Limitations and strengths
+The low number of refractory cases prevented us from analyzing the data with more advanced statistical models.
+However, our analysis is the first published report on a patient sample containing over 30 refractory anaphylaxis cases.
 
-Comparison of refractory anaphylaxis to patients with severe anaphylaxis gave us the possibility to describe which patients have a higher risk of developing a refractory episode, however we can not answer the question which patients have the higher risk of developing anaphylaxis per se.
+Definition  of refractory anaphylaxis is not universal. Also in this study, the confirmation of a refractory anaphylaxis based on the answers to our online questionnaire posed difficulties.
+If a fatal reaction occured before the second dose of adrenaline was administered to the patient - it could not be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline) although it in fact could have been refractory.
+
+Comparison of refractory anaphylaxis with patients suffering from severe anaphylaxis enabled us to distinguish patients with a higher risk of developing a refractory episode.
+However we cannot address the question which patients had a higher risk of experiencing anaphylaxis per se.
 
@@
 
-Refractory anaphylaxis is a rare form of life-threatening hypersensitivity reaction with high mortality.
+Refractory anaphylaxis is a rare form of a life-threatening hypersensitivity reaction with high mortality.
 Its elicitors and cofactors differ from other anaphylaxis cases, and the managment of refractory anaphylaxis needs to improve.
-We propose to increase the use of second line medication such as methylene blue, vasopressin and (in suspicion of a beta adrenergic blockade) glucagon in cases where 2 doses of adrenaline did not result in a rapid normalization of anaphylaxis symptoms.
+Although, more studies need to be conducted to better understand the pathomechanisms involved in refractory anaphylaxis, we propose to increase the use of second line medication such as methylene blue, vasopressin and (in suspicion of a beta adrenergic blockade) glucagon in cases where 2 doses of adrenaline did not result in a rapid normalization of anaphylaxis symptoms.
 
@@
 
+#####
+
 ```{r theory, fig.cap="Visual representation of the three forces determining the severity of anaphylaxis.
-The natural ability of the body to compensate the anaphylaxis symptoms and therapeutic measures act to restore homeostasis to a patient by whom elicitors and cofactors influence synergistically the severity of a given episode."}
+The natural ability of the body to compensate the anaphylaxis symptoms and therapeutic measures acting to restore homeostasis to a patient by whom elicitors and cofactors influence synergistically the severity of a given episode."}
 
@@
 
+#####
+
 # Tables
@@
 ```{r demoTab}
-knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. Age is represented by a mean value, other as fractions [%]. DM - diabetes mellitus",)
+knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. Age is represented by a mean value, other as fractions [%]. DM - diabetes mellitus")
 ```
+
+#####
 
@@
 ```
+
+#####
 
@@
 
+#####
+
 ```{r symptTab}
@@
 
+#####
+
 ```{r cofactorsTab}

### 2018-09-14 R/analysis.R

@@ -509,6 +509,8 @@ v$mortality <- rcalc("q_140_fatal")# ** Tables ----
 sdb <- data3[which(data3$rANA=="no"),]
 
 #### 1. Demography ################
+
+
 demoTab <- cbind(n = rdb$b_sex %>% summary(),
                 Age = rdb$d_age %>% split(rdb$b_sex) %>%
                               lapply(.,function(x){mean(x) %>% signif(3)}),
@@ -520,7 +522,8 @@ demoTab <- cbind(n = rdb$b_sex %>% summary(),
                               f3,
                  Mastocytosis = rdb$q_410_masto_cur %>% split(rdb$b_sex) %>% f3,
                  Malignancy = rdb$q_410_malig_cur %>% split(rdb$b_sex)  %>% f3,
-                `Atopic dermatitis` = rdb$q_410_ad_cur %>% split(rdb$b_sex)  %>% f3
+                `Atopic dermatitis` = rdb$q_410_ad_cur %>% split(rdb$b_sex)  %>% f3,
+                `tryptase [median]` = rdb$q_212_tryptase_value_v5 %>% split(rdb$b_sex) %>% lapply(median,na.rm=T)
                 )
 
 
@@ -532,7 +535,8 @@ demoTabs <- cbind(n = sdb$b_sex %>% summary(),
                  `Food allergy` = sdb$q_410_foodallergy_cur_v6 %>%  split(sdb$b_sex) %>%f3,
                  Mastocytosis = sdb$q_410_masto_cur %>% split(sdb$b_sex) %>% f3,
                  Malignancy = sdb$q_410_malig_cur %>% split(sdb$b_sex)  %>% f3,
-                 `Atopic dermatitis` = sdb$q_410_ad_cur %>% split(sdb$b_sex)  %>% f3
+                 `Atopic dermatitis` = sdb$q_410_ad_cur %>% split(sdb$b_sex)  %>% f3,
+                 `tryptase [median]` = sdb$q_212_tryptase_value_v5 %>% split(sdb$b_sex) %>% lapply(median,na.rm=T)
 )
 
 demoTabsP <- cbind(n = data3$b_sex %>% table(data3$rANA) %>% summary() %>% {.$p.value} %>% signif(3),
@@ -542,7 +546,9 @@ demoTabsP <- cbind(n = data3$b_sex %>% table(data3$rANA) %>% summary() %>% {.$p.
[…]

### 2018-09-14 analysis/paper/paper.Rmd
figure and p value formatting changed

@@
 
-knitr::include_graphics("../figures/diag.png")
+knitr::include_graphics("../figures/diag2.png")
 ```
@@
 ```{r demoTab}
-knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases. Age is represented by a mean value, other as fractions [%]. DM - diabetes mellitus")
+knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases.
+Age is represented by a mean value, other variables as fractions [%].
+DM - diabetes mellitus, p value is derived from a Mann-Whitney U test or a Chi^2^ test for interval and categorical variables respectively.",
+             digits = 2,
+             align = rep("r",10),
+             nsmall=2)
 ```
@@
 ANA - anaphylaxis, p-value derived from the Fisher exact test.",
-             col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]"),
-             digits = 1)
+             col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]")
+            #digits = 1,
+             # align = rep("r",10),
+             # trim=T,
+             # scientific=F,
+             # nsmall=3,
+             #drop0trailing = F
+             )
 ```

### 2018-09-14 R/analysis.R

@@ -41,20 +41,26 @@ rcalc <- function(var,level="yes"){
   np <- data3$rANA %>% factor %>% summary %>% {.[2]}-pp
   eval <- c(pn,pp,nn,np) %>% matrix(byrow = T,ncol = 2)
   prt <- (prop.table(eval,2)*100) %>% {signif(.,3)}
-  p <- eval %>% fisher.test() %>% .$p.value %>% signif(3)
+  p <- eval %>% fisher.test() %>% .$p.value %>% roP
   return(c(var,prt[1,1],prt[1,2],p,l))
+}
+
+roP<- function(x){
+  ifelse(x > 0.001,
+  round(x,3),ifelse(x==0,0,0.0001)
+  )
 }
 
 f1 <- function(x){
   names(x) <- NULL
-  return(do.call(t.test,x)$p.value %>% signif(3))
+  return(do.call(t.test,x)$p.value %>% roP)
 }
 f2 <- function(x){
-  x %>% table(data3$rANA) %>% {.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% signif(3)
+  x %>% table(data3$rANA) %>% {.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% roP
 }
 f3 <- function(x){
   x %>%
-    lapply(function(x){(length(which(x=="yes"))/length(x)*100) %>% round(0)})
+    lapply(function(x){(length(which(x=="yes"))/length(x)*100) %>% roP})
 }
 
 
@@ -548,9 +554,10 @@ demoTabsP <- cbind(n = data3$b_sex %>% table(data3$rANA) %>% summary() %>% {.$p.
                   Malignancy = data3$q_410_malig_cur %>% f2,
                   `Atopic dermatitis` = data3$q_410_ad_cur %>% f2,
                   `tryptase [mean]` = data3$q_212_tryptase_value_v5[data3$rANA=="no"] %>%
-                    wilcox.test(data3$q_212_tryptase_value_v5[data3$rANA=="yes"]) %>% {.$p.value}
+                    wilcox.test(data3$q_212_tryptase_value_v5[data3$rANA=="yes"]) %>% {.$p.value} %>% round(3)
 
 )
+
 demoTab <- rbind(demoTab,demoTabs,demoTabsP)
 demoTab <- cbind(Group = c("refractory","refractory","severe","severe","p value"),demoTab)
 
[…]

### 2018-09-18 analysis/paper/paper.Rmd
glucagon discussion add

@@
 Other (not specified) medications were also more frequent in refractory cases.
+In `r table(data3$rANA,data3$q_423_beta)[2,2]` cases of refractory anaphylaxis (`r (table(data3$rANA,data3$q_423_beta)[2,2]/sum(table(data3$rANA,data3$q_423_beta)[2,1:3])*100) %>% round(1)`%) patients reported receiving beta-blockers as a concomitant medication, but none of these patients recieved a glucagon infusion.
 
@@
 
+Suprisingly, even though multiple anaphylaxis management guidelines recommend glucagon infusions in cases of concomitant beta-blocker therapy in anaphylactic patients [@Simons2011;@Lieberman;@Muraro2014], it has not been administered in any of the severe or refractory cases.
+Glucagon has been reported to sucessfully relive refractory anaphylaxis [@Zaloga1986;@Javeed1996] by directly activating the adenylyl cyclase and therefore bypassing the \beta-adrenergic receptor blockade [@Zaloga1986].
+
 ## Limitations and strengths
@@
 Definition  of refractory anaphylaxis is not universal. Also in this study, the confirmation of a refractory anaphylaxis based on the answers to our online questionnaire posed difficulties.
-If a fatal reaction occured before the second dose of adrenaline was administered to the patient - it could not be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline) although it in fact could have been refractory.
+If a fatal reaction occured before the second dose of adrenaline was administered to the patient - it could not be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although it in fact could have been refractory.
 

### 2018-09-18 analysis/paper/paper.Rmd
corrections after final revision

@@
 <!-- 350 Words unstructured -->
-Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 mg adrenaline) is a rare and often fatal hypersensitivity reaction.
+Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 µg adrenaline) is a rare and often fatal hypersensitivity reaction.
 Comprehensive data on its definition, prevalence, and risk factors are missing.
@@
 The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).
-Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r elicitExact[4,3]`% vs `r elicitExact[4,2]`%, p = `r elicitExact[4,4]`).
+Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r 5/8*100`% vs `r round(449/(77+1607+4+9+170+3+449)*100,1)`%, p = `r fisher.test(matrix(c(449,5,77+1607+4+9+170+3,3),nrow=2))$p.value %>% round(3)`).
 The refractory cases occurred more frequently in a perioperative setting (`r v$perioperative[3]`% vs.
@@
 Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these guidelines can be improved.
-The lack of second line medication (i.e.
-methylene blue or dopamine) use in refractory cases has to be addressed in future efforts to improve the managment (and guidelines adherence) of severe refractory anaphylaxis cases.
+The neglegence of second line medication (i.e.
+methylene blue or dopamine) in refractory cases has to be addressed in future efforts to improve the managment (and guidelines adherence) of severe refractory anaphylaxis cases.
 
@@
 This variability introduces a wide range of possible reaction-symptom severities.
-Therefore multiple sub-types of anaphylaxis have been previously identified (i.e.
+Therefore, multiple sub-types of anaphylaxis have been previously identified (i.e.
 food dependent exercise induced anaphylaxis, venom anaphylaxis, biphasic anaphylaxis).
@@
 The statistical analysis was performed in the R Statistical Package [@R].
-Simple comparison of nominal variables was performed using Fischer's exact test, continous variables were analysed using Mann-Whithney U test.
+Simple comparison of categorical variables was performed using Fisher's exact test, continous variables were analysed using Mann-Whithney U test.
 We defined statistical significance as α = 0.05.
-Data along with the analysis script can be accessed at github.com/wolass/RefractoryAnaOrg.
+Data along with the analysis script can be accessed at www.github.com/wolass/RefractoryAnaOrg.
 
@@
 Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the cardiopulmonary resuscitation was unsuccessful.
-In cases where patients responded to life support procedures, but failed to be reanimated due to post-resuscitative complication (e.g.
+In cases where patients responded to life support, but failed to be reanimated due to post-resuscitative complication (e.g.
 hypoxic brain injury), death occured in the next 3-8 days.
@@
 `r therapyTab[1,3]`%, p < 0.0001).
-Median time to the second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes vs. 15 minutes in non-refractory cases, p <0.0001). 
+Median time to the second dose of adrenaline was also shorter in refractory cases (`r med.time2[[2]]` minutes vs.
+15 minutes in non-refractory cases, p <0.0001).
 <!-- `r med.timep$p.value %>% signif(3)`). -->
@@
 were the second most frequently administered group of drugs in refractory cases (as a first and second line treatment), outpacing antihistaminic drugs and volume replacement therapy, and were significantly more often given in refractory cases.
-Volume replacement therapy was given initially in `r therapyTab[4,3]`% of refractory cases but was sustained only in `r therapyTab[5,3]`% as the therapy progressed in the hospital environment.
+Volume replacement therapy was given initially in `r therapyTab[4,3]`% of refractory cases and was sustained only in `r therapyTab[5,3]`% as the therapy progressed in the hospital environment.
 <!--(possible counfounding variable by patients who died!!! they did not receive second line therapy and might need to be discarded from the analysis)-->
@@
 An early use of drugs (e.g.
-methylene blue and dopamine) might increse the survival chance.
+methylene blue and dopamine) might increse their survival chance.
 
 Overall, the severity of an anaphylaxis and its probable transition into a refractory episode might be described by four co-influencing mechanisms: 
-1) elicitors - the type of an eliciting agent and duration of the exposition, which correlates to the amount of the allergen and its metabolisation; 
+1) elicitors - the type of an eliciting agent and duration of the exposition, which correlates to the amount of the allergen and its metabolism; 
 2) cofactors - the presence of other factors which may increase the severity of a given reaction i.e. concomitant use of a betablocker etc.;
@@
 Recent study on IgE-independent anaphylaxis showed that peroperative drugs (i.e.
-rocuronium, turbocurarine, fluorochinolones, atracurium) may trigger anaphylaxis by activating mast cells directly through the MrgprX2 receptor [@Mcneil2015].
+rocuronium, turbocurarine, fluorochinolones, atracurium) may trigger anaphylaxis by activating mast cells directly through the MRGPRX2 receptor [@Mcneil2015].
 Therefore patients who undergo surgical procedures and have a history of anaphylaxis should remain under extraordinary caution.
@@
 Therefore, the ratio of yellow-jacket to honey-bee venom elicited anaphylaxis was inverted in the refractory group.
-Reason for this observation might include higher allergen exposition due to either longer exposition to the allergen (bee's sting remains in the skin) or the venom volume being 10 times higher than in a wasp sting [@Visscher1996].
+Reasons for this observation might include higher allergen exposition due to either longer exposition to the allergen (bee's sting remains in the skin) or the venom volume being 10 times higher than in a wasp sting [@Visscher1996].
 <!-- Patients should remove the bee sting as quickly as possible (optimal < 2 seconds) . -->
@@
 The underlying reason is unclear.
-It is however interesting that past malignant diseases were mostly recognized in male refractory anaphylaxis patients.
 
@@
 ```
+It is however interesting that past malignant diseases were mostly recognized in male refractory anaphylaxis patients.
 Many cases of hypersenitivity due to cancer drugs were described in the literature, but malignancy in medical history has never been previously associated with severe anaphylaxis.
-It can be possible that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]] %>% round(1)` vs `r mean_age_malig[[2]] %>% round(1)` years, p <
+It is possible that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]] %>% round(1)` vs `r mean_age_malig[[2]] %>% round(1)` years, p <
 <!-- `r fisher.test(data3$d_age[which(data3$rANA=="yes")],data3$q_410_malig_prev_v5[which(data3$rANA=="yes")])`  --> 
@@
 Psychological stress in temporal proximity to the reaction might increase the severity of the reaction as we previously reported [@Worm2018].
-It has been reported that stress induces the complement cascade activation [@Burns2008].
-Substance P which is released during stress [@Arck2003] is a known vasodilator [@Bos92] and can activate mast cells directly [@Singh1999].
+Stress induces the complement cascade activation [@Burns2008].
+Substance P, which is released during stress [@Arck2003], is a known vasodilator [@Bos92] and can activate mast cells directly [@Singh1999].
 These mediators may contribute to an increased mast cell response and more severe anaphylaxis.
@@
 Suprisingly, even though multiple anaphylaxis management guidelines recommend glucagon infusions in cases of concomitant beta-blocker therapy in anaphylactic patients [@Simons2011;@Lieberman;@Muraro2014], it has not been administered in any of the severe or refractory cases.
-Glucagon has been reported to sucessfully relive refractory anaphylaxis [@Zaloga1986;@Javeed1996] by directly activating the adenylyl cyclase and therefore bypassing the \beta-adrenergic receptor blockade [@Zaloga1986].
+Glucagon has been reported to sucessfully relive refractory anaphylaxis [@Zaloga1986;@Javeed1996] by directly activating the adenylyl cyclase and therefore bypassing the β-adrenergic receptor blockade [@Zaloga1986].
 
@@
 
-Definition  of refractory anaphylaxis is not universal. Also in this study, the confirmation of a refractory anaphylaxis based on the answers to our online questionnaire posed difficulties.
-If a fatal reaction occured before the second dose of adrenaline was administered to the patient - it could not be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although it in fact could have been refractory.
+Definition  of refractory anaphylaxis is not universal. The confirmation of a refractory anaphylaxis based on the answers to our online questionnaire posed difficulties.
+If a fatal reaction occured before the second dose of adrenaline was administered to the patient - it could not be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although, it in fact could have been refractory.
 
@@
 Age is represented by a mean value, other variables as fractions [%].
-DM - diabetes mellitus, p value is derived from a Mann-Whitney U test or a Chi^2^ test for interval and categorical variables respectively.",
+DM - diabetes mellitus, p value is derived from a Mann-Whitney U test or a Chi^2^ test for interval and categorical variables respectively.
+Refractory group was compared to severe anaphylaxis cases without differentiating into male and female subgroups.",
              digits = 2,
@@
 ```{r elicitorTab}
-knitr::kable(elicitorTab, caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
-ANA - anaphylaxis, p-value derived from the Fisher exact test.",
-             col.names = c("n","refractory ANA [%]","severe ANA [%]","p value","Age","Male sex [%]","Perioperative [%]","Food allergy [%]")
+knitr::kable(elicitorTab[,c(1,5,6,7,8,2,3,4)], caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
+ANA - anaphylaxis, Age is represented as a mean, p-value derived from the Fisher's exact test.",
+             col.names = c("n","Age","Male sex [%]","Perioperative [%]","Food allergy [%]","refractory ANA [%]","severe ANA [%]","p value")
             #digits = 1,
@@
 ```{r elicitExact}
-knitr::kable(cbind(elicitExact[,5],elicitExact[,2:4]), caption = "Summary of the elicitors broken down to specific elicitors from each group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control.
-ANA - anaphylaxis, p-value derived from the Fisher exact test.
+
+knitr::kable(cbind(c("Antibiotics","X-ray medium","Muscle relaxant","Legumes","Bee venom","Yellow-jacket venom"),elicitExact[,2:4]), caption = "Summary of the elicitors broken down into specific elicitors from each group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control.
+ANA - anaphylaxis, p-value derived from the Fisher's exact test.
 ",
@@
 ```
-
-p-value derived from the Fisher exact test.
 
@@
 knitr::kable(symptTab[,1:4], caption = "Summary of the symptoms in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
-ANA - anaphylaxis, p-value derived from the Fisher exact test.
+ANA - anaphylaxis, p-value derived from the Fisher's exact test.
 ",
@@
 knitr::kable(therapyTab[,1:4], caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
-2nd line therapy was performed after the initial rescue individual called for a professional medical emergeny team.
-ANA - anaphylaxis, p-value derived from the Fisher exact test.
+'2nd line' refers to the therapy performed by the professional medical emergency team after the initial rescue procedures.
+ANA - anaphylaxis, p-value derived from the Fisher's exact test.
 ",
@@
 knitr::kable(cofactorsTabF[,1:4], caption = "Factors potentially increasing the risk of a severe anaphylaxis investigated in refractory cases.
-ANA - anaphylaxis, p-value derived from the Fisher exact test.
+ANA - anaphylaxis, p-value derived from the Fisher's exact test.
 ",

### 2018-09-18 R/analysis.R

@@ -723,15 +723,15 @@ nodes <-
     type = "a",
     label= c(paste0("All cases\n",length(data3$b_submitdate)),
              paste0("At least 2 doses of Epinephrine\nn = ", length(which(data3$d_adre_no>=2))),
-             paste0("Severe reaction \nrequiring hospitalisation\nn = ",length(which(refractory == "yes"))),
+             paste0("Severe reaction \nrequiring hospitalization\nn = ",length(which(refractory == "yes"))),
              paste0("Treatment with at least\ntwo 2nd line drugs\nn = ", length(which(refractory.strict=="yes"&refractory.ICU=="no"))),
              paste0("Final number of cases included in the study\nn = ",length(which(refractory.f=="yes"))-R1-R2-R3),
              paste0("Fatal reaction\nn = ",length(which(refractory.death=="yes"))),
              paste0("Reactions requiring\nICU\nn = ",length(which(refractory.ICU=="yes"&refractory.death=="no"))),
              paste0("Cases eliminated after manual revision:\nMultiple elicitors (n = 1)\n",
                     "Biphasic reactions responsive to Adrenaline (n = ", R1-1,")\n",
-                    "Long time between adrenalin doses indicating responiveness (n = ",R2,")\n",
-                    "Adequate response after second dose of adrenaline (n = ",R3,")")),
+                    "Extended time between adrenalin doses indicating responiveness (n = ",R2,")\n",
+                    "Adequate response after a second dose of adrenaline (n = ",R3,")")),
     #color = c("red", "green",
     #          "grey", "blue"),
     #value = c(3.5, 2.6, 9.4, 2.7),

### 2018-11-21 analysis/paper/paper.Rmd
Final version after corrections from coauthors

@@
 author:
-  - Wojciech Francuzik*
-  - Sabine Dölle-Bierke*
-  - Franziska Ruëff**
-  - Margitta Worm*
+  - Wojciech Francuzik^1^
+  - Sabine Dölle-Bierke^1^
+  - Macarena Knop^2^
+  - Kathrin Scherer Hofmeier^3^
+  - Ewa Cichocka-Jarosz^4^
+  - Blanca E. García^5^
+  - Roland Lang^6^
+  - Ioana Maris^7^
+  - Jean-Marie Renaudin^8^
+  - Margitta Worm^1^ 
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 require(magrittr)
+options(scipen=1, digits=4)
+#require(pander)
+#panderOptions('digits', 2)
+#panderOptions('round', 2)
+#panderOptions('keep.trailing.zeros', TRUE)
 ```
@@
   as.factor %>%  summary()
+nperiopAna <- data3$rANA[which(data3$q_152_location!="medical practice, hospital")] %>% as.factor %>% summary
 
@@
 
-*Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
+^1^ Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,
 
-**Klinik und Poliklinik für Dermatologie und Allergologie, Frauenlobstraße 9–11, 80337 München
+^2^ Department of Dermatology and Allergology, Klinikum der Universität München, Munich, Germany
+
+^3^ Department of Dermatology, University Hospital Basel, Basel, Switzerland
+
+^4^ Department of Pediatrics, Jagiellonian University Medical College, Krakow, Poland
+
+^5^ Service of Allergology, Complejo Hospitalario de Navarra, Pamplona, Spain
+
+^6^ Department of Dermatology, Paracelsus Private Medical University Salzburg, Salzburg, Austria
+
+^7^ Bon Secours Hospital, Cork/Department of Paediatrics and Child Health, University College Cork, Ireland
+
+^8^ Réseau d’Allergo-Vigilance (Allergy Vigilance Network), Vandoeuvre les Nancy, France
 
@@
 Comprehensive data on its definition, prevalence, and risk factors are missing.
-Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified refractory anaphylaxis cases (n = `r v$casesR`) and analysed these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
+Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified refractory anaphylaxis cases (n = `r v$casesR`) and analyzed these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
 
-The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) than in other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).
-Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees in refractory cases (`r 5/8*100`% vs `r round(449/(77+1607+4+9+170+3+449)*100,1)`%, p = `r fisher.test(matrix(c(449,5,77+1607+4+9+170+3,3),nrow=2))$p.value %>% round(3)`).
-The refractory cases occurred more frequently in a perioperative setting (`r v$perioperative[3]`% vs.
+The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) compared to other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).
+Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees than wasps in refractory cases (`r 5/8*100`% vs `r round(449/(77+1607+4+9+170+3+449)*100,1)`%, p = `r fisher.test(matrix(c(449,5,77+1607+4+9+170+3,3),nrow=2))$p.value %>% round(3)`).
+The refractory cases occurred mostly in a perioperative setting (`r v$perioperative[3]`% vs.
 `r v$perioperative[2]`, p < 0.0001).
@@
 Second line treatment options (e.g.
-vasopression with dopamine, methylene blue, glucagon) were not used at all in the treatment of refractory cases.
+vasopression with dopamine, methylene blue, glucagon) were not used at all for the treatment of refractory cases.
 The mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).
 
-Although multiple guidelines on anaphylaxis have been published, the clinical adherence to these guidelines can be improved.
-The neglegence of second line medication (i.e.
-methylene blue or dopamine) in refractory cases has to be addressed in future efforts to improve the managment (and guidelines adherence) of severe refractory anaphylaxis cases.
+Refractory anaphylaxis is associated with drug induced anaphylaxis in particular if given intravenously.
+Although adrenaline (even i.v.) is frequently applied in these patients, not all patients are responding.
+Whether a delay in recognition of anaphylaxis is responsible for the refractory case or whether these cases are due to an overflow with mast cell activating substances - requires further studies.
+Reasons for the low use of second line medication (i.e.
+methylene blue or dopamine) in refractory cases are unknown, but their use might improve the outcome of severe refractory anaphylaxis cases.
 
@@
 The mainstay of anaphylaxis management is the intramuscular dose of adrenaline [@Grabenhenrich2018], but in the most severe cases of anaphylaxis it might be insufficient to restore a stable patient status.
-Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least 2 doses of minimum 300 µg adrenaline, does not achieve normalization of clinical symptoms.
+Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least 2 doses of minimum 300 µg adrenaline, does not achieve normalization of the clinical symptoms in a given individual.
 Typical elicitors and symptoms of refractory anaphylaxis, as well as the therapeutic strategy for the most severe cases, differ from the usual reactions [@Francuzik2018] and call for a specific research and targeted guideline development for refractory anaphylaxis cases.
 
-We aimed to distinguish the prevalence of refractory anaphylaxis among anaphylaxis cases and to describe its symptoms and factors which may increase the risk of a refractory anaphylaxis episode.
+We aimed to distinguish the prevalence of refractory anaphylaxis among anaphylaxis cases and to describe symptoms and factors which may increase the risk of a refractory anaphylaxis episode.
 
 # Methods
-The European Anaphylaxis Registry (described in detail elsewhere [@Grabenhenrich2016]) provided data for this analysis status from May 2018.
+The European Anaphylaxis Registry (described in detail elsewhere [@Grabenhenrich2016]) provided data for this analysis status from March 2018.
 We selected cases where patients received at least two doses of adrenaline, and failed to recover adequately and assigned them to the "refractory anaphylaxis group".
@@
 We compared these to a group of severe, non-refractory cases of anaphylaxis.
-Severe reactions were identified based on the definition by NIAID/FAAN [@Sampson2006] and presented with significant hypoxia, hypotension, confusion, collapse, loss of consciousness, or incontinence.
-We compared the frequency of various elicitors, symptoms, and factors known to incease the risk of severe anphylaxis [@Worm2018] in both groups. We also compared how the cases were managed.
+Severe reactions were identified based on the definition by NIAID/FAAN [@Sampson2006] and presented with significant hypoxia, hypotension, confusion, collapse and loss of consciousness, or incontinence.
+We compared the frequency of various elicitors, symptoms, and factors known to increase the risk of severe anaphylaxis [@Worm2018] in both groups. We also analyzed how the cases were managed.
 
@@
 The statistical analysis was performed in the R Statistical Package [@R].
-Simple comparison of categorical variables was performed using Fisher's exact test, continous variables were analysed using Mann-Whithney U test.
+Simple comparison of categorical variables was performed using Fisher's exact test, continuous variables were analyzed using Mann-Whitney U test.
 We defined statistical significance as α = 0.05.
@@
 Each year approximately 1% (`r table(data3$d_year_reaction, data3$rANA) %>% {cbind(.[,1]/(.[,1]+.[,2]),.[,2]/(.[,1]+.[,2]))[which(rownames(.) %in% 2007:2017),2]} %>% {paste0(mean(.) %>% signif(3)*100, "% $\\pm$ ", sd(.) %>% signif(3)*100,"%")}`) of severe anaphylactic episodes are refractory to treatment with adrenaline.
-When considering patients undergoing a medical procedure (who experience anaphylaxis in a perioperative setting), nearly `r (periopAna[2]/(periopAna[2]+periopAna[1])) %>% signif(3)*100`% (`r (periopAna[2]/(periopAna[2]+periopAna[1]) %>% signif(3)*100 / 0.853) %>% signif(3)` times more) patients present with reactions that do not respond to adrenaline.
+When considering patients who experienced anaphylaxis in a perioperative setting or a medical facility, nearly `r (periopAna[2]/(periopAna[2]+periopAna[1])) %>% signif(3)*100`% patients present with reactions that do not respond to adrenaline vs.
+`r (nperiopAna[2]/(nperiopAna[2]+nperiopAna[1])) %>% signif(3)*100`% in non-medical setting (`r 3.72/0.448 %>% signif(1)` times more).
 
@@
 The percentage of males within the refractory anaphylaxis group was `r (as.numeric(demoTab[2,2])/sum(as.numeric(demoTab[2,2])+as.numeric(demoTab[1,2]))*100) %>% {round(.,1)}`%.
-Significantly more patients suffered from a concomitant malignant disease in the refractory anaphylaxis group.
+More patients suffered from a concomitant malignant disease in the refractory anaphylaxis group.
 Most strikingly, patients with refractory reactions more often had a previous anaphylactic reaction in their medical history (p = `r data3$q_410_prev_v5 %>% table(data3$rANA) %>%{.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% signif(3)`).
+Baseline tryptase levels were significantly higher in the refractory anaphylaxis group as 7 refractory patients (`r {7/41*100} %>% round(1)`%) had tryptase level above 11.5 µg/L (vs.
+`r summary(sdb$d_210_tryptase_cat) %>% {.[2]/sum(.[1:3])} %>% {.*100} %>% round(1)`%)).
 The demographic summary of refractory cases is shown in table \@ref(tab:demoTab).
@@
 Refractory anaphylaxis was most commonly elicited by drugs (significantly more often than in severe, non-refractory cases), followed by food and insects (table \@ref(tab:elicitorTab)).
-The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics (`r elicitExact[1,3]`%) and Radio-contrast media (RCM, `r elicitExact[2,3]`%).
+The most common drugs eliciting anaphylaxis refractory to adrenaline were antibiotics (`r elicitExact[1,3]`%) and radiocontrast media (RCM, `r elicitExact[2,3]`%).
 Patients with refractory anaphylaxis more frequently experienced the reaction while undergoing a medical procedure (`r (locTab[3,2]/sum(locTab[,2])*100) %>% {signif(.,3)}`% vs.
@@
 
-`r elicitorTab[1,8]`% of food elicited cases had a previously confirmed diagnosis of food allergy. Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
+`r elicitorTab[1,8]`% of food elicited refractory cases had a previously confirmed diagnosis of food allergy. Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
 Insect-venom-elicited, refractory cases were significantly less often elicited by yellow-jacket-stings (but not by bee-stings) than severe, non-refractory cases (table \@ref(tab:elicitExact)).
 
-## Refractory anaphylaxis is life-threathening
+## Refractory anaphylaxis is life-threatening
 <!-- ## Symptoms -->
@@
 <!-- ## Death  -->
-Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the cardiopulmonary resuscitation was unsuccessful.
-In cases where patients responded to life support, but failed to be reanimated due to post-resuscitative complication (e.g.
-hypoxic brain injury), death occured in the next 3-8 days.
+Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the cardiopulmonary resuscitation was unsuccessful and were highly associated with refractory cases (`r rcalc("q_140_fatal")[3]`% vs.
+only `r rcalc("q_140_fatal")[2]`% of severe anaphylaxis cases, p < 0.0001) In cases where patients responded to life support, but failed to be reanimated due to post-resuscitative complication (e.g.
+hypoxic brain injury), death occurred in the next 3-8 days.
 
@@
 <!-- ## Cofactors -->
-Patients with refractory anaphylaxis more often had concomitant asthma and malignant diseases in medical history.
+Patients with refractory anaphylaxis more often had concomitant asthma and malignant diseases in their medical history.
 Also other unspecified concomitant conditions were significantly more often reported in refractory cases.
-It is worth noticing that concomitant cardiologic conditions, diabetes and mastocytosis were similarly frequent in both groups.
+Concomitant cardiologic conditions, diabetes and mastocytosis were similarly frequent in both groups.
 
-Patients with refractory anaphylaxis more often reported concomitant proton pump inhibitors (PPI) and acetylsalicilic acid (ASA) use compared to patients with severe non-refractory anaphylaxis.
+Patients with refractory anaphylaxis more often reported concomitant intake of proton pump inhibitors (PPI) and acetylsalicilic acid (ASA) compared with patients with severe non-refractory anaphylaxis.
 Other (not specified) medications were also more frequent in refractory cases.
-In `r table(data3$rANA,data3$q_423_beta)[2,2]` cases of refractory anaphylaxis (`r (table(data3$rANA,data3$q_423_beta)[2,2]/sum(table(data3$rANA,data3$q_423_beta)[2,1:3])*100) %>% round(1)`%) patients reported receiving beta-blockers as a concomitant medication, but none of these patients recieved a glucagon infusion.
+In `r table(data3$rANA,data3$q_423_beta)[2,2]` cases of refractory anaphylaxis (`r (table(data3$rANA,data3$q_423_beta)[2,2]/sum(table(data3$rANA,data3$q_423_beta)[2,1:3])*100) %>% round(1)`%) patients reported receiving beta-blockers as a concomitant medication, but none of these patients received a glucagon infusion.
 
-The intensity of physical exercise exceeding the reaction was indifferent between groups, however the psychological burden was reported 3 times more frequently in refractory cases (see table \@ref(tab:cofactorsTab)).
+The intensity of physical exercise exceeding the reaction was indifferent between groups, however psychological burden (defined as stressfull event preceeding the reaction, rated by the physician) was reported 3 times more frequently in refractory cases (see table \@ref(tab:cofactorsTab)).
 
@@
 
-## Frequency and patients at risk
+## Frequency of refractory anaphylaxis and patients at risk
 Our findings suggest that around 1 in 100 severe anaphylaxis patients will not respond to the standard therapy with adrenaline, commonly outlined in anaphylaxis management guidelines [@Muraro2014].
 Such cases have the highest risk of a fatal outcome, and therefore need to be treated accordingly.
-An early use of drugs (e.g.
-methylene blue and dopamine) might increse their survival chance.
+An early use of adrenaline and maybe other second line drugs (e.g.
+methylene blue and dopamine) might increase their survival chance.
 
-Overall, the severity of an anaphylaxis and its probable transition into a refractory episode might be described by four co-influencing mechanisms: 
-1) elicitors - the type of an eliciting agent and duration of the exposition, which correlates to the amount of the allergen and its metabolism; 
-2) cofactors - the presence of other factors which may increase the severity of a given reaction i.e. concomitant use of a betablocker etc.;
-3) compensation - how well is the body able to compensate reaction symptoms e.g. hypotension; 
-4) therapy - how fast and accurately the therapeutic procedures were introduced (Fig. \@ref(fig:theory)).
+Overall, the severity of anaphylaxis and its probable transition into a refractory episode might be described by several co-influencing mechanisms: 
+1) elicitors - the type and dose of an eliciting agent and route of exposure; 
+2) cofactors - the presence of other individual and external factors which may increase the severity of a given reaction i.e. concomitant use of a beta-blocker etc.;
+3) compensation - how well the affected individual can compensate reaction symptoms e.g. hypotension; 
+4) therapy - how fast and accurately the therapeutic interventions were applied (Fig. \@ref(fig:theory)).
 <!-- This theory might explain why younger patients are able to better compensate the same trigger which would elicit a refractory reaction in eldery patients.  -->
@@
 Multiple medications with mast cell activation potential [@Hepner2003] given in a perioperative setting increase the chance of a hypersensitivity reaction and drug interactions [@Ebo2007].
-Recent study on IgE-independent anaphylaxis showed that peroperative drugs (i.e.
+A recent study on IgE-independent anaphylaxis showed that perioperative drugs (i.e.
 rocuronium, turbocurarine, fluorochinolones, atracurium) may trigger anaphylaxis by activating mast cells directly through the MRGPRX2 receptor [@Mcneil2015].
-Therefore patients who undergo surgical procedures and have a history of anaphylaxis should remain under extraordinary caution.
+Therefore, patients undergoing surgical procedures and having a history of anaphylaxis should remain under extraordinary caution.
 
@@
 RCM on the other hand, commonly elicit IgE-independent hypersensitivity episodes that are refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [@Greenberger1991].
-We previously saw RCM to be the most frequent elicitor of refractory anaphylaxis [@Francuzik2018].
-It might be that RCM promote  complement activation [@Wang2011] and thus - unspecific activation of multiple immune cell classes (mast cells, basophiles, platelets and neutophiles).
+We previously reported RCM to be the most frequent elicitor of refractory anaphylaxis [@Francuzik2018].
+It might be that RCM promote  complement activation [@Wang2011] and thus - unspecific activation of multiple immune cell classes (mast cells, basophils, platelets and neutophils).
 Therefore, treatment with adrenaline may be less effective in these reactions.
@@
 ## Risk factors 
-Although we have previously identified older age and male sex to be associated with severe anaphylaxis [@Worm2018], these factors seem to be less important when refractory anaphylaxis is considered.
-Asthma and malignant diseases in the medical history were more frequently present in the refractory anaphylaxis group.
-The underlying reason is unclear.
+Although we have previously identified higher age and male sex to be associated with severe anaphylaxis [@Worm2018], these factors seem to be less important when refractory anaphylaxis is considered.
+By contrast, concomitant asthma was previously associated with less severe anaphylaxis in our registry data [@Worm2018].
+However, the present analysis as well as other previous data [@Mullins2003] suggest concomitant asthma as a risk factor for a severe episode.
+Probably the control of the asthmatic condition is essential in this context and should be therefore evaluated in patients at risk.
+Moreover, it might be possible that during a refractory episode the bronchospasm is additionally triggered via mediators derived from basophils and eosinophils or the vegetative nerve system what makes it refractory to adrenaline and results in a prolonged anaphylactic episode.
+Following this line we observed more respiratory distress symptoms in the refractory anaphylaxis cases.
 
@@
 ```
-It is however interesting that past malignant diseases were mostly recognized in male refractory anaphylaxis patients.
-Many cases of hypersenitivity due to cancer drugs were described in the literature, but malignancy in medical history has never been previously associated with severe anaphylaxis.
-It is possible that patients with reported malignancies were simply older (mean age of `r mean_age_malig[[1]] %>% round(1)` vs `r mean_age_malig[[2]] %>% round(1)` years, p <
-<!-- `r fisher.test(data3$d_age[which(data3$rANA=="yes")],data3$q_410_malig_prev_v5[which(data3$rANA=="yes")])`  --> 
-0.001) and therefore less able to compensate the anaphylactic shock.
-
-We previously observed concomitant asthma to be associated with less severe  anaphylaxis [@Worm2018].
-However, the present analysis suggests concomitant asthma as a risk factor for a refractory episode.
-It is possible that during a refractory episode the bronchospasm is mediated via other mediators (deriving from basophiles, and eosinophiles) what makes it refractory to adrenaline and therefore results in a prolonged anaphylactic episode.
-This might also explain why we saw more respiratory distress symptoms in the refractory anaphylaxis cases.
+The higher rate of malignancy in patients with refractory anaphylaxis diseases might be a confounder as certain chemotherapeutics are given intravenously and are well known to activate mast cells (e.g.
+taxanes) [@lee2009].
+On the other hand, it is possible that patients with reported malignancies were older (mean age of `r mean_age_malig[[1]] %>% round(1)` vs `r mean_age_malig[[2]] %>% round(1)` years, p <  0.001) and therefore less able to compensate the anaphylactic shock.
 
@@
 These mediators may contribute to an increased mast cell response and more severe anaphylaxis.
+Some published case reports indicated for stress as the anaphylaxis trigger [@alevizos2014].
 
-<!-- We have previously proven the role of ASA in increasing the risk of severe anaphylaxis [@].  Could not find the reference -->
 The role of PPI as a risk factor for severe anaphylaxis has been indicated in prolonging the exposition to an oral allergen (preventing its degradation due to lower gastric acid production) [@Wlbing2013].
-Moreover, there are reports on PPI as elicitors of anaphylaxis [@Natsch2000].
+Moreover, there are reports on PPI themselves as elicitors of anaphylaxis @Natsch2000].
 We observed a significantly higher rate of PPI intake in the refractory anaphylaxis group.
@@
 We and others have shown that older age may correlate with the decreased abillity to retain homeostasis on increased strain [@fleg1995].
-Patients who underwent refractory anahpylaxis more often had perioperative reactions  and therefore a decreased ablity to compensate the reaction symptoms with reflexory renal or cardiopulmonary response [@Rogers2013].
+Patients who underwent refractory anaphylaxis more often had perioperative reactions  and therefore a decreased ability to compensate the reaction symptoms with reflexory renal or cardiopulmonary response [@Rogers2013].
+
+Compensation mechanisms demonstrated in animal models indicated that anaphylactic hypotension activates the hypothalamic paraventricular nucleus, medullary nucleus tractus solitarii and rostral ventrolateral medulla, independently of the baroreflex pathway.
+Further, it stimulates efferent sympathetic nerve activity to the adrenal gland and kidney to restore blood pressure [@tanida2018].
 
@@
 Grabenhenrich et al.
-[@Grabenhenrich2018] recently evaluated the epinephrine use in anaphylaxis patients and concluded that, even in this state of the art drug, large discrepancies between recommenden use and actual treatment practice exist.
+[@Grabenhenrich2018] recently evaluated the epinephrine use in anaphylaxis patients and concluded that, even in this state of the art drug, large discrepancies between recommended use and actual treatment practice exist.
+Similarly, US studies documented poor adherence in patients and caregivers to anaphylaxis guidelines recommending more than 1 adrenaline autoinjector available at all times [@song2018].
 Therefore, more effort needs to be dedicated to promote and develop the consensus guidelenes as practically as possible in order to increase adherence.
@@
 within minutes.
-Its mode of action is by blocking the guanylate cyclase and therefore preventing further nitric-oxide-dependent vasodilation in a distributive shock [@dfaz].
+Methylene blue blocks the guanylate cyclase and therefore prevents further nitric-oxide-dependent vasodilation in a distributive shock [@dfaz].
+Its potential role in neuroprotection has also been indicated [@tucker2017].
 
 Suprisingly, even though multiple anaphylaxis management guidelines recommend glucagon infusions in cases of concomitant beta-blocker therapy in anaphylactic patients [@Simons2011;@Lieberman;@Muraro2014], it has not been administered in any of the severe or refractory cases.
+Similarly, Royal Collage of Anesthetists reported one glucagon administration in 266 severe intraoperative anaphylaxis episodes [@harper2018].
 Glucagon has been reported to sucessfully relive refractory anaphylaxis [@Zaloga1986;@Javeed1996] by directly activating the adenylyl cyclase and therefore bypassing the β-adrenergic receptor blockade [@Zaloga1986].
@@
 The low number of refractory cases prevented us from analyzing the data with more advanced statistical models.
-However, our analysis is the first published report on a patient sample containing over 30 refractory anaphylaxis cases.
+However, our analysis is the first report on a patient cohort exceeding 30 refractory anaphylaxis cases.
 
-Definition  of refractory anaphylaxis is not universal. The confirmation of a refractory anaphylaxis based on the answers to our online questionnaire posed difficulties.
-If a fatal reaction occured before the second dose of adrenaline was administered to the patient - it could not be diagnosed as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although, it in fact could have been refractory.
+There is no universal definition  of refractory anaphylaxis is not universal.
+The confirmation ofrefractory anaphylaxis is based on the answers to our online questionnaire.
+If a fatal reaction occurred before the second dose of adrenaline was administered to the patient - it was not categorized as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although, it in fact might have been refractory.
 
-Comparison of refractory anaphylaxis with patients suffering from severe anaphylaxis enabled us to distinguish patients with a higher risk of developing a refractory episode.
+The comparison of refractory anaphylaxis with patients suffering from severe anaphylaxis enabled us to distinguish patients with a higher risk of developing a refractory episode.
 However we cannot address the question which patients had a higher risk of experiencing anaphylaxis per se.
@@
 Refractory anaphylaxis is a rare form of a life-threatening hypersensitivity reaction with high mortality.
-Its elicitors and cofactors differ from other anaphylaxis cases, and the managment of refractory anaphylaxis needs to improve.
+Its elicitors and cofactors differ from other anaphylaxis cases, and the management of refractory anaphylaxis needs to improve.
 Although, more studies need to be conducted to better understand the pathomechanisms involved in refractory anaphylaxis, we propose to increase the use of second line medication such as methylene blue, vasopressin and (in suspicion of a beta adrenergic blockade) glucagon in cases where 2 doses of adrenaline did not result in a rapid normalization of anaphylaxis symptoms.
@@
 # Acknowledgements
-We thank the members of The European Anaphylaxis Registry.
-<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
+We thank all patients, parents, and their children for their support in providing data on the
+occurrence of anaphylaxis for this study. We thank the study personnel for patients
+counseling and data entry and we thank the members of The European Anaphylaxis Registry
+in detail:
+J. Grünhagen (Berlin, Germany),
+K. Beyer (Berlin, Germany),
+A. Möser (Jena, Germany),
+T. Fuchs (Göttingen, Germany),
+F. Ruëff (Munich, Germany),
+B. Wedi (Hanover, Germany),
+F. Friedrichs (Aachen, Germany),
+H. Dickel (Bochum, Germany),
+H. Merk (Aachen, Germany),
+U. Hillen (Essen, Germany),
+A. Bauer (Dresden, Germany),
+N. Wagner (Erlangen, Germany),
+E. Rietschel (Cologne, Germany),
+R. Treudler, S. Aurich (Leipzig, Germany),
+L. Klimek, O. Pfaar (Wiesbaden, Germany),
+N. Reider (Innsbruck, Germany),
+W. Aberer (Graz, Austria),
+A. Köhli (Zurich, Switzerland),
+F. Riffelmann (Schmallenberg, Germany),
+B. Kreft (Halle, Germany),
+K. Nemat (Dresden, Germany),
+T. Kinaciyan (Vienna, Austria),
+R. Brehler (Münster, Germany),
+J. Witte (Hamburg, Germany),
+N. Hunzelmann und I. Huseynow (Cologne, Germany),
+T. Bieber (Bonn, Germany),
+U. Rabe (Treuenbritzen, Germany),
+P. Schmid-Grendelmeier (Zurich, Switzerland),
+W. Brosi (Würzburg, Germany),
+S. Nestoris (Lippe-Lemgo, Germany),
+T. Hawranek (Salzburg, Austria),
+R. Bruns (Greifswald, Germany),
+S. Lehmann (Aachen, Germany),
+L. Lange (Bonn, Germany),
+G. Hansen (Hanover, Germany),
+C. Pföhler (Homburg, Germany),
+E. Varga (Graz, Austria),
+Z. Szepfalusi (Vienna, Austria),
+P. Eng (Aarau, Switzerland),
+P. Eng (Lucerne, Switzerland),
+T. Reese (Rheine, Germany),
+M. Polz (Rüsselsheim, Germany),
+S. Schweitzer-Krantz (Düsseldorf, Germany),
+H. Rebmann (Tübingen, Germany),
+G. Stichtenoth (Lübeck, Germany),
+S. Theis (Schwedt, Germany),
+I. Yildiz (Neumünster, Germany),
+M. Gerstlauer (Augsburg, Germany),
+A. Nordwig (Dresden, Germany),
+R. Schlags (Wangen im Allgäu, Germany),
+I. Neustädter (Fürth, Germany),
+C. Stadlin (Zurich, Switzerland),
+M. Bücheler (Bonn, Germany),
+S. Volkmuth (Velbert, Germany),
+J. Fischer (Tübingen, Germany),
+A. Henschel (Berlin, Germany),
+S. Plank-Habibi (Alzenau, Germany),
+B. Schilling (Passau, Germany),
+A. Kleinheinz (Buxtehude, Germany),
+K. Schäkel (Heidelberg, Germany),
+N. Papadopoulus (Athen, Greece),
+M. Kowalski (Lodz, Poland),
+M. Fernandez-Rivas (Madrid, Spain),
+K. Solarewicz-Madajek (Wroclaw, Poland),
+C. Körner-Rettberg (Bochum, Germany),
+T. Mustakov (Sofia, Bulgaria),
+K. Hartmann (Lübeck, Germany),
+C. Kemen (Hamburg, Germany),
+U. Miehe (Leipzig, Germany),
+C. Ebner (Vienna, Austria),
+S. Haak (Oldenburg, Deutschland),
+V. Cardona (Barcelona, Spain),
+S. Hämmerling (Heidelberg, Germany),
+E. Arroabarren (Pamplona, Spain),
+M. Bilo (Ancona, Italy),
+N. Cabañes Higuero (Toledo, Spain),
+A. Vega Castro (Guadalajara, Spain),
+I. Poziomkowska-Gęsicka (Szczecin, Poland),
+S. Büsing (Osnabrück, Germany),
+U. Klettke (Berlin, Germany),
+C. Virchow (Rostock, Germany),
+G. Christoff (Sofia, Bulgaria),
+U. Jappe (Borstel, Germany),
+K. Breuer (Hamburg, Germany),
+S. Müller und T. Jakob (Freiburg, Germany),
+H. Straube (Darmstadt, Germany),
+C. Vogelberg (Dresden, Germany),
+F. Knöpfel (Norderney, Germany),
+J. Hourihane (Cork, Ireland),
+B. Rogala (Silesia, Poland),
+A. Montoro (Madrid, Spain),
+A. Brandes (Frankfurt/Oder, Germany),
+A. Muraro (Padua, Italy),
+T. Buck und J. Büsselberg (Hanover-Misburg, Germany),
+N. Zimmermann (Potsdam, Germany),
+D. Hernandez (Valencia, Spain),
+P. Minale (Genua, Italy),
+J. Niederwimmer und B. Zahel (Linz, Austria),
+A. Fiocchi (Rome, Italy),
+A. Reissig (Gera, Germany),
+F. Horak (Vienna, Austria),
+N. Klossowiski (Düsseldorf, Germany),
+F. Eitelberger (Wels, Austria),
+H. Ott (Hanover, Germany),
+R. Asero (Milan, Italy),
+S. Pistauer (Sylt/Westerland, Germany),
+M. Geißler (Ribnitz-Damgarten, Germany),
+L. Ensina (Sao Paulo, Brazil),
+A. Plaza Martin (Barcelona, Spain),
+J. Meister (Aue, Germany),
+S. Hompes (Hamburg, Germany),
+S. Stieglitz (Wuppertal, Germany).
 
@@
 ```{r demoTab}
-knitr::kable(demoTab, caption = "Summary of the refractory anaphylaxis cases.
+knitr::kable(demoTab , caption = "Summary of the refractory anaphylaxis cases.
 Age is represented by a mean value, other variables as fractions [%].
@@
              align = rep("r",10),
-             nsmall=2)
+             nsmall=2) #%>% 
+  # kableExtra::kable_styling("striped")
 ```
@@
 ```{r elicitorTab}
-knitr::kable(elicitorTab[,c(1,5,6,7,8,2,3,4)], caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
-ANA - anaphylaxis, Age is represented as a mean, p-value derived from the Fisher's exact test.",
-             col.names = c("n","Age","Male sex [%]","Perioperative [%]","Food allergy [%]","refractory ANA [%]","severe ANA [%]","p value")
+knitr::kable(elicitorTab[,c(0,1,5,6,7,8,2,3,4)+1], caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
+ANA - anaphylaxis, Age is represented as a mean, p-value derived from the Fisher's exact test."
+             #col.names = c("n","Age","Male sex [%]","Perioperative [%]","Food allergy [%]","refractory ANA [%]","severe ANA [%]","p value")
+
             #digits = 1,
@@
 
-knitr::kable(cbind(c("Antibiotics","X-ray medium","Muscle relaxant","Legumes","Bee venom","Yellow-jacket venom"),elicitExact[,2:4]), caption = "Summary of the elicitors broken down into specific elicitors from each group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control.
+knitr::kable(data.frame(c("Antibiotics","X-ray medium","Muscle relaxant","Legumes","Bee venom","Yellow-jacket venom"),
+                   as.numeric(elicitExact[,2:4]) %>% matrix(nrow = 6)), 
+             caption = "Summary of the elicitors broken down into specific elicitors from each group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control.
 ANA - anaphylaxis, p-value derived from the Fisher's exact test.
 ",
-             col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value"))
+             col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value"),
+             digits = c(0,2,2,4)
+             )
 ```
@@
 ```{r symptTab}
-knitr::kable(symptTab[,1:4], caption = "Summary of the symptoms in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
+knitr::kable(symptTab[,1:4] %>% {data.frame(.[,1],.[,2] %>% as.numeric,.[,3] %>% as.numeric,.[,4] %>% as.numeric)}, caption = "Summary of the symptoms in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
 ANA - anaphylaxis, p-value derived from the Fisher's exact test.
 ",
-             col.names = c("Symptom","severe ANA [%]","refractory ANA [%]","p value"))
+             col.names = c("Symptom","severe ANA [%]","refractory ANA [%]","p value"),
+             digits = c(0,2,2,4))
 ```
@@
 ```{r therapyTab}
-knitr::kable(therapyTab[,1:4], caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
+knitr::kable(therapyTab[,1:4] %>% {data.frame(.[,1],.[,2] %>% as.numeric(),.[,3] %>% as.numeric,.[,4] %>% as.numeric())}, caption = "Summary of therapeutic measures in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
 '2nd line' refers to the therapy performed by the professional medical emergency team after the initial rescue procedures.
@@
 ",
-             col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"))
+             col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"),
+             digits = c(0,2,2,4))
 ```
@@
 ```{r cofactorsTab}
-knitr::kable(cofactorsTabF[,1:4], caption = "Factors potentially increasing the risk of a severe anaphylaxis investigated in refractory cases.
+knitr::kable(cofactorsTabF[,1:4] %>% {data.frame(.[,1],.[,2] %>% as.numeric(),.[,3] %>% as.numeric,.[,4] %>% as.numeric())}, caption = "Factors potentially increasing the risk of a severe anaphylaxis investigated in refractory cases.
 ANA - anaphylaxis, p-value derived from the Fisher's exact test.
 ",
-             col.names = c("Factor","severe ANA [%]","refractory ANA [%]","p value"))
+             col.names = c("Factor","severe ANA [%]","refractory ANA [%]","p value"),
+             digits = c(0,2,2,4))
 ```

### 2018-11-21 R/analysis.R

@@ -558,8 +558,18 @@ demoTabsP <- cbind(n = data3$b_sex %>% table(data3$rANA) %>% summary() %>% {.$p.
 
 )
 
+require(plyr)
 demoTab <- rbind(demoTab,demoTabs,demoTabsP)
 demoTab <- cbind(Group = c("refractory","refractory","severe","severe","p value"),demoTab)
+cols <- dimnames(demoTab)[[2]]
+demoTab %<>%  unlist %>% matrix(nrow = 5,byrow = F)
+demoTab %<>% as.data.frame()
+demoTab %<>% {data.frame(.[,1],apply(.[2:10],MARGIN = 2,FUN =  function(x){
+  as.character(x) %>% as.numeric()
+}))}
+demoTab %<>% {data.frame(Sex = c("female","male","female","male",""),.)}
+colnames(demoTab) <- c("Sex",cols)
+
 
 rdb$q_410_masto_cur%>% split(rdb$b_sex) %>% lapply(.,function(x){(length(which(x=="yes"))/length(x)*100) %>% signif(3)})
 rdb$q_410_asthma_cur%>% split(rdb$b_sex) %>% lapply(.,function(x){(length(which(x=="yes"))/length(x)*100) %>% signif(3)})
@@ -600,6 +610,17 @@ elicitorTab <- cbind(n = rdb$d_elicitor_gr5 %>% summary(),
                  `Food allergy` = rdb$q_410_foodallergy_cur_v6%>% split(rdb$d_elicitor_gr5) %>%
                    lapply(.,function(x){(length(which(x=="yes"))/length(x)*100) %>% signif(3)})
 )
+
+rn <- rownames(elicitorTab)
+cn <- colnames(elicitorTab)
+elicitorTab %<>%  unlist %>% matrix(nrow = 5,byrow = F) %>% as.data.frame()
+# elicitorTab %>% {data.frame(.[,1],apply(.[2:10],MARGIN = 2,FUN =  function(x){
+#   as.character(x) %>% as.numeric()
+# }))}
+elicitorTab %<>% {data.frame(Elicitor = rn,.)}
+colnames(elicitorTab) <- c("Elicitor",cn)
+
+
 
 #### 3. Risk Factors #####
 
@@ -646,12 +667,12 @@ therapyTab[,1] <-
     "volume, 2nd line",
     "antihistaminics i.v.",
     "antihistaminics i.v. 2nd line",
-    "corticosteoids, all routes",
[…]

### 2018-11-21 analysis/paper/paper.Rmd
hotfix tables

@@
 knitr::kable(elicitorTab[,c(0,1,5,6,7,8,2,3,4)+1], caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
-ANA - anaphylaxis, Age is represented as a mean, p-value derived from the Fisher's exact test."
+ANA - anaphylaxis, Age is represented as a mean, p-value derived from the Fisher's exact test.",
              #col.names = c("n","Age","Male sex [%]","Perioperative [%]","Food allergy [%]","refractory ANA [%]","severe ANA [%]","p value")
-
+              align = rep("r",10),
+             nsmall=2)
             #digits = 1,
@@
              #drop0trailing = F
-             )
+
 ```
@@
 ",
-             col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value"),
-             digits = c(0,2,2,4)
+             col.names = c("Elicitor","severe ANA [%]","refractory ANA [%]","p value")
+             #digits = c(0,2,2,4)
              )

### 2018-11-22 analysis/paper/paper.Rmd
changed the README

@@
 
-The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) compared to other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).
+The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab[elicitorTab$Elicitor=="drugs","percent"]`% of cases, p < 0.0001) compared to other severe anaphylaxis cases (`r elicitorTab[elicitorTab$Elicitor=="drugs","percANA"]`%).
 Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees than wasps in refractory cases (`r 5/8*100`% vs `r round(449/(77+1607+4+9+170+3+449)*100,1)`%, p = `r fisher.test(matrix(c(449,5,77+1607+4+9+170+3,3),nrow=2))$p.value %>% round(3)`).

### 2019-08-20 R/insect venom.Rmd
pre revisions from Frontiers

--- 2018-11-21 analysis/paper/paper.Rmd
@@
 ---
-title: "Refractory anaphylaxis: Data from the European Anaphylaxis Registry"
+title: "Insect-venom elicited anaphylaxis: Data from the European Anaphylaxis Registry"
 author:
@@
 nperiopAna <- data3$rANA[which(data3$q_152_location!="medical practice, hospital")] %>% as.factor %>% summary
-
 ```
@@
 <!-- 350 Words unstructured -->
-Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 µg adrenaline) is a rare and often fatal hypersensitivity reaction.
-Comprehensive data on its definition, prevalence, and risk factors are missing.
-Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified refractory anaphylaxis cases (n = `r v$casesR`) and analyzed these in comparison to a control group of severe anaphylaxis cases (n = `r v$casesC`).
+Insect-venom elicited anaphylaxis is a common and often severe hypersensitivity reaction.
+It affects older individuals than other anaphylaxis causes (p = `r t.test(data$d_age[data$d_elicitorInsect=="yes"],data$d_age[data$d_elicitorInsect=="no"])$p.val`).
+Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified insect-venom elicited anaphylaxis cases (n = `r v$insectElicited`) and analyzed these in comparison to anaphylaxis elicited by other elicitors (n = `r v$otherEli`).
 
-The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab["drugs","percent"]`% of cases, p < 0.0001) compared to other severe anaphylaxis cases (`r elicitorTab["drugs","percANA"]`%).
-Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees than wasps in refractory cases (`r 5/8*100`% vs `r round(449/(77+1607+4+9+170+3+449)*100,1)`%, p = `r fisher.test(matrix(c(449,5,77+1607+4+9+170+3,3),nrow=2))$p.value %>% round(3)`).
-The refractory cases occurred mostly in a perioperative setting (`r v$perioperative[3]`% vs.
-`r v$perioperative[2]`, p < 0.0001).
-Intramuscular adrenaline (as a first line therapy) was administered in `r therapyTab[1,3]`% of refractory cases, whereas in `r therapyTab[2,3]`% of cases it was applied intravenously (significantly more often than in severe anaphylaxis cases: `r therapyTab[2,2]`%, p < 0.0001).
-Second line treatment options (e.g.
-vasopression with dopamine, methylene blue, glucagon) were not used at all for the treatment of refractory cases.
-The mortality rate in refractory anaphylaxis was significantly higher (`r v$mortality[3]`%) than in severe cases (`r v$mortality[2]`%, p < 0.0001).
+Patients who experienced insect-venom anaphylaxis were older (p < 0.0001), more often had mastocytisis (p < 0.0001) and cardiologic conditions (p < 0.0001) and females more often had concomitant thyroid disease, and less often sufferd from food allergy or atopic dermatitis.
 
-Refractory anaphylaxis is associated with drug induced anaphylaxis in particular if given intravenously.
+The data show that `r v$insectsExact[7]/sum(v$insectsExact)*100`% of all insect elicited cases were elicited  by yellow jackets, followed by bees (`r v$insectsExact[2]/sum(v$insectsExact)*100`%).
+The insect venom elicited cases occurred mostly in outdoor places (`r v$locat[4,3]/sum(v$locat[,3])*100`%) patients' homes (`r v$locat[9,3]/sum(v$locat[,3])*100`%) or urban places (`r v$locat[5,3]/sum(v$locat[,3])*100`%).
+Intramuscular adrenaline (as a first line therapy) was administered significantly less often in insect venom elicited cases (`r therapyTab[1,2]`%, p < 0.0001).
+The mortality rate in insect anaphylaxis was comparable (`r v$mortality[3]`%) to other cases (`r v$mortality[2]`%, p < `r v$mortality[4]`).
+
+Insect venom anaphylaxis is associated with drug induced anaphylaxis in particular if given intravenously.
 Although adrenaline (even i.v.) is frequently applied in these patients, not all patients are responding.
@@
 
+```
+
+## Is elevated serum tryptace aposing risk for severe anaphylaxis in the venom elicited cases more than in other cases?
+
+```{r}
+data$d_210_tryptase_cat %>% table(data$d_elicitorInsect)
+plot(data$d_severity_rm)
+
+plot(data$d_severity_rm[data$d_210_tryptase_cat=="positive"&data$d_elicitorInsect=="yes"])
+plot(data$d_severity_rm[data$d_210_tryptase_cat=="positive"&data$d_elicitorInsect=="no"])
+ggplot(data[data$d_210_tryptase_cat=="positive"],)
 ```
@@
 ",
-             col.names = c("Therapy","severe ANA [%]","refractory ANA [%]","p value"),
+             col.names = c("Therapy","insect-elicited [%]","other [%]","p value"),
              digits = c(0,2,2,4))
@@
 ```
+
+# Figures
+
+```{r age_elicitor_plot, fig.cap="The density of anaphylaxis cases elicited by either insects or other factors according to age"}
+age_elicitors_plot
+```
+
+```{r tryptase_plot, fig.cap="Tryptase values according to insect elicitors."}
+tryptase_plot
+```

### 2019-08-20 R/Insects.R

@@ -11,7 +11,7 @@ require(magrittr)
 #load("analysis/data/raw_data/data3.R")
 # New database
 load("analysis/data/raw_data/data4.R")
-data3 <- data4
+data <- data4
 
 ##### FUNCTIONS #####
 # Put all the functions above the code below,
@@ -26,28 +26,34 @@ sum_adrenalines <- function(vars,data){
     return()
 }
 
+d_elicitorInsect <- rep("NA",length(data$b_submitdate))
+d_elicitorInsect[data$d_elicitor_gr5!="unknown"] <- "no"
+d_elicitorInsect[data$d_elicitor_gr5=="insects"] <- "yes"
+data$d_elicitorInsect <- d_elicitorInsect
+
 rcalc <- function(var,level="yes"){
-  tab <- table(data3[,var],data3$rANA)
-  if(level %in% (data3[,var] %>% levels)){
+  tab <- table(data[,var],data$d_elicitorInsect)
+  if(level %in% (data[,var] %>% levels)){
     pn <- tab[level,"no"]
     pp <- tab[level,"yes"]
     l <- level
   } else {
     pn <- tab[2,"no"]
     pp <- tab[2,"yes"]
-    l <- levels(data3[,var])[2]
+    l <- levels(data[,var])[2]
   }
-  nn <- data3$rANA %>% factor %>% summary %>% {.[1]}-pn
-  np <- data3$rANA %>% factor %>% summary %>% {.[2]}-pp
+  nn <- data$d_elicitorInsect %>% factor %>% summary %>% {.[1]}-pn
+  np <- data$d_elicitorInsect %>% factor %>% summary %>% {.[2]}-pp
   eval <- c(pn,pp,nn,np) %>% matrix(byrow = T,ncol = 2)
   prt <- (prop.table(eval,2)*100) %>% {signif(.,3)}
   p <- eval %>% fisher.test() %>% .$p.value %>% roP
   return(c(var,prt[1,1],prt[1,2],p,l))
 }
 
+
 roP<- function(x){
   ifelse(x > 0.001,
-  round(x,3),ifelse(x==0,0,0.0001)
+         round(x,3),ifelse(x==0,0,0.0001)
   )
 }
 
@@ -56,7 +62,7 @@ f1 <- function(x){
   return(do.call(t.test,x)$p.value %>% roP)
 }
 f2 <- function(x){
[…]

### 2019-08-21 analysis/paper/paper.Rmd
add pval, change line 119

@@
 
-`r elicitorTab[1,8]`% of food elicited refractory cases had a previously confirmed diagnosis of food allergy. Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
-Insect-venom-elicited, refractory cases were significantly less often elicited by yellow-jacket-stings (but not by bee-stings) than severe, non-refractory cases (table \@ref(tab:elicitExact)).
+`r elicitorTab[1,9]`% of food elicited refractory cases had a previously confirmed diagnosis of food allergy. Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
+Refractory anaphylaxis cases elicited by insects were more often elicited by bees than severe, non-refractory cases, p = `r table(data3$q_340_insects,data3$rANA) %>%  {rbind(addmargins(.[3:8,],1) %>% tail(1),.[2,])} %>% as.matrix() %>% fisher.test() %>% {.$p.value} %>% round(4)` (table \@ref(tab:elicitExact)).
 
@@
 ```{r elicitorTab}
-knitr::kable(elicitorTab[,c(0,1,5,6,7,8,2,3,4)+1], caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
+knitr::kable(elicitorTab#[,c(0,1,5,6,7,8,2,3,4)+1]
+             , caption = "Summary of elicitors in the refractory anaphylaxis cases and severe, non-refractory anaphylaxis cases as a control.
 ANA - anaphylaxis, Age is represented as a mean, p-value derived from the Fisher's exact test.",

### 2019-08-21 R/analysis.R

@@ -45,6 +45,25 @@ rcalc <- function(var,level="yes"){
   return(c(var,prt[1,1],prt[1,2],p,l))
 }
 
+rcalc2 <- function(var,level="yes"){
+  tab <- table(data3[,var],data3$rANA)
+  if(level %in% (data3[,var] %>% levels)){
+    # pn <- tab[level,"no"]
+    # pp <- tab[level,"yes"]
+    l <- level
+   } else {
+    # pn <- tab[2,"no"]
+    # pp <- tab[2,"yes"]
+     l <- levels(data3[,var])[2]
+   }
+  # nn <- data3$rANA %>% factor %>% summary %>% {.[1]}-pn
+  # np <- data3$rANA %>% factor %>% summary %>% {.[2]}-pp
+  eval <- tab[1:2,1:2]
+  prt <- (prop.table(eval,2)*100) %>% {signif(.,3)}
+  p <- eval %>% fisher.test() %>% .$p.value %>% roP
+  return(c(var,prt[2,1],prt[2,2],p,l))
+}
+
 roP<- function(x){
   ifelse(x > 0.001,
   round(x,3),ifelse(x==0,0,0.0001)
@@ -58,6 +77,10 @@ f1 <- function(x){
 f2 <- function(x){
   x %>% table(data3$rANA) %>% {.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% roP
 }
+f2_rdbsdb <- function(x){
+  x %>% table(rdbsdb$rANA) %>% {.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% roP
+}
+
 f3 <- function(x){
   x %>%
     lapply(function(x){(length(which(x=="yes"))/length(x)*100) %>% roP})
@@ -507,9 +530,9 @@ v<- list()
 v$allCases <- length(data3$b_submitdate)
 v$casesR <- length(which(data3$rANA=="yes"))
 v$casesC <- length(which(data3$rANA=="no"))
-v$perioperative <- rcalc("perioperative")
+v$perioperative <- rcalc2("perioperative")
 v$casesInsect <- table(data3$q_340_insects[data3$rANA=="yes"])
-v$mortality <- rcalc("q_140_fatal")# ** Tables ----
+v$mortality <- rcalc2("q_140_fatal")# ** Tables ----
 # This section deals with tables. Prepare them here and reference to them in the paper.Rmd
 
[…]

### 2019-08-27 analysis/paper/paper.Rmd
gramarly check

@@
 
-library(RefractoryAnaOrg) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
+#library(RefractoryAnaOrg) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
+source("../../R/analysis.R")
 require(magrittr)
@@
 #panderOptions('keep.trailing.zeros', TRUE)
+rcalc2 <- function(var,level="yes"){
+  tab <- table(data3[,var],data3$rANA)
+  if(level %in% (data3[,var] %>% levels)){
+    # pn <- tab[level,"no"]
+    # pp <- tab[level,"yes"]
+    l <- level
+   } else {
+    # pn <- tab[2,"no"]
+    # pp <- tab[2,"yes"]
+     l <- levels(data3[,var])[2]
+   }
+  # nn <- data3$rANA %>% factor %>% summary %>% {.[1]}-pn
+  # np <- data3$rANA %>% factor %>% summary %>% {.[2]}-pp
+  eval <- tab[1:2,1:2]
+  prt <- (prop.table(eval,2)*100) %>% {signif(.,3)}
+  p <- eval %>% fisher.test() %>% .$p.value %>% roP
+  return(c(var,prt[2,1],prt[2,2],p,l))
+}
 ```
@@
 **Keywords:** 
-anaphylaxis, adrenaline (epinephrine), beta-blockers, insect venom allergy, drug allergic reactions, vasopression, refractory
+anaphylaxis, adrenaline (epinephrine), beta-blockers, insect venom allergy, drug allergic reactions, vasoconstriction, refractory
 
@@
 <!-- 350 Words unstructured -->
-Refractory anaphylaxis (unresponsive to treatment with at least 2 doses of minimum 300 µg adrenaline) is a rare and often fatal hypersensitivity reaction.
+Refractory anaphylaxis (unresponsive to treatment with at least two doses of minimum 300 µg adrenaline) is a rare and often fatal hypersensitivity reaction.
 Comprehensive data on its definition, prevalence, and risk factors are missing.
@@
 
-Refractory anaphylaxis is associated with drug induced anaphylaxis in particular if given intravenously.
-Although adrenaline (even i.v.) is frequently applied in these patients, not all patients are responding.
+Refractory anaphylaxis is associated with drug-induced anaphylaxis in particular if allergens are given intravenously.
+Although physicians frequently use adrenaline in cases of perioperative anaphylaxis, not all patients are responding to treatment.
 Whether a delay in recognition of anaphylaxis is responsible for the refractory case or whether these cases are due to an overflow with mast cell activating substances - requires further studies.
-Reasons for the low use of second line medication (i.e.
-methylene blue or dopamine) in refractory cases are unknown, but their use might improve the outcome of severe refractory anaphylaxis cases.
+Reasons for the low use of second-line medication (i.e., methylene blue or dopamine) in refractory cases are unknown, but their use might improve the outcome of severe refractory anaphylaxis cases.
 
@@
 This variability introduces a wide range of possible reaction-symptom severities.
-Therefore, multiple sub-types of anaphylaxis have been previously identified (i.e.
-food dependent exercise induced anaphylaxis, venom anaphylaxis, biphasic anaphylaxis).
+Therefore, multiple subtypes of anaphylaxis have been previously identified (i.e., food-dependent exercise-induced anaphylaxis, venom anaphylaxis, biphasic anaphylaxis).
 
-The mainstay of anaphylaxis management is the intramuscular dose of adrenaline [@Grabenhenrich2018], but in the most severe cases of anaphylaxis it might be insufficient to restore a stable patient status.
-Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least 2 doses of minimum 300 µg adrenaline, does not achieve normalization of the clinical symptoms in a given individual.
-Typical elicitors and symptoms of refractory anaphylaxis, as well as the therapeutic strategy for the most severe cases, differ from the usual reactions [@Francuzik2018] and call for a specific research and targeted guideline development for refractory anaphylaxis cases.
+The mainstay of anaphylaxis management is the intramuscular dose of adrenaline [@Grabenhenrich2018], but in the most severe cases of anaphylaxis, it might be insufficient to restore a stable patient status.
+Refractory anaphylaxis (although the established definition is lacking) might be defined as anaphylaxis meeting the criteria by NIAID/FAAN [@Sampson2006] which, after the treatment with at least two doses of minimum 300 µg adrenaline, does not achieve normalization of the clinical symptoms in a given individual.
+Common elicitors and symptoms of refractory anaphylaxis, as well as the therapeutic strategy for the most severe cases, differ from the usual reactions [@Francuzik2018] and call for specific research and targeted guideline development for refractory anaphylaxis cases.
 
@@
 The European Anaphylaxis Registry (described in detail elsewhere [@Grabenhenrich2016]) provided data for this analysis status from March 2018.
-We selected cases where patients received at least two doses of adrenaline, and failed to recover adequately and assigned them to the "refractory anaphylaxis group".
-The flowchart in figure \@ref(fig:flowchart) represents the detailed cases selection process.
+We selected cases where patients received at least two doses of adrenaline and failed to recover adequately and assigned them to the "refractory anaphylaxis group." The flowchart in figure \@ref(fig:flowchart) represents the detailed case-selection process.
 
@@
 Severe reactions were identified based on the definition by NIAID/FAAN [@Sampson2006] and presented with significant hypoxia, hypotension, confusion, collapse and loss of consciousness, or incontinence.
-We compared the frequency of various elicitors, symptoms, and factors known to increase the risk of severe anaphylaxis [@Worm2018] in both groups. We also analyzed how the cases were managed.
+We compared the frequency of various elicitors, symptoms, and factors known to increase the risk of severe anaphylaxis [@Worm2018] in both groups as well as their management.
 
-<!-- ## Data handling and statisitcs -->
-The statistical analysis was performed in the R Statistical Package [@R].
-Simple comparison of categorical variables was performed using Fisher's exact test, continuous variables were analyzed using Mann-Whitney U test.
+<!-- ## Data handling and statistics -->
+We performed a statistical analysis in the R Statistical Package [@R].
+A simple comparison of categorical variables was performed using Fisher's exact test; continuous variables were analyzed using the Mann-Whitney U test.
 We defined statistical significance as α = 0.05.
-Data along with the analysis script can be accessed at www.github.com/wolass/RefractoryAnaOrg.
+Data, along with the analysis script, can be accessed at www.github.com/wolass/RefractoryAnaOrg.
 
@@
 Most strikingly, patients with refractory reactions more often had a previous anaphylactic reaction in their medical history (p = `r data3$q_410_prev_v5 %>% table(data3$rANA) %>%{.[1:2,]} %>% fisher.test() %>% {.$p.value} %>% signif(3)`).
-Baseline tryptase levels were significantly higher in the refractory anaphylaxis group as 7 refractory patients (`r {7/41*100} %>% round(1)`%) had tryptase level above 11.5 µg/L (vs.
+Baseline tryptase levels were significantly higher in the refractory anaphylaxis group as 7 refractory patients (`r {7/42*100} %>% round(1)`%) had tryptase level above 11.5 µg/L (vs.
 `r summary(sdb$d_210_tryptase_cat) %>% {.[2]/sum(.[1:3])} %>% {.*100} %>% round(1)`%)).
@@
 
-`r elicitorTab[1,9]`% of food elicited refractory cases had a previously confirmed diagnosis of food allergy. Severe cases of anaphylaxis were most frequently elicited by insects and food triggers.
-Refractory anaphylaxis cases elicited by insects were more often elicited by bees than severe, non-refractory cases, p = `r table(data3$q_340_insects,data3$rANA) %>%  {rbind(addmargins(.[3:8,],1) %>% tail(1),.[2,])} %>% as.matrix() %>% fisher.test() %>% {.$p.value} %>% round(4)` (table \@ref(tab:elicitExact)).
+`r elicitorTab[1,9]`% of food elicited refractory cases had a previously confirmed diagnosis of food allergy. Insect-venom and food allergens most frequently elicited severe cases of anaphylaxis.
+Refractory anaphylaxis cases were more often elicited by bees than severe, non-refractory cases, p = `r table(data3$q_340_insects,data3$rANA) %>%  {rbind(addmargins(.[3:8,],1) %>% tail(1),.[2,])} %>% as.matrix() %>% fisher.test() %>% {.$p.value} %>% round(4)` (table \@ref(tab:elicitExact)).
 
@@
 <!-- ## Symptoms -->
-Milder anaphylaxis symptoms (i.e.
-pruritus, gastrointestinal symptoms, vertigo, chest and throat tightness) were significantly less present in refractory anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were more often associated with the refractory anaphylaxis cases.
+Milder anaphylaxis symptoms (i.e., pruritus, gastrointestinal symptoms, vertigo, chest, and throat tightness) were significantly less present in refractory anaphylaxis cases, whereas respiratory and cardiac arrest, as well as inspiratory and expiratory distress, and death were more often associated with the refractory anaphylaxis cases.
 Table \@ref(tab:symptTab) summarizes the most prominent differences in anaphylaxis symptoms among both groups.
 <!-- ## Death  -->
-Fatal reactions frequently occurred 30 minutes to 120 minutes after exposure - when the cardiopulmonary resuscitation was unsuccessful and were highly associated with refractory cases (`r rcalc("q_140_fatal")[3]`% vs.
-only `r rcalc("q_140_fatal")[2]`% of severe anaphylaxis cases, p < 0.0001) In cases where patients responded to life support, but failed to be reanimated due to post-resuscitative complication (e.g.
-hypoxic brain injury), death occurred in the next 3-8 days.
+Fatal reactions frequently occurred 30 minutes to 120 minutes after allergen exposure and were highly associated with refractory cases (`r rcalc2("q_140_fatal")[3]`% vs.
+only `r rcalc2("q_140_fatal")[2]`% of severe anaphylaxis cases, p < 0.0001) In cases where patients responded to life support, but failed to be reanimated due to post-resuscitative complication (e.g., hypoxic brain injury), death occurred in the next 3-8 days.
 
@@
 Corticosteroids i.v.
-were the second most frequently administered group of drugs in refractory cases (as a first and second line treatment), outpacing antihistaminic drugs and volume replacement therapy, and were significantly more often given in refractory cases.
+were the second most frequently administered group of drugs in refractory cases (as a first and second-line treatment), outpacing antihistaminic drugs and volume replacement therapy, and were significantly more often given in refractory cases.
 Volume replacement therapy was given initially in `r therapyTab[4,3]`% of refractory cases and was sustained only in `r therapyTab[5,3]`% as the therapy progressed in the hospital environment.
-<!--(possible counfounding variable by patients who died!!! they did not receive second line therapy and might need to be discarded from the analysis)-->
 
-Second line medication like dopamine, glucagon and methylene blue were not given in all refractory anaphylaxis cases as well as in severe non-refractory ones.
+Second-line medication like dopamine, glucagon, and methylene blue were not given in all refractory anaphylaxis cases as well as in severe non-refractory ones.
 However, patients with refractory anaphylaxis were more frequently admitted to the hospital (`r therapyTab[18,3]`%) and treated at the ICUs (`r therapyTab[19,3]`%).
@@
 Patients with refractory anaphylaxis more often had concomitant asthma and malignant diseases in their medical history.
-Also other unspecified concomitant conditions were significantly more often reported in refractory cases.
-Concomitant cardiologic conditions, diabetes and mastocytosis were similarly frequent in both groups.
+Also, other unspecified concomitant conditions were significantly more often reported in refractory cases.
+Concomitant cardiologic conditions, diabetes, and mastocytosis were similarly frequent in both groups.
 
@@
 
-The intensity of physical exercise exceeding the reaction was indifferent between groups, however psychological burden (defined as stressfull event preceeding the reaction, rated by the physician) was reported 3 times more frequently in refractory cases (see table \@ref(tab:cofactorsTab)).
+The intensity of physical exercise exceeding the reaction was indifferent between groups, however psychological burden (defined as a stressful event preceding the reaction, rated by the physician) was reported three times more frequently in refractory cases (see table \@ref(tab:cofactorsTab)).
 
@@
 Such cases have the highest risk of a fatal outcome, and therefore need to be treated accordingly.
-An early use of adrenaline and maybe other second line drugs (e.g.
-methylene blue and dopamine) might increase their survival chance.
+Early use of adrenaline and maybe other second-line drugs (e.g., methylene blue and dopamine) might increase their survival chance.
 
-Overall, the severity of anaphylaxis and its probable transition into a refractory episode might be described by several co-influencing mechanisms: 
+Overall, the severity of anaphylaxis and its probable transition into a refractory episode might depend on several co-influencing mechanisms: 
 1) elicitors - the type and dose of an eliciting agent and route of exposure; 
-2) cofactors - the presence of other individual and external factors which may increase the severity of a given reaction i.e. concomitant use of a beta-blocker etc.;
-3) compensation - how well the affected individual can compensate reaction symptoms e.g. hypotension; 
+2) cofactors - the presence of other individual and external factors which may increase the severity of a given reaction, e.g., concomitant use of a beta-blocker;
+3) compensation - how well the affected individual can compensate reaction symptoms, e.g., hypotension; 
 4) therapy - how fast and accurately the therapeutic interventions were applied (Fig. \@ref(fig:theory)).
-<!-- This theory might explain why younger patients are able to better compensate the same trigger which would elicit a refractory reaction in eldery patients.  -->
+<!-- This theory might explain why younger patients can compensate better the same trigger, which would elicit a refractory reaction in elderly patients.  -->
 
@@
 Multiple medications with mast cell activation potential [@Hepner2003] given in a perioperative setting increase the chance of a hypersensitivity reaction and drug interactions [@Ebo2007].
-A recent study on IgE-independent anaphylaxis showed that perioperative drugs (i.e.
-rocuronium, turbocurarine, fluorochinolones, atracurium) may trigger anaphylaxis by activating mast cells directly through the MRGPRX2 receptor [@Mcneil2015].
+A recent study on IgE-independent anaphylaxis showed that perioperative drugs (i.e., rocuronium, tubocurarine, fluoroquinolones, atracurium) might trigger anaphylaxis by activating mast cells directly through the MRGPRX2 receptor [@Mcneil2015].
 Therefore, patients undergoing surgical procedures and having a history of anaphylaxis should remain under extraordinary caution.
 
-Antibiotics, although commonly triggering IgE-dependent reactions [@saxon1983], are rarely reported in the literature as the cause of a refractory anaphylaxis.
-RCM on the other hand, commonly elicit IgE-independent hypersensitivity episodes that are refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [@Greenberger1991].
+Antibiotics, although commonly triggering IgE-dependent reactions [@saxon1983], are rarely reported in the literature as the cause of refractory anaphylaxis.
+RCM, on the other hand, commonly elicit IgE-independent hypersensitivity episodes that are refractory to adrenaline and are responsible for 1-5 deaths per 100,000 administrations  [@Greenberger1991].
 We previously reported RCM to be the most frequent elicitor of refractory anaphylaxis [@Francuzik2018].
-It might be that RCM promote  complement activation [@Wang2011] and thus - unspecific activation of multiple immune cell classes (mast cells, basophils, platelets and neutophils).
+It might be that RCM promotes complement activation [@Wang2011] and thus - unspecific activation of multiple immune cell classes (mast cells, basophils, platelets, and neutrophils).
 Therefore, treatment with adrenaline may be less effective in these reactions.
 
-Only few refractory cases were elicited by yellow-jacket-stings.
+Yellow-jacket-stings elicited only a few refractory cases.
 Therefore, the ratio of yellow-jacket to honey-bee venom elicited anaphylaxis was inverted in the refractory group.
-Reasons for this observation might include higher allergen exposition due to either longer exposition to the allergen (bee's sting remains in the skin) or the venom volume being 10 times higher than in a wasp sting [@Visscher1996].
-<!-- Patients should remove the bee sting as quickly as possible (optimal < 2 seconds) . -->
+Reasons for this observation might include higher allergen exposition due to either extended exposition to the allergen (bee's sting remains in the skin) or the venom volume being ten times higher than in a wasp sting [@Visscher1996].
 
 <!-- Mastoparans are a potent family of mast cell activators through G-protein coupled receptors (GPCRs).
-It is likely that their affinity to GPCRs might vary and the composition of mastoparans might depend on the species [source].
+Their affinity to GPCRs might likely vary, and the composition of mastoparans might depend on the species [source].
 -->
 
-<!-- Carpa ???-->
-
 ## Risk factors 
-Although we have previously identified higher age and male sex to be associated with severe anaphylaxis [@Worm2018], these factors seem to be less important when refractory anaphylaxis is considered.
-By contrast, concomitant asthma was previously associated with less severe anaphylaxis in our registry data [@Worm2018].
-However, the present analysis as well as other previous data [@Mullins2003] suggest concomitant asthma as a risk factor for a severe episode.
+Although we have previously identified higher age and male sex to be associated with severe anaphylaxis [@Worm2018], these factors seem to be less critical in refractory anaphylaxis cases.
+By contrast, we previously associated concomitant asthma with less severe anaphylaxis in our registry data [@Worm2018].
+However, the present analysis, as well as other previous data [@Mullins2003], suggest concomitant asthma as a risk factor for a severe episode.
 Probably the control of the asthmatic condition is essential in this context and should be therefore evaluated in patients at risk.
 Moreover, it might be possible that during a refractory episode the bronchospasm is additionally triggered via mediators derived from basophils and eosinophils or the vegetative nerve system what makes it refractory to adrenaline and results in a prolonged anaphylactic episode.
-Following this line we observed more respiratory distress symptoms in the refractory anaphylaxis cases.
+Following this line, we observed more respiratory distress symptoms in the refractory anaphylaxis cases.
+
+<!-- The influence of tryptase -->
+Baseline serum tryptase (BST) levels were increased in patients with refractory anaphylaxis, although only in 7 patients it was over the threshold level of 11.5 ng/mL - what is commonly recognized as elevated [@klion2003].
+BST may reflect pro-β tryptase due to increased body mast cell content [@CAUGHEY2006] or result from alpha-tryptasemia due to multiple copies of *TPSAB1* gene [@Lyons2018].
+Irrespective of the mechanism - increased BST has been shown to correlate with the severity of anaphylaxis [@Aniceto2019].
 
@@
 Stress induces the complement cascade activation [@Burns2008].
-Substance P, which is released during stress [@Arck2003], is a known vasodilator  [@Bos92] and can activate mast cells directly [@Singh1999].
+Substance P, released during stress [@Arck2003], is a known vasodilator  [@Bos92] and can activate mast cells directly [@Singh1999].
 These mediators may contribute to an increased mast cell response and more severe anaphylaxis.
@@
 
-The role of PPI as a risk factor for severe anaphylaxis has been indicated in prolonging the exposition to an oral allergen (preventing its degradation due to lower gastric acid production) [@Wlbing2013].
+Wölbling et al.
+indicated the role of PPI as a risk factor for severe anaphylaxis by prolonging the exposition to an oral allergen (preventing its degradation due to lower gastric acid production) [@Wlbing2013].
 Moreover, there are reports on PPI themselves as elicitors of anaphylaxis @Natsch2000].
@@
 
-<!--Low-Dose Ethanol Alters the Cardiovascular, Metabolic, and Respiratory Compensation for Severe Blood Loss-->
+Concomitant ASA use was also associated with refractory cases more frequently than in severe anaphylaxis.
+The role of ASA on increasing the severity of anaphylaxis has previously been implicated by increasing the intestinal absorption of allergens [@Wlbing2013].
+Nevertheless, as ASA might be both a cofactor and elicitor of anaphylaxis [@faria2014] and is highly correlated with another cofactor - concomitant cardiologic conditions - it is tough to evaluate its isolated influence on the anaphylaxis severity.
 
 ## Symptom compensation
-Age is the most important factor influencing the risk of developing severe anaphylaxis [@Worm2018].
-We and others have shown that older age may correlate with the decreased abillity to retain homeostasis on increased strain [@fleg1995].
-Patients who underwent refractory anaphylaxis more often had perioperative reactions  and therefore a decreased ability to compensate the reaction symptoms with reflexory renal or cardiopulmonary response [@Rogers2013].
+Age is the most critical factor influencing the risk of developing severe anaphylaxis [@Worm2018].
+We and others have shown that older age may correlate with the decreased ability to retain homeostasis on increased strain [@fleg1995].
+Patients who underwent refractory anaphylaxis more often had perioperative reactions and therefore, a decreased ability to compensate the reaction symptoms with reflexive renal or cardiopulmonary response [@Rogers2013].
 
-Compensation mechanisms demonstrated in animal models indicated that anaphylactic hypotension activates the hypothalamic paraventricular nucleus, medullary nucleus tractus solitarii and rostral ventrolateral medulla, independently of the baroreflex pathway.
+Compensation mechanisms demonstrated in animal models indicated that anaphylactic hypotension activates the hypothalamic paraventricular nucleus, medullary nucleus tractus solitarii, and rostral ventrolateral medulla, independently of the baroreflex pathway.
 Further, it stimulates efferent sympathetic nerve activity to the adrenal gland and kidney to restore blood pressure [@tanida2018].
@@
 We saw significantly more adrenaline use in the refractory cases, but it could have been used even more extensively in these life-threatening reactions.
-This observation might indicate that in fact the emergency teams are using these life-saving measures too cautiously.
+This observation might indicate that in fact, the emergency teams are using these life-saving measures too cautiously.
 -->
 
-Adrenaline i.v.
-as first-line therapy use was high in the refractory cases.
-This was probably due to the fact that most of them occur in a medical setting.
+Physicians and surgeons often used adrenaline i.v.
+as first-line therapy in refractory cases, probably because most of them occurred in a medical setting.
 However, other second-line therapeutic options were rarely used.
 Grabenhenrich et al.
-[@Grabenhenrich2018] recently evaluated the epinephrine use in anaphylaxis patients and concluded that, even in this state of the art drug, large discrepancies between recommended use and actual treatment practice exist.
-Similarly, US studies documented poor adherence in patients and caregivers to anaphylaxis guidelines recommending more than 1 adrenaline autoinjector available at all times [@song2018].
-Therefore, more effort needs to be dedicated to promote and develop the consensus guidelenes as practically as possible in order to increase adherence.
+[@Grabenhenrich2018] recently evaluated the epinephrine usage in anaphylaxis patients and concluded that, even in this state of the art drug, significant discrepancies between recommended use and actual treatment practice exist.
+Similarly, US studies documented poor adherence in patients and caregivers to anaphylaxis guidelines recommending more than one adrenaline autoinjector available at all times [@song2018].
+Therefore, more effort needs to be dedicated to promote and develop consensus guidelines as practically as possible in order to increase adherence.
 
-Methylene blue and vasopressors have been described to successfully restore refractory hypotension and recommended by managment guidelines [@Muraro2014], but their actual use in anaphylaxis patients is scarce.
+Methylene blue and vasopressors have been described to successfully restore refractory hypotension and recommended by management guidelines [@Muraro2014], but their actual use in anaphylaxis patients is scarce.
 There are multiple reports of successful methylene blue use in refractory anaphylaxis [@Francuzik2018].
-Evora [@Evora2000] reported 6 cases of refractory, peroperative anaphylaxis (to RCM and penicillin) which responded to methylene blue i.v.
+Evora [@Evora2000] reported 6 cases of refractory, perioperative anaphylaxis (to RCM and penicillin) which responded to methylene blue i.v.
 within minutes.
@@
 
-Suprisingly, even though multiple anaphylaxis management guidelines recommend glucagon infusions in cases of concomitant beta-blocker therapy in anaphylactic patients [@Simons2011;@Lieberman;@Muraro2014], it has not been administered in any of the severe or refractory cases.
+Surprisingly, even though multiple anaphylaxis management guidelines recommend glucagon infusions in cases of concomitant beta-blocker therapy in anaphylactic patients [@Simons2011;@Lieberman;@Muraro2014], it has not been administered in any of the severe or refractory cases.
 Similarly, Royal Collage of Anesthetists reported one glucagon administration in 266 severe intraoperative anaphylaxis episodes [@harper2018].
-Glucagon has been reported to sucessfully relive refractory anaphylaxis [@Zaloga1986;@Javeed1996] by directly activating the adenylyl cyclase and therefore bypassing the β-adrenergic receptor blockade [@Zaloga1986].
+Glucagon has been reported to successfully relive refractory anaphylaxis [@Zaloga1986;@Javeed1996] by directly activating the adenylyl cyclase and therefore bypassing the β-adrenergic receptor blockade [@Zaloga1986].
 
@@
 
-There is no universal definition  of refractory anaphylaxis is not universal.
-The confirmation ofrefractory anaphylaxis is based on the answers to our online questionnaire.
-If a fatal reaction occurred before the second dose of adrenaline was administered to the patient - it was not categorized as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although, it in fact might have been refractory.
+The definition of refractory anaphylaxis is not universal, and we had to assume it based on the answers to our online questionnaire.
+If a fatal reaction occurred before the second dose of adrenaline was administered to the patient - it was not categorized as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although, it might have been refractory.
+
+The therapy with adrenaline was not weight-adapted, and patients received multiple fixed doses of adrenaline ranging from 300 µg - 1000 µg.
+It is possible that patients experiencing anaphylaxis refractory to standard doses of adrenaline were treated with subtherapeutic doses in the initial phase of the reaction due to their obesity.
+However, we did not gather data on the patient's weight, and therefore, we cannot conclude how it influenced the therapy-outcomes with fixed doses of adrenaline.
+Nevertheless, repeated doses of adrenaline should be sufficient to alleviate the symptoms in such patients if the weight would be the only reason for the refractoriness of anaphylaxis.
 
 The comparison of refractory anaphylaxis with patients suffering from severe anaphylaxis enabled us to distinguish patients with a higher risk of developing a refractory episode.
-However we cannot address the question which patients had a higher risk of experiencing anaphylaxis per se.
+However, we cannot address the question which patients had a higher risk of experiencing anaphylaxis per se.
 
@@
 Its elicitors and cofactors differ from other anaphylaxis cases, and the management of refractory anaphylaxis needs to improve.
-Although, more studies need to be conducted to better understand the pathomechanisms involved in refractory anaphylaxis, we propose to increase the use of second line medication such as methylene blue, vasopressin and (in suspicion of a beta adrenergic blockade) glucagon in cases where 2 doses of adrenaline did not result in a rapid normalization of anaphylaxis symptoms.
+Although, more studies need to be conducted to understand better the pathomechanisms involved in refractory anaphylaxis, we propose to increase the use of second-line medication such as methylene blue, vasopressin and (in suspicion of a beta-adrenergic blockade) glucagon in cases where 2 doses of adrenaline did not result in a rapid normalization of anaphylaxis symptoms.
 
@@
 We thank all patients, parents, and their children for their support in providing data on the
-occurrence of anaphylaxis for this study. We thank the study personnel for patients
+the occurrence of anaphylaxis for this study. We thank the study personnel for patients
 counseling and data entry and we thank the members of The European Anaphylaxis Registry
@@
 K. Schäkel (Heidelberg, Germany),
-N. Papadopoulus (Athen, Greece),
+N. Papadopoulos (Athen, Greece),
 M. Kowalski (Lodz, Poland),
@@
 K. Breuer (Hamburg, Germany),
-S. Müller und T. Jakob (Freiburg, Germany),
+S. Müller and T. Jakob (Freiburg, Germany),
 H. Straube (Darmstadt, Germany),
@@
 ```{r flowchart, fig.cap="Flowchart illustrating the cases selection process for the final database."}
+
 knitr::include_graphics("../figures/flow.png")
+
 ```
@@
 ```{r theory, fig.cap="Visual representation of the three forces determining the severity of anaphylaxis.
-The natural ability of the body to compensate the anaphylaxis symptoms and therapeutic measures acting to restore homeostasis to a patient by whom elicitors and cofactors influence synergistically the severity of a given episode."}
+The natural ability of the body to compensate for the anaphylaxis symptoms and therapeutic measures acting to restore homeostasis.
+Elicitors and cofactors influence the severity of a given episode synergistically."}
 
-knitr::include_graphics("../figures/diag2.png")
+knitr::include_graphics("../figures/diag22.png")
 ```
@@
                    as.numeric(elicitExact[,2:4]) %>% matrix(nrow = 6)), 
-             caption = "Summary of the elicitors broken down into specific elicitors from each group in the refractory anaphylaxis cases and severe anaphylaxis cases as a control.
+             caption = "Comparison of the specific elicitors from each elicitor-group betwen the refractory anaphylaxis cases and severe anaphylaxis cases as a control.
 ANA - anaphylaxis, p-value derived from the Fisher's exact test.

### 2019-08-27 insect venom.Rmd
gramarly check

--- 2019-08-20 R/insect venom.Rmd
@@
 Insect-venom elicited anaphylaxis is a common and often severe hypersensitivity reaction.
-It affects older individuals than other anaphylaxis causes (p = `r t.test(data$d_age[data$d_elicitorInsect=="yes"],data$d_age[data$d_elicitorInsect=="no"])$p.val`).
+It affects older individuals than other anaphylaxis causes (p = `r t.test(data3$d_age[data3$d_elicitorInsect=="yes"],data$d_age[data$d_elicitorInsect=="no"])$p.val`).
 Using the data from the European Anaphylaxis Registry (`r v$allCases` cases in total) we identified insect-venom elicited anaphylaxis cases (n = `r v$insectElicited`) and analyzed these in comparison to anaphylaxis elicited by other elicitors (n = `r v$otherEli`).

### 2019-08-27 R/Insects.R

@@ -126,58 +126,59 @@ varTher <- c(227:299)
 # add Seveirity
 # Define a new variable - reaction type to exclude non ANA cases
 #data <- data3
-sentinel_skin <- rep(F, nrow(data))
-sentinel_skin[which(data$q_111_angioedema=="yes"|
-                      data$q_111_urticaria=="yes"|
-                      data$q_111_erythema_flush_v5=="yes"|
-                      data$q_111_pruritus=="yes")] <- T
+sentinel_skin <- rep(F, nrow(data3))
+sentinel_skin[which(data3$q_111_angioedema=="yes"|
+                      data3$q_111_urticaria=="yes"|
+                      data3$q_111_erythema_flush_v5=="yes"|
+                      data3$q_111_pruritus=="yes")] <- T
 
-sentinel_respiratory <- rep(F, nrow(data))
-sentinel_respiratory[which(data$q_115_cyanosis_pallor_v6=="yes"|
-                             data$q_113_stridor_inspiratory=="yes"|
-                             data$q_113_wheezing_expiratory_distress_v5=="yes"|
-                             data$q_113_dyspnea=="yes"|
-                             data$q_113_respiratory_arrest=="yes")] <- T
+sentinel_respiratory <- rep(F, nrow(data3))
+sentinel_respiratory[which(data3$q_115_cyanosis_pallor_v6=="yes"|
+                             data3$q_113_stridor_inspiratory=="yes"|
+                             data3$q_113_wheezing_expiratory_distress_v5=="yes"|
+                             data3$q_113_dyspnea=="yes"|
+                             data3$q_113_respiratory_arrest=="yes")] <- T
 
-sentinel_cardio <- rep(F, nrow(data))
+sentinel_cardio <- rep(F, nrow(data3))
 sentinel_cardio[which(
-  data$q_112_incontinence=="yes"|
-    data$q_114_reductions_of_alertness=="yes"|
-    data$q_114_loss_of_consciousness=="yes"|
-    data$q_114_cardiac_arrest=="yes"|
-    data$q_114_hypotension_collapse_v5=="yes")] <- T
+  data3$q_112_incontinence=="yes"|
+    data3$q_114_reductions_of_alertness=="yes"|
+    data3$q_114_loss_of_consciousness=="yes"|
+    data3$q_114_cardiac_arrest=="yes"|
[…]

### 2019-08-27 R/analysis.R

@@ -10,7 +10,7 @@ require(magrittr)
 # Load your data in this section - it is crucial to know on what data you are actually working
 #load("analysis/data/raw_data/data3.R")
 # New database
-load("analysis/data/raw_data/data4.R")
+load("../data/raw_data/data4.R")
 data3 <- data4
 
 ##### FUNCTIONS #####
@@ -375,7 +375,7 @@ refractoryCasesBrown%in%refractoryCasesRM
 refractoryCasesRM%in%refractoryCasesBrown
 
 export$casesBrown <- data3[refractoryCasesBrown,]
-save(export, file = "analysis/data/derived_data/export.R")
+#save(export, file = "analysis/data/derived_data/export.R")
 
 
 

### 2019-08-27 analysis/paper/paper.Rmd
minor changes

@@
 
-```{r, setup, echo = FALSE,warning=FALSE,message=FALSE}
+```{r, setup, echo = FALSE,warning=FALSE,message=FALSE, include=F}
 knitr::opts_chunk$set(
@@
 
-The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab[elicitorTab$Elicitor=="drugs","percent"]`% of cases, p < 0.0001) compared to other severe anaphylaxis cases (`r elicitorTab[elicitorTab$Elicitor=="drugs","percANA"]`%).
+The data show that drugs more frequently elicited refractory anaphylaxis (`r elicitorTab[elicitorTab$Elicitor=="drugs","%"]`% of cases, p < 0.0001) compared to other severe anaphylaxis cases (`r elicitorTab[elicitorTab$Elicitor=="drugs","% severe ANA"]`%).
 Cases elicited by insects (n = `r sum(v$casesInsect)`) were more often due to bees than wasps in refractory cases (`r 5/8*100`% vs `r round(449/(77+1607+4+9+170+3+449)*100,1)`%, p = `r fisher.test(matrix(c(449,5,77+1607+4+9+170+3,3),nrow=2))$p.value %>% round(3)`).
@@
 # Methods
-The European Anaphylaxis Registry (described in detail elsewhere [@Grabenhenrich2016]) provided data for this analysis status from March 2018.
+The European Anaphylaxis Registry (described in detail elsewhere [@Grabenhenrich2016]) provided data for this analysis (status from March 2018).
 We selected cases where patients received at least two doses of adrenaline and failed to recover adequately and assigned them to the "refractory anaphylaxis group." The flowchart in figure \@ref(fig:flowchart) represents the detailed case-selection process.
@@
 
-`r elicitorTab[1,9]`% of food elicited refractory cases had a previously confirmed diagnosis of food allergy. Insect-venom and food allergens most frequently elicited severe cases of anaphylaxis.
+`r elicitorTab[elicitorTab$Elicitor =="food","Food allergy [n]"] / elicitorTab[elicitorTab$Elicitor =="food","n"]*100 %>% {round(.,1)}`% of food elicited refractory cases had a previously confirmed diagnosis of food allergy.
+Insect-venom and food allergens most frequently elicited severe cases of anaphylaxis.
 Refractory anaphylaxis cases were more often elicited by bees than severe, non-refractory cases, p = `r table(data3$q_340_insects,data3$rANA) %>%  {rbind(addmargins(.[3:8,],1) %>% tail(1),.[2,])} %>% as.matrix() %>% fisher.test() %>% {.$p.value} %>% round(4)` (table \@ref(tab:elicitExact)).
@@
 Moreover, it might be possible that during a refractory episode the bronchospasm is additionally triggered via mediators derived from basophils and eosinophils or the vegetative nerve system what makes it refractory to adrenaline and results in a prolonged anaphylactic episode.
-Following this line, we observed more respiratory distress symptoms in the refractory anaphylaxis cases.
+Accordingly, we observed more respiratory distress symptoms in the refractory anaphylaxis cases.
 
 <!-- The influence of tryptase -->
-Baseline serum tryptase (BST) levels were increased in patients with refractory anaphylaxis, although only in 7 patients it was over the threshold level of 11.5 ng/mL - what is commonly recognized as elevated [@klion2003].
+Baseline serum tryptase (BST) levels were increased in patients with refractory anaphylaxis, although only in 7 patients they were over the threshold level of 11.5 ng/mL - what is commonly recognized as elevated [@klion2003].
 BST may reflect pro-β tryptase due to increased body mast cell content [@CAUGHEY2006] or result from alpha-tryptasemia due to multiple copies of *TPSAB1* gene [@Lyons2018].
@@
 ```
-The higher rate of malignancy in patients with refractory anaphylaxis diseases might be a confounder as certain chemotherapeutics are given intravenously and are well known to activate mast cells (e.g.
+The higher rate of malignancy in patients with refractory anaphylaxis diseases might be a confounder as certain intravenous chemotherapeutics are well known to directly activate mast cells (e.g.
 taxanes) [@lee2009].
@@
 These mediators may contribute to an increased mast cell response and more severe anaphylaxis.
-Some published case reports indicated for stress as the anaphylaxis trigger [@alevizos2014].
+Some published case reports indicated stressful event as the anaphylaxis trigger [@alevizos2014].
 
@@
 indicated the role of PPI as a risk factor for severe anaphylaxis by prolonging the exposition to an oral allergen (preventing its degradation due to lower gastric acid production) [@Wlbing2013].
-Moreover, there are reports on PPI themselves as elicitors of anaphylaxis @Natsch2000].
+Moreover, there are reports on PPI themselves as elicitors of anaphylaxis [@Natsch2000].
 We observed a significantly higher rate of PPI intake in the refractory anaphylaxis group.
@@
 The definition of refractory anaphylaxis is not universal, and we had to assume it based on the answers to our online questionnaire.
-If a fatal reaction occurred before the second dose of adrenaline was administered to the patient - it was not categorized as refractory (as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis) although, it might have been refractory.
+If a fatal reaction occurred before the second dose of adrenaline was administered to the patient (although, it might have been refractory) - it was not categorized as refractory as we defined at least two doses of minimum 300 µg adrenaline for refractory anaphylaxis.
 
@@
 Although, more studies need to be conducted to understand better the pathomechanisms involved in refractory anaphylaxis, we propose to increase the use of second-line medication such as methylene blue, vasopressin and (in suspicion of a beta-adrenergic blockade) glucagon in cases where 2 doses of adrenaline did not result in a rapid normalization of anaphylaxis symptoms.
+
+# Funding 
+The European Anaphylaxis Registry was supported by the Network for Online-Registration of Anaphylaxis NORA e. V.
+We acknowledge support from the German Research Foundation (DFG) and the Open Access Publication Fund of Charité – Universitätsmedizin Berlin.
+
+# Conflict of Interest Statement
+The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
 

### 2019-08-29 analysis/paper/paper.Rmd
figures and acknowledgements

@@
 
-`r elicitorTab[elicitorTab$Elicitor =="food","Food allergy [n]"] / elicitorTab[elicitorTab$Elicitor =="food","n"]*100 %>% {round(.,1)}`% of food elicited refractory cases had a previously confirmed diagnosis of food allergy.
+`r round(elicitorTab[elicitorTab$Elicitor =="food","Food allergy [n]"] / elicitorTab[elicitorTab$Elicitor =="food","n"]*100, 0) `% of food elicited refractory cases had a previously confirmed diagnosis of food allergy.
 Insect-venom and food allergens most frequently elicited severe cases of anaphylaxis.
@@
 <!-- The influence of tryptase -->
-Baseline serum tryptase (BST) levels were increased in patients with refractory anaphylaxis, although only in 7 patients they were over the threshold level of 11.5 ng/mL - what is commonly recognized as elevated [@klion2003].
+Baseline serum tryptase (BST) levels were increased in patients with refractory anaphylaxis, although only in 7 patients they exceeded the threshold level of 11.5 ng/mL - what is commonly recognized as elevated [@klion2003].
 BST may reflect pro-β tryptase due to increased body mast cell content [@CAUGHEY2006] or result from alpha-tryptasemia due to multiple copies of *TPSAB1* gene [@Lyons2018].
@@
 in detail:
-J. Grünhagen (Berlin, Germany),
-K. Beyer (Berlin, Germany),
-A. Möser (Jena, Germany),
-T. Fuchs (Göttingen, Germany),
-F. Ruëff (Munich, Germany),
-B. Wedi (Hanover, Germany),
-F. Friedrichs (Aachen, Germany),
-H. Dickel (Bochum, Germany),
-H. Merk (Aachen, Germany),
-U. Hillen (Essen, Germany),
-A. Bauer (Dresden, Germany),
-N. Wagner (Erlangen, Germany),
-E. Rietschel (Cologne, Germany),
-R. Treudler, S. Aurich (Leipzig, Germany),
-L. Klimek, O. Pfaar (Wiesbaden, Germany),
-N. Reider (Innsbruck, Germany),
-W. Aberer (Graz, Austria),
-A. Köhli (Zurich, Switzerland),
-F. Riffelmann (Schmallenberg, Germany),
-B. Kreft (Halle, Germany),
-K. Nemat (Dresden, Germany),
-T. Kinaciyan (Vienna, Austria),
-R. Brehler (Münster, Germany),
-J. Witte (Hamburg, Germany),
-N. Hunzelmann und I. Huseynow (Cologne, Germany),
-T. Bieber (Bonn, Germany),
-U. Rabe (Treuenbritzen, Germany),
-P. Schmid-Grendelmeier (Zurich, Switzerland),
-W. Brosi (Würzburg, Germany),
-S. Nestoris (Lippe-Lemgo, Germany),
-T. Hawranek (Salzburg, Austria),
-R. Bruns (Greifswald, Germany),
-S. Lehmann (Aachen, Germany),
-L. Lange (Bonn, Germany),
-G. Hansen (Hanover, Germany),
-C. Pföhler (Homburg, Germany),
-E. Varga (Graz, Austria),
-Z. Szepfalusi (Vienna, Austria),
-P. Eng (Aarau, Switzerland),
-P. Eng (Lucerne, Switzerland),
-T. Reese (Rheine, Germany),
-M. Polz (Rüsselsheim, Germany),
-S. Schweitzer-Krantz (Düsseldorf, Germany),
-H. Rebmann (Tübingen, Germany),
-G. Stichtenoth (Lübeck, Germany),
-S. Theis (Schwedt, Germany),
-I. Yildiz (Neumünster, Germany),
-M. Gerstlauer (Augsburg, Germany),
-A. Nordwig (Dresden, Germany),
-R. Schlags (Wangen im Allgäu, Germany),
-I. Neustädter (Fürth, Germany),
-C. Stadlin (Zurich, Switzerland),
-M. Bücheler (Bonn, Germany),
-S. Volkmuth (Velbert, Germany),
-J. Fischer (Tübingen, Germany),
-A. Henschel (Berlin, Germany),
-S. Plank-Habibi (Alzenau, Germany),
-B. Schilling (Passau, Germany),
-A. Kleinheinz (Buxtehude, Germany),
-K. Schäkel (Heidelberg, Germany),
-N. Papadopoulos (Athen, Greece),
-M. Kowalski (Lodz, Poland),
-M. Fernandez-Rivas (Madrid, Spain),
-K. Solarewicz-Madajek (Wroclaw, Poland),
-C. Körner-Rettberg (Bochum, Germany),
-T. Mustakov (Sofia, Bulgaria),
-K. Hartmann (Lübeck, Germany),
-C. Kemen (Hamburg, Germany),
-U. Miehe (Leipzig, Germany),
-C. Ebner (Vienna, Austria),
-S. Haak (Oldenburg, Deutschland),
-V. Cardona (Barcelona, Spain),
-S. Hämmerling (Heidelberg, Germany),
-E. Arroabarren (Pamplona, Spain),
-M. Bilo (Ancona, Italy),
-N. Cabañes Higuero (Toledo, Spain),
-A. Vega Castro (Guadalajara, Spain),
-I. Poziomkowska-Gęsicka (Szczecin, Poland),
-S. Büsing (Osnabrück, Germany),
-U. Klettke (Berlin, Germany),
-C. Virchow (Rostock, Germany),
-G. Christoff (Sofia, Bulgaria),
-U. Jappe (Borstel, Germany),
-K. Breuer (Hamburg, Germany),
-S. Müller and T. Jakob (Freiburg, Germany),
-H. Straube (Darmstadt, Germany),
-C. Vogelberg (Dresden, Germany),
-F. Knöpfel (Norderney, Germany),
-J. Hourihane (Cork, Ireland),
-B. Rogala (Silesia, Poland),
-A. Montoro (Madrid, Spain),
-A. Brandes (Frankfurt/Oder, Germany),
-A. Muraro (Padua, Italy),
-T. Buck und J. Büsselberg (Hanover-Misburg, Germany),
-N. Zimmermann (Potsdam, Germany),
-D. Hernandez (Valencia, Spain),
-P. Minale (Genua, Italy),
-J. Niederwimmer und B. Zahel (Linz, Austria),
-A. Fiocchi (Rome, Italy),
-A. Reissig (Gera, Germany),
-F. Horak (Vienna, Austria),
-N. Klossowiski (Düsseldorf, Germany),
-F. Eitelberger (Wels, Austria),
-H. Ott (Hanover, Germany),
-R. Asero (Milan, Italy),
-S. Pistauer (Sylt/Westerland, Germany),
-M. Geißler (Ribnitz-Damgarten, Germany),
-L. Ensina (Sao Paulo, Brazil),
-A. Plaza Martin (Barcelona, Spain),
-J. Meister (Aue, Germany),
-S. Hompes (Hamburg, Germany),
-S. Stieglitz (Wuppertal, Germany).
+
+```{r, results="asis"}
+dfca <-readxl::read_excel("../data/derived_data/Co-authors 2019-08-28.xlsx")
+names(dfca) <- "names"
+temp <- dfca %>% 
+  dplyr::as_tibble() %>% 
+  dplyr::filter(!is.na(names)) %>% 
+  dplyr::pull(names) %>% 
+  paste(collapse = ", ")  %>% 
+  cat
+```
 
@@
 
-knitr::include_graphics("../figures/diag22.png")
+knitr::include_graphics("../figures/diag.png")
 ```

### 2019-08-29 R/Insects.R

@@ -744,10 +744,13 @@ nodes <-
     width = c(1.8,3.5,1.8,1.8,3.5,1.8,1.8,5.8),
     x = c(1,1,1,3,1,-1, 1, 1),
     y = c(0,-1,-2,-3,-5,-3,-3,-4),
-    height = c(rep(0.6,7),1))
+    height = c(rep(0.6,7),1),
+    color = "black",
+    fillcolor = "lightgrey")
 
 edges <- create_edge_df(from = c(1,2,3,4,2,6,3,7,8),
-                        to = c(2,3,4,8,6,8,7,8,5))
+                        to = c(2,3,4,8,6,8,7,8,5),
+                        color = "black")
 # Add the node data frame to the
 # graph object to create a graph
 # with nodes
@@ -759,8 +762,11 @@ create_graph() %>%
     value = "ortho",
     attr_type = "graph") %>%
   add_global_graph_attrs(value="black",
+                         attr = "color",
+                         attr_type = "graph") %>%
+  add_global_graph_attrs(value="black",
                          attr = "fontcolor",
-                         attr_type = "node") %>% #render_graph()
+                         attr_type = "node") %>%#render_graph()
   export_graph(file_name = "analysis/figures/flow.png",
                file_type = "png",
                title = NULL,
